---
title: "Characterizing core morphology and scar patterning within a unified spherical harmonic framework"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - uw
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      orcid: 0009-0008-1504-9401
      institute:
        - ALPHA
        - MOE
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China; 
  - MOE: MOE Key Laboratory of Western China's Environmental System, Center for Excellence in Archaeological Science and Cultural Heritage, Lanzhou University, Lanzhou, China
  - uw: Department of Anthropology, University of Washington, Seattle, WA, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx"
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
number-sections: true
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 800
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-method-and-theory.csl"
abstract: |
  Lithic cores preserve the most direct evidence of past reduction strategies, yet their analysis still relies largely on typological classifications that are subjective and difficult to reproduce. Recent 3D methods quantify core attributes more precisely, but morphology and scar organization are handled in separate mathematical frameworks, precluding joint analysis of how the two relate. Here we present a unified framework based on spherical harmonics that encodes both core morphology (M-SPHARM) and scar patterning (SP-SPHARM) within a single rotation-invariant, landmark-free space, the latter being the first application of spherical harmonics to scar orientation data. Co-inertia Analysis examines their covariation, and harmonic reconstruction renders statistical axes back into interpretable three-dimensional forms. while While the methods are complementary, validation on ideal models and experimental cores shows that the scar pattern descriptor resolves finer distinctions among core types than existing metrics, most clearly the radial subtypes others cannot separate. Exploratory analysis on experimental and archaeological assemblages reveals no detectable covariation between morphology and scar organization. We identify core type as the principal factor structuring variation at Sandinggai (Hunan, China). The methods are released as an open-source R package (spharmlithic) with a fully reproducible workflow. More broadly, the framework supports understanding core variability as a continuous techno-morphological space rather than a set of discrete types.
  
editor_options: 
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

```{r}
#| label: setup
#| include: false
library(targets)
library(tidyverse)
tar_config_set(store = "/project/analysis/paper/_targets")
sa          <- tar_read(spharm_analysis)
ph          <- sa$spharm_posthoc
im          <- tar_read(im_comparison)
dist_spi    <- im$dist_spi
dist_fabric <- im$dist_fabric
dist_spharm <- im$dist_spharm
cia         <- tar_read(exp_cia_analysis)
sdg         <- tar_read(sdg_cia_analysis)
```

# Introduction

Cores preserve the most direct evidence of reduction strategies in lithic assemblages. Their analysis informs not only knapping behavior, including flaking strategies, knapping skill levels, and technological traditions, but also broader questions of hominin cognition, mobility patterns, site function, and social learning.

Core analysis has traditionally relied on typological approaches, in which cores are classified according to hierarchically organized discrete attributes such as morphology, scar patterning, flaking angles, and platform preparation into categories such as Levallois, discoidal, and laminar cores. These frameworks, however, are less standardized: different typological systems often employ distinct descriptive vocabularies, and substantial classification inconsistencies arise both between analysts and within the same analyst over time. The emerging debate over Levallois technology in China illustrates this problem clearly. Although researchers commonly apply the criteria proposed by Boëda, disagreements persist over whether certain cores from the Shiyu and Guanyindong sites should be classified as Levallois. Efforts to establish unified classificatory systems, such as the Unified Lithic Taxonomy and the Logic Analytical System, have partially reduced inter-observer variability, yet they remain fundamentally reliant on qualitative judgment.

Growing demands for precision and reproducibility have driven increasing interest in three-dimensional quantitative approaches to lithic analysis. 3D modeling technologies provide substantially higher-resolution data than traditional linear measurements and have been widely applied to flakes, retouched tools, and large cutting tools (LCTs). Studies specifically focused on cores, however, remain limited. Two broad approaches characterize existing work. The first extracts predefined technological variables from 3D models, such as platform angles, cortex proportions, reduction intensity, and scar patterning. The second characterizes overall core morphology, most commonly through landmark-based geometric morphometrics, though recent studies have also applied spherical harmonic methods (SPHARM) to spheroids and polyhedral cores. These two approaches are complementary, and their integration has yielded deeper insights in several studies.

These approaches, however, have notable limitations. Landmark-based methods compress continuous morphological variation into a limited number of discrete points, inevitably losing information. They also rely on homology assumptions: current landmark configurations, which typically place landmarks along core edges and semi-landmarks across surfaces, are mainly applicable to highly standardized systems such as Levallois reduction and difficult to extend to more variable types or to cross-type comparisons. Even within Levallois core studies, landmark configurations vary between researchers, making direct comparisons between morphospaces problematic. Spherical harmonic approaches offer a promising alternative. They do not require homology assumptions, capture morphological information more comprehensively, and produce rotationally invariant power spectra that eliminate the need for manual core orientation. To date, however, their application has been largely restricted to near-spherical types such as spheroids and polyhedral cores, and their potential for more morphologically diverse core types has yet to be tested.

Quantitative approaches to scar pattern analysis are less developed. Only two systematic frameworks exist: the Scar Pattern Index (SPI), a univariate metric measuring directional concentration through the ratio of resultant to total vector length, and fabric analysis, which uses eigenvalues of orientation tensors to provide a high-level summary of scar patterns. Both represent important advances, yet both reduce complex scar-vector configurations to a small number of summary parameters that capture only relatively coarse-grained orientation organization rather than the full spatial structural properties of scar patterning itself. This information loss is reflected in their performance: SPI and fabric approaches effectively distinguish broad technological organizations, such as unidirectional or bidirectional removals and Levallois or discoidal cores (radial types), but struggle to differentiate variability within these categories. A method capable of representing the spatial structure of scar organization more completely, rather than reducing it to a few summary statistics, is therefore needed.

The relationship between how a core is reduced and the shape it ultimately assumes is an important question in lithic analysis. One influential line of work holds that knappers actively structured removal sequences to generate and maintain particular core morphologies, most explicitly in prepared core types such as Levallois. Some research emphasizes that core form is also, and sometimes primarily, shaped by factors largely independent of technological intention, including raw material quality, blank geometry, and reduction intensity. Whether, and to what degree, core morphology reflects deliberate technological design rather than the cumulative by-product of working a given material therefore remains genuinely open, with direct bearing on how we infer planning, skill, and cultural transmission from stone tools. Adjudicating between these alternatives requires that technological organization and core morphology be quantified independently and that their covariation be tested directly, something no existing framework achieves within a single, reproducible representation. The problem is compounded by the available tools: scar pattern metrics such as SPI and fabric analysis compress organization into a handful of summary parameters, while landmark-based geometric morphometrics imposes homology assumptions that are tractable only for highly standardized forms and that differ from analyst to analyst, foreclosing the cross-type, cross-assemblage, and cross-period comparison on which questions of technological origin and transmission ultimately depend.

Here we develop a spherical harmonic framework that represents core morphology and scar organization within a single mathematical space, recasting core variation from a set of discrete typological categories into a continuous, jointly quantified techno-morphological space. Scar vectors are converted into continuous spherical distributions through von Mises–Fisher kernel density estimation and expanded into rotation-invariant power spectra, placing scar organization on the same footing as morphology, which is described by the same expansion of the core surface. Because the method requires neither homologous landmarks nor manual orientation, its descriptors are directly comparable across core types, raw materials, regions, and time periods. Treating technology and morphology as separable but jointly analyzable dimensions, we use Co-inertia Analysis to test their covariation and exploit the reconstructive capacity of spherical harmonics to translate statistical axes back into interpretable three-dimensional forms. We first validate the framework against idealized digital cores and an experimental knapping dataset, and then demonstrate it on the assemblage from Sandinggai, China. This application serves not as a regional end in itself, but as a worked example of a framework designed to travel to broader problems, from the contested status of Levallois in East Asia and the origins of prepared-core technologies to the reconstruction of technological transmission across hominin populations.

# Material and Methods

**2.1 Materials**

Three datasets were used in this study, each serving a distinct analytical role. The Ideal Digital Cores dataset was used to evaluate the sensitivity of spherical harmonic methods to standardized scar patterns. The Experimental Knapping Cores dataset was employed to test the scar pattern analysis framework under more realistic conditions of morphological and technological variability, and to evaluate the integrated statistical framework proposed in this study. The Archaeological Specimens dataset was then analyzed within this validated framework to demonstrate its practical archaeological applicability.

[Ideal digital cores]{.underline} Eight idealized digital core models, sourced from Sam Lin, were used, comprising cylindrical, conical, discoidal, Levallois (preferential, convergent, and laminar), bifacial, and polyhedral types. Each category is associated with distinct scar organizations, described in detail in the supplementary materials.

[Experimental knapping cores]{.underline} The experimental cores were produced by Chris Clarkson and comprise 58 specimens in this study: 16 unidirectional, 6 bidirectional, 21 Levallois (10 preferential, 5 convergent, 2 Nubian, 2 recurrent, and 2 laminar), 5 discoidal, and 10 multiplatform cores. These specimens are curated at the School of Social Sciences, University of Queensland. The cores were digitized by Sam Lin using a Polyga HDI Compact C210 structured-light scanner with FlexiScan3D software.

[Archaeological specimens]{.underline} The archaeological specimens derive from the Sandinggai site (hereafter SDG) in Hunan Province, China, previously reported by Li. The SDG sequence comprises three cultural layers (L2, L3, and L4), dated by optically stimulated luminescence (OSL) to approximately 100–20 ka. The lithic assemblage is characterized by two operational sequences: (1) a small-flake system primarily based on chert, and (2) a large-flake and large cutting tool (LCT) system primarily based on quartz sandstone.

From the previously reported core assemblage, 51 complete and well-preserved specimens retaining clear technological information were selected for analysis, after excluding incomplete, heavily abraded, and test cores. These derive from Layers 2, 3, and 4; however, because only two cores originate from Layer 2, statistical analyses are restricted to Layers 3 and 4. Core classifications follow the original paper, and details of core types and raw materials are provided in Table X.

Archaeological cores were digitized using an EinScan Pro XS handheld 3D scanner (SHINING 3D) with a turntable and fixed camera stand. Model alignment and export were completed using the accompanying EinScan-Pro software.

For both experimental and archaeological datasets, post-processing was conducted in Geomagic Wrap, including the filling of minor holes and the correction of mesh defects such as non-manifold edges and self-intersections.

**2.2 Methods**

**2.2.1 Data Extraction**

Scar orientations were recorded following the protocol established by Lin. Only scars larger than 5 mm were included. Each scar was recorded as a vector defined by the three-dimensional coordinates of its start and end points, using Geomagic Wrap 2021 for data acquisition. A best-fit plane was also calculated for each core model, and the coordinates of the plane centroid and normal vector were exported for model alignment.

Two alignment procedures were employed to eliminate initial phase differences among core models. The first follows Lin, aligning models based on their best-fit planes by rotating the plane normal to coincide with the Z-axis and translating the starting point of the longest scar to the global origin. The second is proposed here for the first time and relies exclusively on the spatial organization of scar data: (1) unit vectors are calculated for all scars and subjected to singular value decomposition (SVD); (2) the third singular vector, representing the direction least correlated with overall scar orientation structure, is selected as the alignment-plane normal and rotated to the Z-axis; (3) the global centroid of all scar coordinates is translated to the origin; and (4) the first singular vector, representing the principal direction of scar organization, is aligned with the X-axis. The original vectors and the vectors aligned under both procedures were then used to evaluate the rotational invariance of the scar pattern metrics described below. To allow direct visual inspection of the alignment outcomes, interactive HTML visualizations of each model under both procedures are provided in the accompanying repository.

**2.2.2 Spherical Harmonic Analysis**

Spherical harmonic analysis represents the three-dimensional extension of elliptic Fourier analysis. The method characterizes surfaces of closed genus-zero objects by mapping each vertex onto the unit sphere through equivalent coordinate functions and representing the surface as a weighted sum of spherical harmonic basis functions. The resulting shape descriptors are invariant to rotation, translation, and scaling, and can be used to reconstruct the original three-dimensional morphology directly. SPHARM was originally developed for morphometric analysis of brain structures in medical imaging and has since been applied across multiple disciplines. In lithic studies, Sholts used spherical harmonics to quantify asymmetry in Clovis points, while Muller and Ye applied the method to spheroid morphology. Beyond morphological characterization, mathematical relationships between spherical harmonic representations and directional data have also been established. In the present study, we exploit this relationship by applying SPHARM to scar orientation data for the first time.

The analytical workflow for scar pattern analysis proceeds as follows. Discrete scar vectors are first transformed into a continuous density distribution on the sphere using von Mises–Fisher spherical kernel density estimation (VMF-sKDE):

$$f(\mathbf{g}) = \frac{1}{n} \sum_{i=1}^{n} \exp\left(\kappa \, \mathbf{g} \cdot \mathbf{x}_i\right)$$

where $\mathbf{g}$ is a unit vector on the spherical grid, $\mathbf{x}_i$ is the unit vector representing the orientation of the i-th scar, and $\mathbf{κ}$ is the concentration parameter of the von Mises–Fisher kernel, related to the bandwidth $\mathbf{h}$ by:

$$\kappa = \frac{1}{h^2}$$

A bandwidth of 0.35 ($\mathbf{κ}$ ≈ 8.16) was selected to balance directional resolution against oversmoothing. This choice was validated by reconstructing the original distributions, in which distinct scar patterns remained clearly identifiable.

The KDE distributions were interpolated onto a 64×128 Driscoll–Healy grid and subjected to spherical harmonic expansion. Each distribution $f(\theta, \phi)$ is represented as a weighted combination of spherical harmonic basis functions $Y_l^m(\theta, \phi)$:

$$f(\theta, \phi) = \sum_{l=0}^{\infty} \sum_{m=-l}^{l} \hat{f}(l, m) \, Y_l^m(\theta, \phi)$$

where the coefficients $\hat{f}(l, m)$ quantify the contribution of each basis function, and $\mathbf{l}$ and $\mathbf{m}$ denote harmonic degree and order, respectively. A truncation threshold of $l_{\max}$ = 20 was adopted in our study, consistent with previous studies demonstrating that this order sufficiently captures detailed artifact morphological variation. The resulting coefficients retain phase information and can be used to reconstruct original three-dimensional forms, but phase sensitivity may introduce noise in comparative analyses. Accordingly, scar pattern spherical harmonic (hereafter SP-SPHARM) power spectra were also calculated by summing squared coefficients within each harmonic order and normalizing to unit total, providing a phase-independent, rotationally invariant representation of scar pattern structure.

For core morphological analysis, we followed the procedures established by Ye. Three-dimensional models (.stl/.ply) were downsampled to 20,000 faces, and Laplacian smoothing was applied to reduce surface noise and regularize vertex distribution. Centroid normalization and principal-axis alignment were then performed to position all vertices within a common coordinate system, which was transformed into spherical coordinates for harmonic expansion. Spherical harmonic coefficients and morphological spherical harmonic (hereafter M-SPHARM) power spectra were calculated at $l_{\max}$ = 20, as for the scar pattern analysis. The complete workflow is illustrated in Figure X.

Although coefficients were computed to $l_{\max}$ = 20 for faithful reconstruction, retaining all 20 orders for statistical comparison would add dimensionality and, for scar patterns, noise. We therefore screened orders prior to all statistical analyses using two diagnostics: the across-specimen coefficient of variation (CV), indexing the reliability of each order's between-specimen signal, and cumulative power, indexing the share of total variance each order carries. The two spectra behaved differently and were truncated on different grounds. For SP-SPHARM, power was effectively exhausted by the sixth order (cumulative energy ≈ 99.9%), and higher orders were noise-dominated (CV > 100% from $\mathbf{l}$ ≈ 9); we therefore retained $\mathbf{l}$ = 1–6. For M-SPHARM, between-specimen signal remained reliable across all orders (CV < 90% throughout), but variance was strongly concentrated in the low orders, with ~98% of cumulative power captured by the eighth order; we retained $\mathbf{l}$ = 1–8, since the reliable but very low-energy higher orders (together under 2% of power) would inflate dimensionality and dilute multivariate comparisons without materially improving discrimination. This contrast confirms that scar pattern spectra are markedly lower-dimensional than morphological spectra (diagnostics in Fig. SX).

**2.2.3 Data Analysis**

[Rotation Invariance Assessment.]{.underline} Rotational invariance of three scar pattern metrics (SPI, fabric Elongation and Isotropy, and SP-SPHARM power spectra) was assessed using the Bland–Altman approach. This method evaluates agreement between paired measurements by plotting their difference against their mean. The mean difference represents systematic bias, and the 95% limits of agreement (LoA) define the range within which 95% of differences are expected to fall; narrow LoA with no systematic trends indicate high agreement. Each metric was computed under three coordinate conditions: (1) original scar-vector data, (2) morphological-plane alignment, and (3) technological-plane alignment. Pairwise Bland–Altman comparisons among these conditions were conducted, with high agreement interpreted as evidence of rotational invariance.

[Method Comparison.]{.underline} For the ideal digital cores, where only one model exists per type, discriminatory power was assessed using pairwise standardized Euclidean distances in feature space, with greater distances indicating stronger inter-type separation. Results were visualized using heatmaps. For the experimental knapping cores, a significance-testing framework was employed. SPI, as a univariate metric, was evaluated using the Kruskal–Wallis test. Fabric metrics and SP-SPHARM descriptors were analyzed using PERMANOVA, which tests overall structural differences between groups, complemented by PERMDISP to assess whether differences in within-group dispersion contribute to the observed separation. Significant PERMDISP results warrant cautious interpretation of PERMANOVA outcomes. All permutation procedures used 9,999 permutations. Where overall tests were significant, post hoc pairwise comparisons were performed with Holm–Bonferroni correction.

[Techno-morphological statistical analytical framework.]{.underline} To investigate techno–morphology relationships, we employed a two-step analytical framework: first testing whether overall covariation exists between M-SPHARM and SP-SPHARM power spectra, then exploring interaction patterns through joint dimensional reduction.

*Isometric log-ratio transformation.* Spherical harmonic power spectra are compositional data: the total power is constrained to a fixed sum, so variation in one harmonic order necessarily induces spurious covariance among others. To eliminate this closure effect, Isometric log-ratio transformation (ILR) was applied to all power spectrum data, projecting them into Euclidean space through log-ratio calculations among harmonic orders. All subsequent statistical analyses were conducted on ILR-transformed data.

*Mantel test and RV coefficient.* Overall covariation was evaluated using two complementary approaches. The Mantel test assesses correlation between two distance matrices via Spearman's rank correlation coefficient, with significance determined by 9,999 permutations. The RV coefficient quantifies the overall degree of association between two multivariate datasets, ranging from 0 (no association) to 1 (identical structure), with significance likewise assessed by permutation. The Mantel test emphasizes pairwise distance relationships, whereas the RV coefficient focuses on multivariate covariance structure; their combined use therefore provides a more robust assessment of dataset correspondence.

*Co-inertia analysis.* Co-inertia Analysis (CoIA) is a multivariate method for evaluating structural relationships between two datasets. Its core principle is to identify directions of variation along which the two datasets are maximally concordant while maximizing their covariance. PCA is first performed separately on each dataset, after which projections are used to identify correspondences among samples and variables across the two data spaces. CoIA is particularly well suited for comparing datasets that differ in physical meaning but share identical sampling units, making it appropriate for jointly analyzing M-SPHARM and SP-SPHARM power spectra, which describe different aspects of core variability across the same set of specimens.

The CoIA ordination axes represent the principal directions of shared variation between the two datasets. Each sample is represented by two points, one from each dataset, connected by a line segment. Line length reflects the degree of discrepancy between a sample's position in the two data spaces: shorter lines indicate stronger correspondence between its morphological and scar pattern signatures, whereas longer lines indicate that the two dimensions characterize that specimen differently. Line direction reflects the trend of covariation for that sample within the shared space. If line directions are broadly consistent across a group of samples, this suggests a stable and internally coherent covariation structure; dispersed directions suggest weaker structural correspondence or the presence of localized variation. Combined with the reconstructive capacity of spherical harmonics, this framework allows the morphological and scar pattern characteristics associated with each end of the CoIA axes to be directly visualized as three-dimensional forms, thereby reconnecting statistical patterns to concrete archaeological interpretation.

To further characterize sample correspondences, we employed Kruskal–Wallis tests to compare line lengths among groups, assessing whether the degree of technology–morphology coupling differs across core types. Rayleigh tests were applied to line directions to evaluate whether orientations exhibit significant directional clustering. Because line directions in CoIA represent axial correspondences rather than unidirectional vectors, they were treated as axial data, with directions differing by 180° considered equivalent. The combined analysis of line lengths and directions enables simultaneous evaluation of coupling strength and structural consistency from both distance-based and directional perspectives.

*Reconstruction along CoIA axes.* To visualize the coupled techno–morphological variation captured by CoIA, we developed a continuous reconstruction framework operating directly in spherical harmonic coefficient space. Spherical harmonic coefficients of individual specimens were projected onto the CoIA axes, and at evenly spaced positions along each axis, reconstructed forms were generated by Gaussian kernel-weighted averaging of neighboring specimens' coefficients. The weighted coefficients were then transformed back into three-dimensional meshes (M-SPHARM) or spherical density surfaces (SP-SPHARM) through inverse spherical harmonic reconstruction. The resulting forms represent localized structural tendencies along covariance gradients rather than static group averages, enabling gradual transitions in morphology and scar organization to be visualized as continuous trajectories within the shared CoIA space.

**2.2.4 Reproducibility and open source materials**

Computational reproducibility remains an ongoing challenge in archaeological science, yet it is essential for the credibility and cumulative progress of quantitative research. To support reproducibility and re-use, all materials necessary to reproduce the analyses presented here are provided in a version-controlled online repository at XX. This includes the complete R and Python scripts used for all analyses and visualizations, and the raw datasets underlying all reported results. The repository is organized following established best practices for research compendia, with a logical folder structure and relative file paths that allow all code to be executed without manual modification. All figures, tables, and statistical results presented in this paper can therefore be independently reproduced from these materials. 

In addition, we provide an open-source R package (spharmlithic) that implements the essential analytical methods developed in this study, including model alignment, SPI and fabric computation, spherical harmonic analysis of both scar patterns and core morphology, and three-dimensional visual reconstruction. This package is designed as a general-purpose tool to enable other researchers to apply these methods to their own datasets.

# Results

**3.1 Validity Analysis**

**3.1.1 Rotational Invariance of Scar Pattern Analysis Measures**

To assess whether scar pattern metrics require coordinate alignment prior to analysis, rotational invariance was evaluated using the Bland–Altman approach. SPI and fabric metrics showed near-perfect agreement across all three coordinate conditions (original, morphological-plane aligned, and technological-plane aligned), with biases and 95% LoA on the order of 10⁻¹⁶, well within floating-point numerical precision (@fig-rotation-validity). Both measures are therefore strictly rotationally invariant, and coordinate alignment is not required prior to their computation.

SP-SPHARM descriptors exhibited small systematic biases in low-order power ($\mathbf{l}$ = 1–2; 95% LoA = ±0.02), which diminished rapidly with increasing degree; no substantive differences were observed for $\mathbf{l}$ ≥ 3 (@fig-rotation-validity). Random rotational perturbations around the Z-axis produced differences within machine precision, confirming that directional uncertainty within the coordinate frame has no effect on the power spectrum given a fixed Z-axis.

The low-order biases most likely originate from upstream computational steps such as KDE interpolation and coordinate mapping rather than from the power spectrum formulation itself. Crucially, a consistent alignment method applied uniformly across all specimens ensures that this systematic error does not alter relative differences among them. All subsequent analyses therefore employ the technological fitted plane for alignment.

```{r}
#| label: fig-rotation-validity
#| fig-cap: "Bland-Altman analysis of rotational invariance. **(A-C)** SPI across three coordinate conditions. **(D-F)** Fabric elongation ratio across three coordinate conditions. **(G-I)** Fabric isotropy ratio across three coordinate conditions. **(J)** SP-SPHARM power spectra bias across three coordinate conditions and random Z-axis perturbation."
#| fig-width: 12
#| fig-height: 14
#| echo: false
#| dpi: 150
tar_read(p_rotational_invariance_validity)
```

**3.1.2 Validity of the SPHARM Method for Scar Pattern Analysis**

**Test of Ideal models** Inter-type separation increased progressively from SPI to fabric metrics to SP-SPHARM descriptors, as revealed by pairwise standardized Euclidean distance heatmaps (@fig-im-comparison). Mean distances confirmed this ordering: SP-SPHARM yielded the greatest overall separation ($\bar{d}_{\mathrm{E,std}}$ = `r dist_spharm`), substantially exceeding fabric metrics ($\bar{d}_{\mathrm{E,std}}$ = `r dist_fabric`) and SPI ($\bar{d}_{\mathrm{E, std}}$ = `r dist_spi`).

```{r}
#| label: fig-im-comparison
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| fig-cap: !expr >
#|   paste0("Pairwise standardised Euclidean distances between ideal core 
#|   models across three methods. Mean distances: SPI = ", dist_spi,
#|   ", Fabric = ", dist_fabric, ", SPHARM = ", dist_spharm, ".")

im$p_heatmap
```

SPI produced moderate distances between unidirectional types (cylindrical, conical) and radial or centripetal types (discoidal, Levallois), but near-zero distances among Levallois subtypes (preferential, convergent, and laminar) and between discoidal variants, indicating that it primarily captures the degree of directional concentration while remaining insensitive to finer structural differences. Fabric metrics substantially improved separation between unidirectional and radial categories but, like SPI, failed to distinguish scar patterns sharing parallel but opposing orientations: distances between unipolar and bipolar cylindrical types remained near zero. SP-SPHARM produced the highest pairwise distances across nearly all comparisons. Notably, it achieved clear separation between discoidal and Levallois types and between bifacial and unifacial discoidal patterns, distinctions that neither SPI nor fabric metrics resolved. Multiplatform cores were most strongly differentiated from all other types under SP-SPHARM. However, distances between types sharing bipolar symmetry in the power spectrum domain, such as bifacial discoidal and bipolar cylindrical patterns, remained comparatively modest, foreshadowing a limitation also observed in the experimental core analysis below.

**Test of experimental cores** To evaluate the sensitivity of each method on empirical data, we computed all three measures for experimental cores and conducted statistical comparisons by core type. All three methods detected significant inter-type differences (@fig-exp-method-compare), but they resolved different sets of distinctions. SP-SPHARM descriptors separated the broadest set of type pairs, including several that neither summary metric could. Fabric metrics distinguished the majority of type pairs but failed to differentiate unidirectional from bidirectional cores. SPI was the most limited, detecting differences only where unidirectional cores were involved.

```{r}
#| label: fig-exp-method-compare
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 12
#| echo: false

tar_read(spharm_analysis)$p_exp_method_combined
```

These results suggest that the three methods are complementary. SP-SPHARM descriptors capture the overall structure of scar patterning for most pairwise comparisons, though separation is reduced for patterns with bipolar symmetry, such as discoidal-bidirectional pairs ($P_{\text{adj}}$ = `r ph$p_dir_disc_bi`). Fabric metrics effectively distinguishes differences between unidirectional and bidirectional cores from Levallois and discoidal cores, but cannot differentiate unidirectional from bidirectional cores ($P_{\text{adj}}$ = `r ph$p_fab_uni_bi`), nor Levallois from discoidal cores ($P_{\text{adj}}$ = `r ph$p_fab_lev_disc`). SPI is most sensitive to the unidirectional type but shows limited discriminative capacity for other types. Notably, although SP-SPHARM descriptors failed to separete discoidal from bidirectional cores due to their similarity structures in SPHARM power spectrum (@fig-im-comparison), this distinction is successfully recovered by fabric metrics ($P_{\text{adj}}$ = `r ph$p_fab_disc_bi`). PERMDISP tests further revealed that discoidal and multiplatform cores exhibit greater within-group dispersion than other types, suggesting that these core types may be operationally more variable. It should also be noted that bidirectional (n = 6) and discoidal (n = 5) are represented by considerably fewer specimens than other types, which limits statistical power for comparisons involving these groups.

**3.2 Joint Analysis of Scar Patterns and Core Morphology in Experimental Knapping Specimens**

PERMANOVA tests using M-SPHARM descriptors showed core type to account for only a modest, non-significant fraction of morphological variation ($R^2$ = `r sa$perm_morph_r2`, $\text{pseudo-}F$ = `r sa$perm_morph_f`, $P$ = `r sa$perm_morph_p`). The global Mantel test likewise revealed no significant correlation between M-SPHARM and SP-SPHARM descriptor sets (Mantel's $r$ = `r cia$mantel_global_r`, $P$ = `r cia$mantel_global_p`), and the RV coefficient from CoIA did not reach significance ($\text{RV}$ = `r cia$rv`, $P$ = `r cia$rv_p`). Both association measures were negligible in magnitude, indicating that core morphology and scar organization vary largely independently at the assemblage level, rather than reflecting a mere failure to detect a relationship.

Because the global tests above provide no evidence of covariation, the CoIA axes are interpreted here descriptively — as the directions of maximal shared variation extracted by the method — rather than as a statistically supported covariance structure. The first two CoIA axes account for 85.6% of total covariation (Axis 1 = 65.5%, Axis 2 = 20.1%; @fig-exp-cia). Along Axis 1, M-SPHARM reconstructions shift from rounded, equidimensional forms to progressively flattened profiles, while SP-SPHARM reconstructions transition from concentrated, directionally organized density distributions to more dispersed, multidirectional patterns characteristic of centripetal or multiplatform scar organizations. Axis 2 captures a secondary morphological gradient from laterally convex, disk-shaped cross-sections to more uniform, flattened elliptical forms. On the scar pattern side, both axes capture broadly similar transitions in scar organization, reflecting a shift from ordered, directionally structured arrangements, such as unidirectional and bidirectional patterns, toward more dispersed, multidirectional configurations typical of discoidal and multiplatform cores (@fig-exp-cia).

```{r}
#| label: fig-exp-cia
#| echo: false
#| fig-width: 10
#| fig-height: 14
#| fig-cap: "CoIA biplot, line length, and directional distribution of techno-morphological coupling in experimental cores."

cia$p_final
```

Kruskal–Wallis tests revealed no significant difference in line lengths across core types (@fig-exp-cia), indicating that the degree of morphology–scar correspondence does not differ detectably across reduction strategies. Overall, the experimental assemblage shows no detectable covariation between core morphology and scar pattern: both the global Mantel test and the RV coefficient are negligible. At the within-type level, the only directional signal is the clustering of CoIA line directions among bidirectional cores ($P$ = `r cia$rayleigh_bi_p`); because this rests on just six specimens and emerges from an ordination that itself shows no significant assemblage-level covariation, we treat it as a tentative observation rather than established within-type structure. The broader pattern is consistent with morphology and scar organization varying along largely independent axes, while underscoring that larger per-type samples are needed to test for within-type structure directly.

**3.3 Archaeological Case Study: Core Analysis from Sandinggai**

As with the experimental cores, the global Mantel test revealed no significant correlation between M-SPHARM and SP-SPHARM descriptors for the SDG assemblage (Mantel's $r$ = `r sdg$mantel_global_r`, $P$ = `r sdg$mantel_global_p`). Cross-frequency Mantel tests remained non-significant after Holm–Bonferroni correction, confirming that no covariation is detectable in any frequency band. The RV coefficient likewise did not reach significance ($\text{RV}$ = `r sdg$rv`, $P$ = `r sdg$rv_p`).

As in the experimental assemblage, the non-significant global tests mean that the CoIA axes below are read descriptively rather than as evidence of a real covariance structure. The first two CoIA axes account for 90.6% of total covariation (Axis 1 = 57.6%, Axis 2 = 33.0%; @fig-coia-composite). Along Axis 1, M-SPHARM reconstructions transition from flattened, dorsoventrally compressed forms to more rounded, equidimensional profiles, while SP-SPHARM reconstructions shift from concentrated, unidirectionally organized density distributions to more symmetrically opposed, bidirectional patterns. Axis 2 captures additional variation in both dimensions. On the scar pattern side, it distinguishes unidirectional and bidirectional organizations from more centripetal, dispersed configurations. On the morphology side, it primarily reflects mid-frequency shape variation associated with three-fold and four-fold rotational symmetry, likely corresponding to differences in core cross-sectional geometry (@fig-coia-composite).

```{r}
#| label: fig-coia-composite
#| echo: false
#| fig-width: 12
#| fig-height: 18
#| fig-cap: "CoIA biplot, line length, and line direction by layer, raw material, and core type."

sdg$fig_coia_composite
```

Line lengths showed no significant differences among layers ($\chi^2$ = `r sdg$kw_arrow_layer_chi2`, $\text{df}$ = 1, $P$ = `r sdg$kw_arrow_layer_p`), raw materials ($\chi^2$ = `r sdg$kw_arrow_rawmat_chi2`, $\text{df}$ = 1, $P$ = `r sdg$kw_arrow_rawmat_p`), or core types ($\chi^2$ = `r sdg$kw_arrow_type_chi2`, $\text{df}$ = 5, $P$ = `r sdg$kw_arrow_type_p`; @fig-coia-composite), indicating that the degree of morphology–scar correspondence does not differ detectably across any grouping condition. Rayleigh tests on line direction revealed no significant directional concentration in any layer, raw material, or core type (all $P$ > 0.05). The absence of any directionally structured group indicates that the decoupling between morphology and scar organization holds consistently down to the within-type level, with no core type showing a structured techno-morphological trajectory.

PERMANOVA tests identified core type as the dominant factor structuring both dimensions, with significant and comparable effects on morphological variation ($R^2$ = `r sdg$perm_morph_type_r2`, $\text{pseudo-}F$ = `r sdg$perm_morph_type_f`, $P$ = `r sdg$perm_morph_type_p`) and scar pattern variation ($R^2$ = `r sdg$perm_scar_type_r2`, $\text{pseudo-}F$ = `r sdg$perm_scar_type_f`, $P$ = `r sdg$perm_scar_type_p`). Raw material and layer accounted for substantially less variation. Raw material explained a small fraction of morphological variation that did not reach significance ($R^2$ = `r sdg$perm_morph_rawmat_r2`, $\text{pseudo-}F$ = `r sdg$perm_morph_rawmat_f`, $P$ = `r sdg$perm_morph_rawmat_p`) and had a negligible effect on scar pattern ($R^2$ = `r sdg$perm_scar_rawmat_r2`, $\text{pseudo-}F$ = `r sdg$perm_scar_rawmat_f`, $P$ = `r sdg$perm_scar_rawmat_p`).  Layer effects were negligible for both dimensions (morphology: $R^2$ = `r sdg$perm_morph_layer_r2`, $\text{pseudo-}F$ = `r sdg$perm_morph_layer_f`, $P$ = `r sdg$perm_morph_layer_p`; scar pattern: $R^2$ = `r sdg$perm_scar_layer_r2`, $\text{pseudo-}F$ = `r sdg$perm_scar_layer_f`, $P$ = `r sdg$perm_scar_layer_p`) consistent with continuity in reduction behavior across the stratigraphic sequence.

Two patterns emerge from these effect sizes. First, core type explained roughly four to five times more variation than either raw material or layer in both dimensions, identifying it as the principal axis of structured variability at SDG. Second, although core type significantly structured both morphology and scar organization, the two dimensions showed no detectable pairwise covariation (Mantel's $r$ = `r sdg$mantel_global_r`; $\text{RV}$ = `r sdg$rv`): type-level differences in form and in scar organization are not aligned along a shared axis, each responding to core type but largely independently of the other. The somewhat larger effect of raw material on morphology ($R^2$ = `r sdg$perm_morph_rawmat_r2`) than on scar organization ($R^2$ = `r sdg$perm_scar_rawmat_r2`) is consistent with material properties influencing form more than reduction organization, but both effects are small and this tendency should not be over-interpreted at the present sample size.

# Discussion

The principal contribution of this study is conceptual as much as methodological. By embedding scar organization and core morphology within a single rotation-invariant, landmark-free mathematical framework, it treats lithic variability as a continuous techno-morphological space rather than a collection of discrete types. This framework enables scar patterning and form to be quantified separately and their relationship assessed directly. The discussion first evaluates the robustness of the approach across diverse reduction systems, then examines what the decoupling of scar organization and morphology reveals about knapping behaviour, and finally considers its implications for long-standing cross-regional questions. By providing a homology-free, period-agnostic descriptor of core variation, the framework offers analytical possibilities beyond those available to conventional landmark-based approaches.

The validity of SP-SPHARM for characterizing scar organization has been confirmed at multiple levels. Visual inspection of reconstructed ideal models, experimental knapping cores, and SDG specimens all exhibit clear and interpretable differences in scar pattern. Statistically, PERMANOVA results demonstrate the ability of SP-SPHARM to discriminate between most scar pattern categories. Two non-significant comparisons warrant discussion. The failure to separate discoidal from bidirectional cores is likely attributable to limited statistical power given the small sample sizes of both groups (n = 5 and n = 6, respectively). The non-significant distinction between discoidal and multiplatform cores may instead reflect the high internal variability of the multiplatform category, within which certain specimens exhibit scar patterns resembling those of discoidal cores. Importantly, this overlap does not imply that these specimens should be reclassified. Rather, it suggests that scar organization does not correspond in a one-to-one manner with existing typological categories but encodes additional technological dimensions not captured by conventional classification. This observation underscores the value of treating scar patterns and morphology as separate yet complementary dimensions of core variability.

Compared with existing quantitative approaches to scar pattern analysis, SP-SPHARM demonstrates clear advantages, although it does not constitute a complete replacement for either the SPI or the fabric method. SPI is fundamentally a measure of directional concentration: it quantifies the degree to which scar orientations converge toward a single resultant direction and is therefore most sensitive to unidirectional patterns, while performing poorly in discriminating among radial or centripetal scar organizations. The fabric method adds a second analytical dimension by characterizing orientational variation perpendicular to the best-fit core plane, thereby enabling separation between bidirectional and radial core types that SPI cannot achieve. However, because fabric metrics are derived from orientation tensors that treat opposing directions as equivalent, the method cannot distinguish parallel scar systems with opposite orientations, such as unipolar versus bipolar configurations. In contrast, SP-SPHARM encodes the spatial structure of scar organization more completely, rather than compressing it into summary parameters. This is what allows it to separate radial subtypes such as Levallois and discoidal cores, a distinction that neither SPI nor fabric metrics achieved. However, SP-SPHARM reconstructions represent continuous density distributions that cannot be straightforwardly mapped onto discrete scar pattern categories. By contrast, SPI and fabric indices retain stronger direct physical interpretability. In practice, an effective workflow is to first assess overall group differences using SP-SPHARM, and then integrate SPI and fabric analyses to refine archaeological interpretation.

At the level of practical implementation, the three methods also differ in their sensitivity to coordinate alignment. SPI and fabric metrics are strictly rotationally invariant, requiring no alignment prior to computation. SP-SPHARM power spectra show small systematic biases in low-order terms attributable to upstream computational steps, necessitating a consistent alignment procedure across all specimens. This additional requirement is a practical cost, but one that is offset by the finer structural distinctions SP-SPHARM resolves. More broadly, all three quantitative approaches share an important advantage over traditional typological classification: they replace subjective categorical judgments with continuous numerical descriptors, thereby reducing inter-observer inconsistency and enabling reproducible, statistically testable comparisons across assemblages.

Beyond method validation, this study integrates M-SPHARM and SP-SPHARM power spectra into a CoIA framework to directly test whether core morphology and scar patterning covary at the assemblage level. This addresses a long-standing gap in core analysis: the lack of a quantitative method capable of simultaneously characterizing both dimensions within a unified analytical framework. Results from the experimental knapping cores indicate that neither the global Mantel test nor the RV coefficient reaches statistical significance,and, aside from a non-robust directional signal in bidirectional cores (n = 6), the CoIA line-direction analyses reveal no structured techno-morphological covariation within core types. Together, these results point to a general absence of detectable covariation between morphology and scar organization: at least at this sample size, similar reduction strategies can produce a range of core morphologies, and conversely, similar morphological outcomes can arise from different technological organizations.

This finding has both substantive and methodological implications. Substantively, the absence of detectable covariation suggests that core morphology and reduction strategy respond to different constraints and should not be treated as interchangeable proxies for one another in archaeological interpretation. Questions about how technological intention is expressed in core form may therefore be better addressed through localized morphology, such as that of the platform or flaking surface, combined with bespoke technological variables, rather than through overall core morphology alone. Methodologically, it validates the analytical approach adopted here, in which morphology and scar pattern are characterized as separate yet complementary dimensions rather than collapsed into a single typological classification. The CoIA framework, combined with the reconstructive capacity of spherical harmonics, further enables this variation to be visualized as continuous trajectories along covariance gradients rather than as discrete typological categories, moving beyond static classification toward a dynamic representation of techno-morphological variability. As an exploratory tool, CoIA remains informative even under decoupling, since it characterizes the overall structure of variation within an assemblage; where sample sizes permit, a more powerful application would be to run it separately within each core type, capturing the within-type spectrum of techno-morphological variation that the present per-type samples are too small to resolve.

Application of the framework to the SDG assemblage provides a direct test of its performance on archaeological materials while yielding new behavioral insights. Previous studies proposed that lithic technology at SDG remained broadly continuous across layers, with two principal operational sequences: a small-flake system based on chert and a large-flake and LCT system based on quartz sandstone. However, whether these two raw material systems involve distinct reduction strategies at a finer behavioral level has not been fully explored. The present results provide quantitative support for technological continuity: no significant differences in either core morphology or scar pattern were detected between Layers 4 and 3, confirming that reduction behavior remained stable across the stratigraphic sequence.

At the same time, the framework clarifies the relative roles of the factors structuring core variation. Core type was the dominant influence on both morphology and scar organization, explaining several times more variation than raw material or stratigraphic layer. Raw material had only a small, non-significant effect on morphology and a negligible effect on scar organization, while layer effects were negligible throughout. This pattern indicates that reduction behavior at SDG was structured primarily by the kind of core being produced rather than by raw material or by time, consistent with a stable technological tradition. The capacity to attribute variation to these competing factors, and to show that morphology and scar organization respond to them largely independently, would be difficult to achieve without the separate quantification of both dimensions afforded by the framework. More broadly, wherever assemblages combine multiple raw materials, the same approach can test whether material differences are expressed in form, in reduction organization, or in both — a question of broad relevance to the role of raw material in the lithic record. This decoupling held consistently across the assemblage: no core type, raw material, or stratigraphic layer exhibited a structured techno-morphological trajectory, reinforcing the conclusion that core morphology and reduction organization vary independently at SDG.

At a broader perspective, the framework proposed here offers potential analytical pathways for addressing several unresolved issues in current studies. For example, the presence of Levallois technology in China has been a matter of persistent contention, with disagreements over whether specific cores from sites such as Guanyindong and Shiyu fulfil Levallois criteria remaining unresolved. The SPHARM framework, by providing continuous quantitative descriptors independent of subjective typological boundaries, may offer additional analytical perspectives for such debates. Similarly, the evolutionary relationship between Levallois technology and Acheulean industries remains a central question in Lower–Middle Paleolithic transition research, with competing hypotheses including independent invention, derivation from handaxe traditions, and development from earlier prepared-core technologies. Recent landmark-based geometric morphometric studies have compared Levallois cores, simple prepared cores, and handaxes, suggesting multiple regional trajectories of origin. However, landmark configurations vary between researchers, limiting cross-study integration and large-scale comparative potential. Because SPHARM does not rely on homologous landmarks and produces standardized spectral descriptors, it offers a potential solution to this limitation. Although the current application focuses on cores and does not yet include large cutting tools, the method is in principle applicable to such artifact classes. To facilitate adoption of this framework, we provide an open-source R package (spharmlithic) that implements the core analytical methods developed in this study, including model alignment, SPI and fabric computation, spherical harmonic analysis of both scar patterns and core morphology, and three-dimensional visual reconstruction. This package enables researchers to apply the methods developed here to their own datasets without requiring specialist programming expertise.

More broadly, the approach has potential for integration with emerging analytical paradigms. Recent work has incorporated elliptic Fourier series into phylogenetic analyses, though these applications have been restricted to two-dimensional morphologies of flaking products. As a three-dimensional generalization of Fourier methods, spherical harmonic analysis could be extended to phylogenetic frameworks. The scar pattern component developed here is particularly relevant in this regard, as it provides a direct line of evidence for reconstructing reduction strategies. Future integration with phylogenetic modeling may offer new opportunities to investigate technological transmission and processes of social learning among hominin groups.

Several limitations should be acknowledged. First, sample sizes for certain core types, particularly bidirectional and discoidal cores in the experimental assemblage, are small, and the current results should therefore be interpreted as exploratory rather than definitive. Second, the experimental cores were produced by a single knapper, which may not fully capture the range of inter-individual variability expected in archaeological assemblages. Finally, the present study operates at the core level, characterizing overall morphology and scar organization as aggregate properties. Finer-grained technological attributes, such as platform preparation, flaking surface maintaining details, are not captured by the current framework. Future study should combine quantitative analyses across multiple scales, progressing from macroscopic to microscopic levels, in order to comprehensively characterize lithic variability and more clearly elucidate hominin technological strategies.

# Conclusion

This study introduces a spherical harmonic framework for the integrated three-dimensional analysis of lithic cores, extending SPHARM beyond near-spherical artifacts to a broader range of morphologically diverse core types. The framework characterizes core morphology (M-SPHARM) and scar pattern (SP-SPHARM) within the same mathematical representation, using von Mises–Fisher kernel density estimation to convert discrete scar vectors into continuous spherical distributions amenable to harmonic expansion.

Validation against ideal digital cores and experimental knapping cores demonstrates that SP-SPHARM resolves finer distinctions than existing approaches, most notably separating radial core subtypes such as Levallois and discoidal that neither SPI nor fabric metrics can resolve. The three methods nonetheless capture different aspects of scar organization and are best used in combination: SP-SPHARM for overall structural comparison, SPI and fabric metrics for interpretively grounded refinement.

Building on these descriptors, we further propose a Co-inertia Analysis framework as a tentative companion approach for jointly examining technological and morphological variation. Applied to experimental and archaeological assemblages, this exploratory framework reveals no detectable covariation between core morphology and scar organization at the assemblage level, a decoupling that held broadly across core types, raw materials, and stratigraphic layers. These findings should be regarded as preliminary, but they illustrate the analytical possibilities opened up by treating morphology and scar pattern as separate yet jointly analyzable dimensions.

To support broader adoption of the framework, we provide an open-source R package (spharmlithic) implementing the essential analytical methods developed in this study, including model alignment, SPI and fabric computation, SPHARM analysis of both morphology and scar patterns, and three-dimensional visual reconstruction. The package, together with the version-controlled repository containing the complete analysis scripts and data for this paper, enables both independent reproduction of the reported results and application of the methods to new datasets.

Future work should focus on expanding sample sizes and assemblage diversity, and integrating SPHARM with phylogenetic and other emerging quantitative paradigms to investigate broader questions of the technological behavior of past humans.

# CRediT authorship contribution statement

# Data availability statement

# Funding

# Declaration of competing interest

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```
