---
title: "New methods based on SPHARM in lithic analysis"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
  - Hao Li:
      correspondence: "yes"
      orcid: 
      institute:
        - ALPHA
        - UCAS
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - UCAS: University of Chinese Academy of Sciences, Beijing 100049, China
  - ALPHA: Group of Alpine Paleoecology and Human Adaptation (ALPHA), Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing 100101; 
  - TPESRE: State Key Laboratory of Tibetan Plateau Earth System, Resources and Environment (TPESRE), Beijing 100101
  - uw: Department of Anthropology, University of Washington, Seattle, WA, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
number-sections: true
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  
editor_options: 
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

```{r}
#| label: setup
#| include: false

library(targets)
tar_config_set(store = "_targets")

ph <- tar_read(spharm_analysis)$spharm_posthoc
```

# Introduction

Cores preserve the most complete record of reduction strategies among all lithic artifact types, making them indispensable for understanding hominin technological behavior. Traditional analyses have relied primarily on the *chaîne opératoire* framework, employing qualitative observations of scar patterns and core morphology to infer reduction strategies. Over the past two decades, lithic studies have witnessed a marked shift toward quantitative approaches, substantially enhancing the scientific rigor and reproducibility of lithic analysis. These advances include the application of multivariate statistical methods, 2D geometric morphometric analyses based on landmarks or Fourier series, and morphological analyses derived from 3D models.

Despite these developments, existing quantitative studies have largely focused on morphological description, leaving the systematic analysis of flaking direction comparatively underdeveloped. To date, only two studies have addressed this issue directly — @clarkson2006 and @lin2024 — yet both rely on one or two variables to characterize overall scar patterns, inevitably entailing a loss of information. This limitation is evident in their capacity to discriminate certain scar configurations: the SPI method struggles to differentiate discoidal cores from Levallois cores, while the Lin's method has difficulty distinguishing unipolar from bipolar reduction patterns. More broadly, most existing approaches are methodologically fragmented, with variables selected *a priori* to match predefined technological categories. This constrains their applicability across different core types and assemblages from different periods, and ultimately limits their generalizability.

Against this background, the present study proposes a new quantitative analytical framework for cores, designed to address two interrelated problems: (1) to improve methods for analyzing flaking pattern on cores; and (2) to integrate scar pattern analysis and morphological analysis within a unified mathematical framework, supported by multivariate statistical methods. By presenting this comprehensive analytical framework, the study aims to offer a methodological contribution to the quantitative study of core analysis.

# Material and Methods

**2.1 Materials**

Ideal core models: one ideal model per core type, sourced from @lin2024.

Experimental knapping models by Clarkson: 1–22 specimens per core type, 59 specimens in total, sourced from Lin et al. (2024) (awaiting email confirmation).

Archaeological specimens from the Sandinggai site: after excluding incomplete cores and tested cores, 55 specimens remain. Specimens derive from Layers 2, 3, and 4 of the site; however, as only two specimens come from Layer 2, statistical analyses in this study are restricted to Layers 3 and 4. Data sourced from Li et al. (2022).

**2.2 Methods**

**2.2.1 Data Extraction**

Scar orientations were digitized in Geomagic Wrap 2021, and their vector coordinates were exported directly.

**2.2.2 Spherical Harmonic Analysis**

Spherical harmonic analysis represents the three-dimensional extension of elliptic Fourier analysis. Its mathematical foundations... Several studies have already applied spherical harmonic methods to lithic analysis: @sholts2017 used spherical harmonics to quantify three-dimensional asymmetry in Clovis points, while @muller2023 and Ye et al. (2026, in reviewed) applied the method to investigate the morphology of spheroids.

In the present study, we apply this method to scar orientation data for the first time, performing spherical harmonic analysis on core scars. For morphological data... For orientation data, a continuous density surface is first fitted using von Mises–Fisher spherical kernel density estimation based on directional statistics, which is then decomposed into spherical harmonics. The spherical harmonic variance and power spectrum are subsequently computed, and an appropriate spectral degree is selected by inspecting the variance contribution across degrees.

**2.2.3 Data Analysis**

*Rotation Invariance Assessment.* To evaluate whether different scar pattern metrics are rotation-invariant, we employed the Bland–Altman approach. The Bland–Altman analysis is widely used to assess agreement between two sets of measurements and provides a quantitative evaluation through the Bland–Altman plot. In this plot, the x-axis represents the mean of the two measurements, while the y-axis represents their difference. The mean difference is defined as the bias, and the 95% limits of agreement (LoA) are calculated based on this bias. A high level of agreement is inferred when the limits of agreement are narrow, approximately 95% of the data points fall within these limits, and the differences are randomly distributed without systematic trends.

In this study, the Bland–Altman framework was used to test the rotation invariance of three metrics: SPI, fabric analysis (Elongation and Isotropy), and the SPHARM power spectrum. Specifically, each metric was computed under three conditions: (1) using the original scar-vector data, (2) using morphological plane aligned data, and (3) using technological plane aligned data. Pairwise Bland–Altman comparisons were then conducted among these conditions. High agreement between the paired measurements was interpreted as evidence of rotation invariance for the corresponding metric.

*Method comparison.* To evaluate the effectiveness of spherical harmonic analysis for characterizing scar patterns, we compare the classificatory performance of three methods — SPI, Lin's fabric analysis, and spherical harmonics — using the ideal models. Since only one ideal model exists per type, we assess performance using pairwise Euclidean distances: specifically, how far apart different ideal model types are in a standardized feature space. Greater pairwise distances indicate stronger differentiation between types, thereby serving as a measure of each method's discriminatory power. Results are visualized as heatmaps, and mean pairwise distances are calculated for each method.

To evaluate the differences in information captured by each method, we applied Multiple Regression on Distance Matrices (MRM). Conceptually, MRM is equivalent to standard multiple linear regression; however, it operates on the lower-triangular elements of distance matrices as observational units and employs permutation tests instead of parametric inference to account for the non-independence among matrix entries.

The analytical procedure consisted of four steps. First, distance matrices for the three scar pattern metrics were constructed using standardized Euclidean distance. Second, each method was treated in turn as the dependent variable, with the remaining two distance matrices used as predictors, resulting in three separate MRM models. The coefficient of determination (R²) from each model represents the proportion of variance in the focal method’s distance structure explained jointly by the other two methods, whereas 1 − R² reflects the proportion of method-specific (unique) information.

Third, with the SPHARM power spectrum distance matrix as the dependent variable, we constructed a full model including both SPI and Fabric as predictors, along with two reduced models each containing a single predictor. Based on standard variance partitioning, the total variance in SPHARM was decomposed into four components: the unique contribution of SPI, the unique contribution of Fabric, the shared contribution of both predictors, and the unexplained variance. The latter component represents the proportion of information uniquely captured by SPHARM relative to SPI and Fabric.

Finally, the residual matrix was extracted from the full model. Positive residuals indicate specimen pairs for which dissimilarity in SPHARM space exceeds the joint prediction from SPI and Fabric, suggesting that SPHARM captures additional directional distribution differences for these pairs; negative residuals indicate the opposite pattern. By ranking standardized residuals, specimen pairs that most strongly reflect the unique discriminative capacity of SPHARM were identified and subsequently visualized.

*Dimensionality reduction.* Because spherical harmonics describe a nonlinear manifold, linear methods such as PCA are ill-suited to recovering the global structure of spherical harmonic data. We therefore apply UMAP to reduce the dimensionality of the spherical harmonic power spectra.

*Correlation analysis and Mantel test.* Spearman rank correlation coefficients and Mantel tests are used to explore whether systematic associations exist between the technological and morphological characteristics of cores.

*Co-inertia analysis.* This method performs joint dimensionality reduction on two data matrices simultaneously, and is used here to interpret the overall covariation between technological and morphological patterns across cores.

# Results

**3.1 Validity Analysis**

**3.1.1 Rotational Invariance of Scar Pattern Analysis Measures**

```{r}
#| label: fig-rotation-validity
#| fig-cap: "Bland-Altman analysis of rotational invariance. **(A-C)** SPI across three coordinate conditions. **(D-F)** Fabric elongation ratio across three coordinate conditions. **(G-I)** Fabric isotropy ratio across three coordinate conditions. **(J)** SP-SPHARM power spectra bias across three coordinate conditions and random Z-axis perturbation."
#| fig-width: 12
#| fig-height: 14
#| echo: false

tar_read(p_rotational_invariance_validity)
```

To evaluate rotational invariance, SPI, fabric metrics, and SP-SPHARM descriptors were computed in three coordinate systems: the original system, a system aligned to the morphological fitted plane, and one aligned to the technological fitted plane. Bland-Altman analysis was used to assess agreement across coordinates. SPI and fabric metrics showed near-perfect agreement across all three conditions, with biases and 95% limits of agreement (LoA) on the order of 10⁻¹⁶ — well within the range of floating-point numerical precision (@fig-rotation-validity) — confirming that both measures are strictly rotationally invariant. Consequently, coordinate alignment of the core model is not required prior to analysis.

The SP-SPHARM descriptors exhibited limited variation across alignment methods. Bland-Altman analysis revealed small systematic biases in low-order power (*l* = 1–2; 95% LoA = ±0.02), which diminished rapidly with increasing degree; no substantive differences were observed for *l* ≥ 3 (@fig-rotation-validity). To further evaluate this, we applied random rotational perturbations around the Z-axis to technological aligned data and recomputed the power spectra. Differences between perturbed and unperturbed spectra remained within machine precision, indicating that directional uncertainty within the coordinate frame — given a fixed Z-axis — has no effect on the power spectrum.

These results suggest that the SP-SPHARM descriptors is theoretically rotationally invariant by design, and that the systematic bias observed in low-order terms more likely originates from upstream computational steps such as KDE-based interpolation and coordinate mapping. Crucially, when a consistent alignment method is applied uniformly across all specimens, this systematic error affects all specimens equally and does not alter relative differences among them, leaving subsequent statistical comparisons unaffected. All subsequent analyses therefore employ the technological fitted plane for coordinate alignment.

**3.1.2 Validity of the SPHARM Method for Scar Pattern Analysis**

```{r}
#| label: setup-im-comparison
#| include: false

library(here)
library(tidyverse)
library(readxl)

# ---- 公共函数 ----
compute_R <- function(dx, dy, dz) {
  resultant_magnitude <- sqrt(sum(dx)^2 + sum(dy)^2 + sum(dz)^2)
  total_length        <- sum(sqrt(dx^2 + dy^2 + dz^2))
  resultant_magnitude / total_length
}

compute_EI <- function(ux, uy, uz) {
  n      <- length(ux)
  U      <- cbind(ux, uy, uz)
  T_mat  <- (t(U) %*% U) / n
  eig    <- eigen(T_mat, symmetric = TRUE)
  lambda <- sort(eig$values, decreasing = TRUE)
  lambda <- pmax(lambda, 0)
  list(
    E = ifelse(lambda[1] > 1e-10, 1 - lambda[2] / lambda[1], NA_real_),
    I = ifelse(lambda[1] > 1e-10,     lambda[3] / lambda[1], NA_real_)
  )
}

# ---- 读入数据 ----
raw <- read_excel(
  here("analysis/data/raw_data/Scar_orientation_data.xlsx"), sheet = 1)
SPHARM_direction <- read_csv(
  here("analysis/data/derived_data/SPHARM_direction.csv"),
  show_col_types = FALSE)

# ---- IM 标本 SPHARM ----
SPHARM_IM <- SPHARM_direction %>%
  filter(str_starts(ID, "IM_")) %>%
  select(ID, Typology, spectral_entropy, power_l1:power_l20)

# ---- IM 标本 Fabric（R/E/I）----
fabric_IM <- raw %>%
  filter(str_starts(ID, "IM_")) %>%
  mutate(
    dx  = End_X - Start_X,
    dy  = End_Y - Start_Y,
    dz  = End_Z - Start_Z,
    len = sqrt(dx^2 + dy^2 + dz^2)
  ) %>%
  filter(len > 1e-10) %>%
  mutate(ux = dx / len, uy = dy / len, uz = dz / len) %>%
  group_by(ID) %>%
  summarise(
    R = compute_R(ux, uy, uz),
    E = compute_EI(ux, uy, uz)$E,
    I = compute_EI(ux, uy, uz)$I,
    .groups = "drop"
  )

# ---- 合并 ----
df_im <- SPHARM_IM %>%
  left_join(fabric_IM, by = "ID") %>%
  mutate(
    label = ID %>%
      str_remove("^IM_") %>%
      str_replace_all("_", " ") %>%
      str_to_sentence()
  )

# ---- 距离矩阵 ----
id_order <- c(
  "Cylindrical unipolar cortical", "Cylindrical unipolar scarred",
  "Cylindrical bipolar",
  "Conical unipolar cortical",     "Conical unipolar scarred",
  "Discoid",                       "Discoid unifacial",
  "Levallois preferential",        "Levallois convergent",
  "Levallois laminar",
  "Biface",                        "Multiplatform"
)

make_dist_df <- function(X, labels, method_name) {
  d <- as.matrix(dist(scale(X), method = "euclidean"))
  rownames(d) <- colnames(d) <- labels
  as.data.frame(d) %>%
    rownames_to_column("From") %>%
    pivot_longer(-From, names_to = "To", values_to = "distance") %>%
    mutate(method = method_name)
}

labs <- df_im$label

dist_all <- bind_rows(
  make_dist_df(df_im %>% select(R)                 %>% as.matrix(), labs, "SPI"),
  make_dist_df(df_im %>% select(E, I)              %>% as.matrix(), labs, "Fabric"),
  make_dist_df(df_im %>% select(power_l1:power_l4) %>% as.matrix(), labs, "SPHARM")
) %>%
  mutate(
    From   = factor(From,   levels = id_order),
    To     = factor(To,     levels = id_order),
    method = factor(method, levels = c("SPI", "Fabric", "SPHARM"))
  )

# ---- 平均距离 ----
avg_dist <- dist_all %>%
  filter(From != To) %>%
  group_by(method) %>%
  summarise(mean_dist = round(mean(distance), 2), .groups = "drop")

dist_spi    <- avg_dist$mean_dist[avg_dist$method == "SPI"]
dist_fabric <- avg_dist$mean_dist[avg_dist$method == "Fabric"]
dist_spharm <- avg_dist$mean_dist[avg_dist$method == "SPHARM"]
```

```{r}
#| label: fig-im-comparison
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| fig-cap: !expr >
#|   paste0("Pairwise standardised Euclidean distances between ideal core 
#|   models across three methods. Mean distances: SPI = ", dist_spi,
#|   ", Fabric = ", dist_fabric, ", SPHARM = ", dist_spharm, ".")
dist_all_upper <- dist_all %>%
  dplyr::filter(as.numeric(From) < as.numeric(To))
ggplot(dist_all_upper, aes(x = To, y = From, fill = distance)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(
    aes(label = ifelse(From != To, sprintf("%.1f", distance), "")),
    size = 2.2
  ) +
  facet_wrap(~ method, ncol = 3) +
  scale_fill_gradient2(
  low  = "#5C7F71",  
  mid  = "#F5EDDC", 
  high = "#802520",
  midpoint = 2,
  name = "Euclidean\ndistance\n(standardised)"
) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  scale_y_discrete(limits = rev) +
  theme_bw(base_size = 8) +
  labs(x = NULL, y = NULL) +
  theme(
    strip.text       = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "#EBEBEB", color = "#EBEBEB"),
    axis.text        = element_text(size = 6.5)
  )
```

**Test of Ideal models** Using ideal core models, we computed SPI, fabric metrics, and SP-SPHARM descriptors and evaluated their discriminative performance. Heatmaps revealed a progressive increase in discriminative capacity from SPI to fabric metrics to SP-SPHARM descriptors (@fig-im-comparison). This ordering was corroborated by mean standardised Euclidean distance: the SP-SPHARM descriptors yielded the highest mean distance ($\bar{d}_{\mathrm{E,std}}$ = `r dist_spharm`), substantially exceeding those of fabric metrics ($\bar{d}_{\mathrm{E,std}}$ = `r dist_fabric`) and SPI ($\bar{d}_{\mathrm{E, std}}$ = `r dist_spi`). These results demonstrate that, under idealized core scar patterns, SP-SPHARM descriptors most effectively distinguishes among different scar distribution structures.\`

```{r}
#| label: fig-exp-method-compare
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 12
#| echo: false

tar_read(p_exp_method_combined)
```

**Test of experimental cores** To evaluate the sensitivity of each method on empirical data, we computed all three measures for experimental cores and conducted statistical comparisons by core type. All three methods detected significant inter-type differences (@fig-exp-method-compare). Post-hoc comparisons with Holm-Bonferroni correction are presented in @fig-exp-method-compare. The SP-SPHARM descriptors achieved the highest overall discrimination, identifying the greatest number of significant pairwise differences. Fabric metrics distinguished the majority of type pairs but failed to differentiate unidirectional from bidirectional cores. SPI was the most limited, detecting differences only where unidirectional cores were involved.

These results suggest that the three methods are complementary in their capacity. SP-SPHARM descriptors capture the overall structure of scar patterning for most pairwise comparisons, although their discriminative power is reduced for patterns with bipolar symmetry, such as discoidal-bidirectional pairs ($P_{\text{adj}}$ = `r ph$p_dir_disc_bi`). Fabric metrics effectively distinguishes differences between unidirectional and bidirectional cores from Levallois and discoidal cores, but cannot differentiate unidirectional from bidirectional cores ($P_{\text{adj}}$ = `r ph$p_fab_uni_bi`), nor Levallois from discoidal cores ($P_{\text{adj}}$ = `r ph$p_fab_lev_disc`). SPI is most sensitive to the unidirectional type but shows limited discriminative capacity for other types. Notably, although SP-SPHARM descriptors failed to separete discoidal from bidirectional cores due to their similarity structures in SPHARM power spectrum (@fig-im-comparison), this distinction is successfully recovered by fabric metrics ($P_{\text{adj}}$ = `r ph$p_fab_disc_bi`).

**3.2 Joint Analysis of Scar Patterns and Core Morphology in Experimental Knapping Specimens**

The analyses above demonstrate that SP-SPHARM descriptors exhibit broadly valid and significant differences across core types. To explore morphological variation among types, we further performed PERMANOVA tests using M-SPHARM descriptors. No significant differences in overall core morphology were detected across types (R2 = , *P* = ), suggesting that core morphology and reduction strategies may constitute two independent dimensions of variation. To test this hypothesis, we integrated morphological and SP-SPHARM descriptors using Mantel tests and Co-Inertia Analysis (CoIA). The global Mantel test revealed no significant correlation between the two descriptor sets, whereas the RV coefficient derived from CoIA was statistically significant (RV = , *P* = ). These results indicate that, at the level of pairwise specimen comparisons, morphological similarity does not consistently correspond to scar pattern similarity, and vice versa. However, at the level of the overall multivariate structure, core morphology and scar patterning exhibit significant covariation.

To rule out the possibility that global structure might obscure local signals, we conducted within-group Mantel tests stratified by core type. A significant positive correlation between morphology and scar patterning was found for discoidal cores (Mantel's r = , *P* = ), indicating that within this type, cores with more similar morphologies tend to produce more similar scar direction distributions. No significant correlations were observed for the remaining types (Levallois: r = 0.047; Unidirectional: r = −0.179; Multiplatform: r = −0.183; Bidirectional: r = −0.100; all *P* \> 0.05). Notably, the negative r values observed for unidirectional, multiplatform, and bidirectional cores suggest that morphological similarity does not predict directional similarity within these groups, consistent with greater operational flexibility or reduced dependence on core geometry in these reduction strategies.

The CoIA results provide further detail on the covariation between morphology and scar patterns. The first two CoIA axes together account for [X]% of total covariation (Axis 1 = [X]%, Axis 2 = [X]%; Figure X). Axis 1 primarily contrasts cores with elongated morphologies and longitudinally organised scars against those with flatter forms and more centripetal scar arrangements. Axis 2 captures residual variation in scar asymmetry and lateral organisation: on the scar pattern side, it distinguishes bidirectional  removals from unidirectional flaking pattern; on the morphology side, it reflects differences in core thickness relative to width — that is, how plump or flat the core is in cross-section (Figure X).

In the CoIA biplot, each specimen's morphological and scar pattern SPHARM spectra are mapped into the same feature space; the length of the line connecting the two projections reflects the degree of coupling between morphology and scar pattern, with shorter lines indicating closer correspondence. A Kruskal-Wallis test revealed no significant difference in line lengths across core types (Fig. X), suggesting that the overall degree of morphology–scar coupling is broadly comparable across reduction strategies. The spatial orientation of the lines reflects the direction of techno-morphological covariation within each type: concentrated orientations imply a consistent axis of joint variation, potentially indicative of an underlying operational schema. Rayleigh tests identified significant directional concentration in three core types — Bidirectional (mean direction = 112°, p = 0.004), Discoid (mean direction = 126°, p = 0.035), and Levallois (mean direction = 299°, p = 0.026) — while Unidirectional and Multiplatform cores showed no significant concentration (Fig. X). Watson two-sample tests further revealed that Levallois line orientations differ significantly from those of Bidirectional, Discoid, and Multiplatform cores (all p < 0.05), confirming that Levallois cores occupy a distinct techno-morphological trajectory in CoIA space. Given the interpretation of the CoIA axes, the Levallois mean direction of 299° projects primarily along the negative end of Axis 1, indicating that Levallois technology is consistently associated with elongated core morphology and longitudinally organised scar patterns. The convergence of Bidirectional and Discoid mean directions (~112–126°) along the positive region of Axis 2 suggests that these two types share a broadly similar techno-morphological trajectory despite their typological differences.

To quantify the structural complexity of the power spectra, we computed spectral entropy for each specimen. Kruskal-Wallis tests showed that the scar pattern spectral entropy of Levallois cores was significantly lower than that of all other types except Unidirectional cores, indicating that Levallois scar pattern exhibits the highest structural regularity. Read in conjunction with the CoIA results, this suggests that while Levallois reduction sequences are organizationally complex, they conform to a highly structured set of operational rules. By contrast, morphological spectral entropy did not differ significantly across types, indicating no systematic differences in morphological complexity among core types (Fig. X).

**3.3 Archaeological Case Study: Core Analysis from Sandinggai**

Based on the analytical framework outlined in 3.2, we conducted a joint analysis of core morphology and scar patterns on 51 cores excavated from Sandinggai.

The global Mantel test revealed no significant pairwise distance correlation between the morphological spectrum and the scar pattern spectrum (r = 0.051, p = 0.254), indicating that core morphology and scar patterns are not significantly similar at the level of specimen pairs. Cross-frequency Mantel tests remained non-significant after Holm-Bonferroni correction, confirming that this independence holds consistently across all frequency bands. The RV coefficient from CoIA likewise failed to reach significance (RV = 0.050, p = 0.794), indicating that no systematic covariation exists between the two dimensions at the level of overall multivariate structure either.

To rule out the possibility that global structure masks local signals, Mantel tests were further conducted within subgroups stratified by layer, raw material, and core type. None of the subgroups showed significant covariation after Holm-Bonferroni correction (Table X), indicating that the independence between morphology and scar patterns holds across different time periods, raw materials, and core types.

The CoIA dimensionality reduction results provide further detail on the covariation between morphology and scar patterns. The first two CoIA axes together account for 90.6% of total covariation (Axis 1 = 57.6%, Axis 2 = 33.0%; Figure X). Axis 1 primarily contrasts cores with flattened morphologies and unidirectional removals against those with more rounded forms and birectional removals. Axis 2 captures additional variation in both scar pattern structure and morphological detail. On the scar pattern side, it distinguishes unidirectional and bidirectional removals from more centripetal patterns. On the morphology side, it primarily reflects mid-frequency shape variation associated with three-fold and four-fold rotational symmetry, likely corresponding to differences in core cross-sectional geometry (Figure X).

To examine how line features derived from CoIA vary across grouping conditions, we analyzed line length and direction separately. Kruskal–Wallis tests showed no significant differences in line length among layers (χ² = 1.93, p = 0.165), raw materials (χ² = 2.02, p = 0.155), or core types (χ² = 1.30, p = 0.935; Figure X), indicating that the decoupling between morphology and scar patterns is consistent in magnitude across all grouping conditions. Rayleigh tests on line direction showed random distributions for most groups and core types, with the only exception of Unifacial unidirectional cores, which exhibited significantly concentrated directionality (U = 0.826, p = 0.002) with a mean direction of approximately 354°, indicating that their techno–morphology relationship is specifically structured along Axis 1. More specifically, regardless of how individual specimens vary in morphology and scar pattern space, Unifacial unidirectional cores consistently shift toward the flattened end of the morphological spectrum and the unidirectional end of the scar pattern spectrum — a directional tendency that is stable across all specimens of this type.

To quantify the structural complexity of scar direction spectra, we calculated the spectral entropy of scar directions for each specimen. Kruskal–Wallis tests revealed significant differences in scar direction spectral entropy among core types (χ² = 12.782, df = 5, p = 0.026). Dunn post-hoc tests with Holm correction showed that Multifacial cores had significantly higher scar pattern spectral entropy than Unifacial centripetal cores (p.adj = 0.028). These results indicate that Multifacial cores show highly dispersed scar pattern energy, while Unifacial centripetal cores show the most concentrated scar direction energy among all core types. In contrast, morphological spectral entropy showed no significant differences across any grouping variable — layer, raw material, or core type — indicating that differences in flaking choices are primarily expressed in the scar pattern dimension rather than in morphological complexity (Figure X).

To explore the factors driving variation in morphology and scar patterns, PERMANOVA tests were conducted for the three grouping variables. Core type had a significant effect on both morphological variation (R² = 0.172, F = 1.82, p = 0.043) and scar pattern variation (R² = 0.168, F = 1.78, p = 0.035), indicating systematic differences in both dimensions among core types. Raw material had a significant effect on morphology (R² = 0.101, F = 5.51, p = 0.004), though pairwise comparisons between chert and sandstone did not survive Holm correction (p = 0.109). Raw material had no significant effect on scar direction variation (R² = 0.018, p = 0.435), indicating that knappers applied consistent reduction strategies regardless of raw material — the physical properties of raw materials shaped core morphology outcomes but did not influence technological decision-making. Layer had no significant effect on either morphology (R² = 0.013, p = 0.612) or scar pattern (R² = 0.020, p = 0.419), suggesting continuity in reduction strategies across layers, with no clear evidence of technological change.

Taken together, the core assemblage from Sandinggai shows a systematic decoupling between core morphology and reduction strategy. This decoupling is not limited to any specific condition but appears as a general pattern across different time periods, raw materials, and core types. Morphological variation is primarily associated with raw material and core type, while scar direction variation shows no significant response to these factors. This suggests that reduction strategies may be relatively stable and largely independent of changes in raw material or core type — possibly reflecting a consistent technological tradition maintained by knappers throughout the core reduction process.

# Discussion

# Conclusion

# CRediT authorship contribution statement

# Data availability statement

# Funding

# Declaration of competing interest

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```
