---
title: "Characterizing core scar patterning and morphology within a unified spherical harmonic framework"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
  - Hao Li:
      correspondence: "yes"
      orcid: 
      institute:
        - ALPHA
        - UCAS
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - UCAS: University of Chinese Academy of Sciences, Beijing 100049, China
  - ALPHA: Group of Alpine Paleoecology and Human Adaptation (ALPHA), Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing 100101; 
  - TPESRE: State Key Laboratory of Tibetan Plateau Earth System, Resources and Environment (TPESRE), Beijing 100101
  - uw: Department of Anthropology, University of Washington, Seattle, WA, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
number-sections: true
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  
editor_options: 
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

```{r}
#| label: setup
#| include: false
library(targets)
library(tidyverse)
tar_config_set(store = "/project/analysis/paper/_targets")
sa          <- tar_read(spharm_analysis)
ph          <- sa$spharm_posthoc
im          <- tar_read(im_comparison)
dist_spi    <- im$dist_spi
dist_fabric <- im$dist_fabric
dist_spharm <- im$dist_spharm
cia         <- tar_read(exp_cia_analysis)
sdg         <- tar_read(sdg_cia_analysis)
```

# Introduction

Cores preserve the most direct evidence of reduction strategies in lithic assemblages. Their analysis informs not only knapping behavior, including flaking strategies, knapping skill levels, and technological traditions, but also broader questions of hominin cognition, mobility patterns, site function, and social learning.

Core analysis has traditionally relied on typological approaches, in which cores are classified according to hierarchically organized attributes such as core morphology, scar patterning, flaking angles, and platform characteristics into categories such as Levallois, discoidal, and laminar cores. These frameworks, however, are far from standardized: different typological systems often employ distinct descriptive vocabularies, and substantial classification inconsistencies arise both between analysts and within the same analyst over time. The long-standing debate over Levallois technology in China illustrates this problem clearly. Although researchers commonly apply the criteria proposed by Boëda, disagreements persist over whether certain cores from the Shiyu and Guanyindong sites should be classified as Levallois. Efforts to establish unified classificatory systems, such as the Unified Lithic Taxonomy and the Logic Analytical System, have partially reduced inter-observer variability, yet the fundamental reliance on qualitative judgment remains.

Growing demands for precision and reproducibility have driven increasing interest in three-dimensional quantitative approaches to lithic analysis. 3D modeling technologies provide substantially higher-resolution data than traditional linear measurements and have been widely applied to flakes, retouched tools, and large cutting tools (LCTs). Studies specifically focused on cores, however, remain limited. Two broad approaches characterize existing work. The first extracts predefined technological variables from 3D models, such as platform angles, cortex proportions, reduction intensity, and scar patterning. The second characterizes overall core morphology, most commonly through landmark-based geometric morphometrics, though recent studies have also applied spherical harmonic methods to stone balls and polyhedral cores. These two approaches are complementary, and their integration has yielded deeper insights in several studies.

These approaches, however, have notable limitations. Landmark-based methods compress continuous morphological variation into a limited number of discrete points, inevitably losing information. They also rely on homology assumptions: current landmark configurations, which typically place landmarks along core edges and semi-landmarks across surfaces, are mainly applicable to highly standardized systems such as Levallois reduction and difficult to extend to more variable types or to cross-type comparisons. Even within Levallois core studies, landmark configurations vary considerably between researchers, making direct comparisons between morphospaces problematic. Spherical harmonic (SPHARM or SH) approaches offer a promising alternative. They do not require homology assumptions, capture morphological information more comprehensively, and produce rotationally invariant power spectra that eliminate the need for manual core orientation. To date, however, their application has been largely restricted to near-spherical types such as spheroids and polyhedral cores, and their potential for more morphologically diverse core types has yet to be tested.

Quantitative approaches to scar pattern analysis are less developed. Only two systematic frameworks exist: the Scar Pattern Index (SPI), a univariate metric measuring directional concentration through the ratio of resultant to total vector length, and fabric analysis, which uses eigenvalues of orientation tensors to provide a high-level summary of scar orientation patterns. Both represent important advances, yet both reduce complex scar-vector configurations to a small number of summary parameters that capture only coarse-grained orientation organization rather than the full spatial structural properties of scar patterning itself. This information loss is reflected in their analytical performance: SPI and fabric approaches effectively distinguish broad technological organizations, such as unidirectional or bidirectional removals and Levallois or discoidal cores, but struggle to differentiate variability within these categories. A method capable of encoding the full spatial structure of scar organization, rather than reducing it to summary statistics, is therefore needed.

Beyond these individual dimensions, the relationship between scar organization and core morphology has long constituted a central problem in core analysis. Previous studies have demonstrated that knappers actively structured flaking sequences to produce and maintain particular morphologies throughout reduction. At the same time, core morphology may be strongly constrained by objective factors such as raw material properties and reduction intensity, rather than determined solely by technological intention. Yet no quantitative method currently exists that can simultaneously characterize technological organization and core morphology within a unified framework while directly testing their covariation. This gap has hindered efforts to evaluate the relative contributions of technological intention and objective constraints within core reduction sequences.

The present study systematically explores the application of spherical harmonic methods to core analysis and develops an integrated quantitative framework linking technology and morphology. For morphological analysis, we evaluate the applicability of spherical harmonics across diverse core types rather than restricting the method to near-spherical forms. For scar pattern analysis, we introduce a new workflow that transforms discrete scar vectors into continuous spherical distributions via kernel density estimation and then subjects them to spherical harmonic expansion, enabling scar organization to be represented within the same mathematical framework as core morphology. Building on this integration, we employ Co-inertia Analysis (CoIA) to examine technological–morphological covariation, and exploit the reconstructive capacity of spherical harmonics to link statistical patterns directly to interpretable three-dimensional shapes. In this way, variation among cores is no longer treated as a static arrangement of typological categories, but can be understood as part of a continuous spectrum of technological and morphological variability. We first validate the method using idealized digital core models and an experimental knapping dataset, then apply the framework to excavated cores from the Sandinggai site in Hunan Province, China, to demonstrate its practical interpretive potential.

# Material and Methods

**2.1 Materials**

Three datasets were used in this study, each serving a distinct analytical role. The Ideal Digital Cores dataset was used to evaluate the sensitivity of spherical harmonic methods to standardized scar patterns. The Experimental Knapping Cores dataset was employed to test the scar-pattern analysis framework under more realistic conditions of morphological and technological variability, and to evaluate the integrated statistical framework proposed in this study. The Archaeological Specimens dataset was then analyzed within this validated framework to demonstrate its practical archaeological applicability.

[Ideal digital cores]{.underline} Eight idealized digital core models, sourced from Sam Lin, were used, comprising cylindrical, conical, discoidal, Levallois (preferential, convergent, and laminar), bifacial, and polyhedral types. Each category is associated with distinct scar organizations, described in detail in the supplementary materials.

[Experimental knapping cores]{.underline} The experimental cores were produced by Chris Clarkson and comprise 58 specimens in this study: 16 unidirectional, 6 bidirectional, 21 Levallois (10 preferential, 5 convergent, 2 Nubian, 2 recurrent, and 2 laminar), 5 discoidal, and 10 multiplatform cores. These specimens are curated at the School of Social Sciences, University of Queensland. The cores were digitized by Sam Lin using a Polyga HDI Compact C210 structured-light scanner with FlexiScan3D software.

[Archaeological specimens]{.underline} The archaeological specimens derive from the Sandinggai site (hereafter SDG) in Hunan Province, China, previously reported by Li. The SDG sequence comprises three cultural layers (L2, L3, and L4), dated by optically stimulated luminescence (OSL) to approximately 100–20 ka. The lithic assemblage is characterized by two operational sequences: (1) a small-flake system primarily based on chert, and (2) a large-flake and large cutting tool (LCT) system primarily based on quartz sandstone.

From the previously reported core assemblage, 51 complete and well-preserved specimens retaining clear technological information were selected for analysis, after excluding incomplete, heavily abraded, and test cores. These derive from Layers 2, 3, and 4; however, because only two cores originate from Layer 2, statistical analyses are restricted to Layers 3 and 4. Core classifications follow the original paper, and details of core types and raw materials are provided in Table X.

Archaeological cores were digitized using an EinScan Pro XS handheld 3D scanner (SHINING 3D) with a turntable and fixed camera stand. Model alignment and export were completed using the accompanying EinScan-Pro software.

For both experimental and archaeological datasets, post-processing was conducted in Geomagic Wrap, including the filling of minor holes and the correction of mesh defects such as non-manifold edges and self-intersections.

**2.2 Methods**

**2.2.1 Data Extraction**

Scar orientations were recorded following the protocol established by Lin. Only scars larger than 5 mm were included. Each scar was recorded as a vector defined by the three-dimensional coordinates of its start and end points, using Geomagic Wrap 2021 for data acquisition. A best-fit plane was also calculated for each core model, and the coordinates of the plane centroid and normal vector were exported for model alignment.

Two alignment procedures were employed to eliminate initial phase differences among core models. The first follows Lin, aligning models based on their best-fit planes by rotating the plane normal to coincide with the Z-axis and translating the starting point of the longest scar to the global origin. The second is proposed here for the first time and relies exclusively on the spatial organization of scar data: (1) unit vectors are calculated for all scars and subjected to singular value decomposition (SVD); (2) the third singular vector, representing the direction least correlated with overall scar orientation structure, is selected as the alignment-plane normal and rotated to the Z-axis; (3) the global centroid of all scar coordinates is translated to the origin; and (4) the first singular vector, representing the principal direction of scar organization, is aligned with the X-axis. The original vectors and the vectors aligned under both procedures were then used to evaluate the rotational invariance of the scar-pattern metrics described below.

**2.2.2 Spherical Harmonic Analysis**

Spherical harmonic analysis represents the three-dimensional extension of elliptic Fourier analysis. The method characterizes surfaces of closed genus-zero objects by mapping each vertex onto the unit sphere through equivalent coordinate functions and representing the surface as a weighted sum of spherical harmonic basis functions. The resulting shape descriptors are invariant to rotation, translation, and scaling, and can be used to reconstruct the original three-dimensional morphology directly. SPHARM was originally developed for morphometric analysis of brain structures in medical imaging and has since been applied across multiple disciplines. In lithic studies, Sholts used spherical harmonics to quantify asymmetry in Clovis points, while Muller and Ye applied the method to spheroid morphology. Beyond morphological characterization, mathematical relationships between spherical harmonic representations and directional data have also been established. In the present study, we exploit this relationship by applying SPHARM to scar orientation data for the first time.

The analytical workflow for scar pattern analysis proceeds as follows. Discrete scar vectors are first transformed into a continuous density distribution on the sphere using von Mises–Fisher spherical kernel density estimation (VMF-sKDE):

$$f(\mathbf{g}) = \frac{1}{n} \sum_{i=1}^{n} \exp\left(\kappa \, \mathbf{g} \cdot \mathbf{x}_i\right)$$

where $\mathbf{g}$ is a unit vector on the spherical grid, $\mathbf{x}_i$ is the unit vector representing the orientation of the i-th scar, and $\mathbf{κ}$ is the concentration parameter of the von Mises–Fisher kernel, related to the bandwidth $\mathbf{h}$ by:

$$\kappa = \frac{1}{h^2}$$

A bandwidth of 0.35 ($\mathbf{κ}$ ≈ 8.16) was selected to balance directional resolution against oversmoothing. This choice was validated by reconstructing the original distributions, in which distinct scar-pattern organizations remained clearly identifiable.

The KDE distributions were interpolated onto a 64×128 Driscoll–Healy grid and subjected to spherical harmonic expansion. Each distribution $f(\theta, \phi)$ is represented as a weighted combination of spherical harmonic basis functions $Y_l^m(\theta, \phi)$:

$$f(\theta, \phi) = \sum_{l=0}^{\infty} \sum_{m=-l}^{l} \hat{f}(l, m) \, Y_l^m(\theta, \phi)$$

where the coefficients $\hat{f}(l, m)$ quantify the contribution of each basis function, and $\mathbf{l}$ and $\mathbf{m}$ denote harmonic order and degree, respectively. A truncation threshold of $l_{\max} = 20$ was adopted in our study, consistent with previous studies demonstrating that this order sufficiently captures detailed artifact morphological variation. The resulting coefficients retain phase information and can be used to reconstruct original three-dimensional forms, but phase sensitivity may introduce noise in comparative analyses. Accordingly, scar pattern spherical harmonic (hereafter SP-SPHARM) power spectra were also calculated by summing squared coefficients within each harmonic order and normalizing to unit total, providing a phase-independent, rotationally invariant representation of scar pattern structure.

For core morphological analysis, we followed the procedures established by Ye. Three-dimensional models (.stl) were downsampled to 20,000 faces, and Laplacian smoothing was applied to reduce surface noise and regularize vertex distribution. Centroid normalization and principal-axis alignment were then performed to position all vertices within a common coordinate system, which was transformed into spherical coordinates for harmonic expansion. Spherical harmonic coefficients and morphological spherical harmonic (hereafter M-SPHARM) power spectra were calculated at $l_{\max} = 20$, as for the scar-pattern analysis. The complete workflow is illustrated in Figure X.

**2.2.3 Data Analysis**

[Rotation Invariance Assessment.]{.underline} Rotational invariance of three scar-pattern metrics (SPI, fabric Elongation and Isotropy, and SP-SPHARM power spectra) was assessed using the Bland–Altman approach. This method evaluates agreement between paired measurements by plotting their difference against their mean. The mean difference represents systematic bias, and the 95% limits of agreement (LoA) define the range within which 95% of differences are expected to fall; narrow LoA with no systematic trends indicate high agreement. Each metric was computed under three coordinate conditions: (1) original scar-vector data, (2) morphological-plane alignment, and (3) technological-plane alignment. Pairwise Bland–Altman comparisons among these conditions were conducted, with high agreement interpreted as evidence of rotational invariance.

The effectiveness of different scar pattern metrics was evaluated using both datasets. Because SP-SPHARM power spectra comprise 20 harmonic orders, dimensional screening was first performed based on coefficients of variation and cumulative power contributions to reduce noise and mitigate the curse of dimensionality.

For the ideal digital cores, where only one model exists per type, discriminatory power was assessed using pairwise standardized Euclidean distances in feature space, with greater distances indicating stronger inter-type separation. Results were visualized using heatmaps. For the experimental knapping cores, a significance-testing framework was employed. SPI, as a univariate metric, was evaluated using the Kruskal–Wallis test. Fabric metrics and SP-SPHARM descriptors were analyzed using PERMANOVA, which tests overall structural differences between groups, complemented by PERMDISP to assess whether differences in within-group dispersion contribute to the observed separation. Significant PERMDISP results warrant cautious interpretation of PERMANOVA outcomes. All permutation procedures used 9,999 permutations. Where overall tests were significant, post hoc pairwise comparisons were performed with Holm–Bonferroni correction.

[Techno-morphological statistical analytical framework.]{.underline} To investigate techno–morphology relationships, we employed a two-step analytical framework: first testing whether overall covariation exists between M-SPHARM and SP-SPHARM power spectra, then exploring interaction patterns through joint dimensional reduction.

*Isometric log-ratio transformation.* Spherical harmonic power spectra are compositional data: the total power is constrained to a fixed sum, so variation in one harmonic order necessarily induces spurious covariance among others. To eliminate this closure effect, Isometric log-ratio transformation (ILR) was applied to all power spectrum data, projecting them into Euclidean space through log-ratio calculations among harmonic orders. All subsequent statistical analyses were conducted on ILR-transformed data.

*Mantel test and RV coefficient.* Overall covariation was evaluated using two complementary approaches. The Mantel test assesses correlation between two distance matrices via Spearman's rank correlation coefficient, with significance determined by 9,999 permutations. The RV coefficient quantifies the overall degree of association between two multivariate datasets, ranging from 0 (no association) to 1 (identical structure), with significance likewise assessed by permutation. The Mantel test emphasizes pairwise distance relationships, whereas the RV coefficient focuses on multivariate covariance structure; their combined use therefore provides a more robust assessment of dataset correspondence.

*Co-inertia analysis.* Co-inertia Analysis (CoIA) is a multivariate method for evaluating structural relationships between two datasets. Its core principle is to identify directions of variation along which the two datasets are maximally concordant while maximizing their covariance. PCA is first performed separately on each dataset, after which projections are used to identify correspondences among samples and variables across the two data spaces. CoIA is particularly well suited for comparing datasets that differ in physical meaning but share identical sampling units, making it appropriate for jointly analyzing M-SPHARM and SP-SPHARM power spectra, which describe different aspects of core variability across the same set of specimens.

The CoIA ordination axes represent the principal directions of shared variation between the two datasets. Each sample is represented by two points, one from each dataset, connected by a line segment. Line length reflects the degree of discrepancy between a sample's position in the two data spaces: shorter lines indicate stronger correspondence between its morphological and scar pattern signatures, whereas longer lines indicate that the two dimensions characterize that specimen differently. Line direction reflects the trend of covariation for that sample within the shared space. If line directions are broadly consistent across a group of samples, this suggests a stable and internally coherent covariation structure; dispersed directions suggest weaker structural correspondence or the presence of localized variation. Combined with the reconstructive capacity of spherical harmonics, this framework allows the morphological and scar-pattern characteristics associated with each end of the CoIA axes to be directly visualized as three-dimensional forms, thereby reconnecting statistical patterns to concrete archaeological interpretation.

To further characterize sample correspondences, we employed Kruskal–Wallis tests to compare line lengths among groups, assessing whether the degree of technology–morphology coupling differs across core types. Rayleigh tests were applied to line directions to evaluate whether orientations exhibit significant directional clustering. Because line directions in CoIA represent axial correspondences rather than unidirectional vectors, they were treated as axial data, with directions differing by 180° considered equivalent. The combined analysis of line lengths and directions enables simultaneous evaluation of coupling strength and structural consistency from both distance-based and directional perspectives.

*Reconstruction along CoIA axes.* To visualize the coupled techno–morphological variation captured by CoIA, we developed a continuous reconstruction framework operating directly in spherical harmonic coefficient space. Spherical harmonic coefficients of individual specimens were projected onto the CoIA axes, and at evenly spaced positions along each axis, reconstructed forms were generated by Gaussian kernel-weighted averaging of neighboring specimens' coefficients. The weighted coefficients were then transformed back into three-dimensional meshes (M-SPHARM) or spherical density surfaces (SP-SPHARM) through inverse spherical harmonic reconstruction. The resulting forms represent localized structural tendencies along covariance gradients rather than static group averages, enabling gradual transitions in morphology and scar organization to be visualized as continuous trajectories within the shared CoIA space.

**2.2.4 Reproducibility and open source materials**

Computational reproducibility remains an ongoing challenge in archaeological reserch, yet it is essential for the credibility and cumulative progress of quantitative research. To support reproducibility and re-use, all materials necessary to reproduce the analyses presented here are provided in a version-controlled online repository at XX. This includes the complete R and Python scripts used for all analyses and visualizations, the raw datasets underlying all reported results, and an R package that implements the analytical workflows proposed in this study for application to other datasets. The repository is organized following established best practices for research compendia, with a logical folder structure and relative file paths that allow all code to be executed without manual modification. All figures, tables, and statistical results presented in this paper can therefore be independently reproduced from these materials.

# Results

**3.1 Validity Analysis**

**3.1.1 Rotational Invariance of Scar Pattern Analysis Measures**

To evaluate rotational invariance, SPI, fabric metrics, and SP-SPHARM descriptors were computed in three coordinate systems: the original system, a system aligned to the morphological fitted plane, and one aligned to the technological fitted plane. Bland-Altman analysis was used to assess agreement across coordinates. SPI and fabric metrics showed near-perfect agreement across all three conditions, with biases and 95% limits of agreement (LoA) on the order of 10⁻¹⁶ — well within the range of floating-point numerical precision (@fig-rotation-validity) — confirming that both measures are strictly rotationally invariant. Consequently, coordinate alignment of the core model is not required prior to analysis.

```{r}
#| label: fig-rotation-validity
#| fig-cap: "Bland-Altman analysis of rotational invariance. **(A-C)** SPI across three coordinate conditions. **(D-F)** Fabric elongation ratio across three coordinate conditions. **(G-I)** Fabric isotropy ratio across three coordinate conditions. **(J)** SP-SPHARM power spectra bias across three coordinate conditions and random Z-axis perturbation."
#| fig-width: 12
#| fig-height: 14
#| echo: false
#| dpi: 150
tar_read(p_rotational_invariance_validity)
```

The SP-SPHARM descriptors exhibited limited variation across alignment methods. Bland-Altman analysis revealed small systematic biases in low-order power (*l* = 1–2; 95% LoA = ±0.02), which diminished rapidly with increasing degree; no substantive differences were observed for *l* ≥ 3 (@fig-rotation-validity). To further evaluate this, we applied random rotational perturbations around the Z-axis to technological aligned data and recomputed the power spectra. Differences between perturbed and unperturbed spectra remained within machine precision, indicating that directional uncertainty within the coordinate frame — given a fixed Z-axis — has no effect on the power spectrum.

These results suggest that the SP-SPHARM descriptors is theoretically rotationally invariant by design, and that the systematic bias observed in low-order terms more likely originates from upstream computational steps such as KDE-based interpolation and coordinate mapping. Crucially, when a consistent alignment method is applied uniformly across all specimens, this systematic error affects all specimens equally and does not alter relative differences among them, leaving subsequent statistical comparisons unaffected. All subsequent analyses therefore employ the technological fitted plane for coordinate alignment.

**3.1.2 Validity of the SPHARM Method for Scar Pattern Analysis**

**Test of Ideal models** Using ideal core models, we computed SPI, fabric metrics, and SP-SPHARM descriptors and assessed how well each method separates different patterns in feature space. Heatmaps revealed a progressive increase in inter-type separation from SPI to fabric metrics to SP-SPHARM descriptors (@fig-im-comparison). This ordering was corroborated by mean standardised Euclidean distance: the SP-SPHARM descriptors yielded the greatest inter-type distance ($\bar{d}_{\mathrm{E,std}}$ = `r dist_spharm`), substantially exceeding those of fabric metrics ($\bar{d}_{\mathrm{E,std}}$ = `r dist_fabric`) and SPI ($\bar{d}_{\mathrm{E, std}}$ = `r dist_spi`). These results indicate that SP-SPHARM descriptors produce the clearest separation among idealized scar patterns.\`

```{r}
#| label: fig-im-comparison
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| fig-cap: !expr >
#|   paste0("Pairwise standardised Euclidean distances between ideal core 
#|   models across three methods. Mean distances: SPI = ", dist_spi,
#|   ", Fabric = ", dist_fabric, ", SPHARM = ", dist_spharm, ".")

im$p_heatmap
```

**Test of experimental cores** To evaluate the sensitivity of each method on empirical data, we computed all three measures for experimental cores and conducted statistical comparisons by core type. All three methods detected significant inter-type differences (@fig-exp-method-compare). Post-hoc comparisons with Holm-Bonferroni correction are presented in @fig-exp-method-compare. The SP-SPHARM descriptors achieved the highest overall discrimination, identifying the greatest number of significant pairwise differences. Fabric metrics distinguished the majority of type pairs but failed to differentiate unidirectional from bidirectional cores. SPI was the most limited, detecting differences only where unidirectional cores were involved.

```{r}
#| label: fig-exp-method-compare
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 12
#| echo: false

tar_read(spharm_analysis)$p_exp_method_combined
```

These results suggest that the three methods are complementary. SP-SPHARM descriptors capture the overall structure of scar patterning for most pairwise comparisons, though separation is reduced for patterns with bipolar symmetry, such as discoidal-bidirectional pairs ($P_{\text{adj}}$ = `r ph$p_dir_disc_bi`). Fabric metrics effectively distinguishes differences between unidirectional and bidirectional cores from Levallois and discoidal cores, but cannot differentiate unidirectional from bidirectional cores ($P_{\text{adj}}$ = `r ph$p_fab_uni_bi`), nor Levallois from discoidal cores ($P_{\text{adj}}$ = `r ph$p_fab_lev_disc`). SPI is most sensitive to the unidirectional type but shows limited discriminative capacity for other types. Notably, although SP-SPHARM descriptors failed to separete discoidal from bidirectional cores due to their similarity structures in SPHARM power spectrum (@fig-im-comparison), this distinction is successfully recovered by fabric metrics ($P_{\text{adj}}$ = `r ph$p_fab_disc_bi`). PERMDISP tests further revealed that discoidal and multiplatform cores exhibit greater within-group dispersion than other types, suggesting that these core types may be operationally more variable. It should also be noted that bidirectional (n = 6) and discoidal (n = 5) are represented by considerably fewer specimens than other types, which limits statistical power for comparisons involving these groups.

**3.2 Joint Analysis of Scar Patterns and Core Morphology in Experimental Knapping Specimens**

The analyses above demonstrate that SP-SPHARM descriptors exhibit broadly valid and significant differences across core types. To explore morphological variation among types, we further performed PERMANOVA tests using M-SPHARM descriptors. No significant differences in overall core morphology were detected across types ($R^2$ = `r sa$perm_morph_r2`, $P$ = `r sa$perm_morph_p`), suggesting that core morphology and reduction strategies may vary along largely decoupled axes at the assemblage level. To test this hypothesis, we integrated morphological and SP-SPHARM descriptors using Mantel tests and Co-Inertia Analysis (CoIA). The global Mantel test revealed no significant correlation between the two descriptor sets (Mantel's $r$ = `r cia$mantel_global_r`, $P$ = `r cia$mantel_global_p`), and the RV coefficient derived from CoIA likewise did not reach significance ($RV$ = `r cia$rv`, $P$ = `r cia$rv_p`).

The CoIA results provide further detail on the covariation between morphology and scar patterns. The first two CoIA axes together account for 88.8% of total covariation (Axis 1 = 70.3%, Axis 2 = 18.5%; @fig-exp-cia). Axis 1 primarily contrasts cores with elongated morphologies and longitudinally scars against those with flatter shape and more centripetal scar arrangements. Axis 2 captures residual variation in scar asymmetry and lateral organisation: on the scar pattern side, it distinguishes bidirectional removals from unidirectional flaking pattern; on the morphology side, it reflects differences in core thickness relative to width — that is, how plump or flat the core is in cross-section (@fig-exp-cia).

In the CoIA biplot, each specimen's morphological and scar pattern SPHARM spectra are mapped into the same feature space; the length of the line connecting the two projections reflects the degree of coupling between morphology and scar pattern, with shorter lines indicating closer correspondence. A Kruskal-Wallis test revealed no significant difference in line lengths across core types (@fig-exp-cia), suggesting that the overall degree of morphology–scar coupling is broadly comparable across reduction strategies. The orientation of lines reflects the direction of techno-morphological covariation within each type: concentrated orientations imply a consistent axis of joint variation, potentially indicative of an underlying operational schema. Rayleigh tests identified significant directional concentration in three core types — bidirectional (mean direction = `r cia$mean_dir_bi`°, $P$ = `r cia$rayleigh_bi_p`), discoidal (mean direction = `r cia$mean_dir_disc`°, $P$ = `r cia$rayleigh_disc_p`), and Levallois (mean direction = `r cia$mean_dir_lev`°, $P$ = `r cia$rayleigh_lev_p`) — while unidirectional and multiplatform cores showed no significant concentration (@fig-exp-cia). Watson two-sample tests further revealed that Levallois line orientations differ significantly from those of bidirectional, discoidal, and multiplatform cores (all $P$ \< 0.05), confirming that Levallois cores occupy a distinct techno-morphological trajectory in CoIA space. Given the interpretation of the CoIA axes, the Levallois mean direction of 299° projects primarily along the negative end of Axis 1, indicating that Levallois technology is consistently associated with elongated core morphology and longitudinally organised scar patterns. The convergence of bidirectional and discoidal mean directions (\~112–126°) along the positive region of Axis 2 suggests that these two types share a similar techno-morphological trajectory despite their typological differences.

```{r}
#| label: fig-exp-cia
#| echo: false
#| fig-width: 10
#| fig-height: 14
#| fig-cap: "CoIA biplot, line length, and directional distribution of techno-morphological coupling in experimental cores."

cia$p_final
```

**3.3 Archaeological Case Study: Core Analysis from Sandinggai**

Based on the analytical framework outlined in 3.2, we conducted a joint analysis of core morphology and scar patterns on 51 cores excavated from Sandinggai.

The global Mantel test revealed no significant pairwise distance correlation between the morphological spectrum and the scar pattern spectrum (r = `r sdg$mantel_global_r`, p = `r sdg$mantel_global_p`), indicating that core morphology and scar patterns are not significantly similar at the level of specimen pairs. Cross-frequency Mantel tests remained non-significant after Holm-Bonferroni correction, confirming that this independence holds consistently across all frequency bands. The RV coefficient from CoIA likewise failed to reach significance (RV = `r sdg$rv`, p = `r sdg$rv_p`), indicating that no systematic covariation exists between the two dimensions at the level of overall multivariate structure either.

The CoIA dimensionality reduction results provide further detail on the covariation between morphology and scar patterns. The first two CoIA axes together account for 90.6% of total covariation (Axis 1 = 57.6%, Axis 2 = 33.0%; @fig-coia-composite). Axis 1 primarily contrasts cores with flattened morphologies and unidirectional removals against those with more rounded forms and birectional removals. Axis 2 captures additional variation in both scar pattern structure and morphological detail. On the scar pattern side, it distinguishes unidirectional and bidirectional removals from more centripetal patterns. On the morphology side, it primarily reflects mid-frequency shape variation associated with three-fold and four-fold rotational symmetry, likely corresponding to differences in core cross-sectional geometry (@fig-coia-composite).

```{r}
#| label: fig-coia-composite
#| echo: false
#| fig-width: 12
#| fig-height: 18
#| fig-cap: "CoIA biplot, line length, and line direction by layer, raw material, and core type."

sdg$fig_coia_composite
```

To examine how line features derived from CoIA vary across grouping conditions, we analyzed line length and direction separately. Kruskal–Wallis tests showed no significant differences in line length among layers (χ² = `r sdg$kw_arrow_layer_chi2`, df = 1, p = `r sdg$kw_arrow_layer_p`), raw materials (χ² = `r sdg$kw_arrow_rawmat_chi2`, df = 1, p = `r sdg$kw_arrow_rawmat_p`), or core types (χ² = `r sdg$kw_arrow_type_chi2`, df = 5, p = `r sdg$kw_arrow_type_p`; @fig-coia-composite), indicating that the decoupling between morphology and scar patterns is consistent in magnitude across all grouping conditions. Rayleigh tests on line direction showed random distributions for most groups and core types, with the only exception of Unifacial unidirectional cores, which exhibited significantly concentrated directionality (U = `r sdg$rayleigh_uni_U`, p = `r sdg$rayleigh_uni_p`) with a mean direction of approximately `r sdg$mean_dir_uni`°, indicating that their techno–morphology relationship is specifically structured along Axis 1. More specifically, regardless of how individual specimens vary in morphology and scar pattern space, Unifacial unidirectional cores consistently shift toward the flattened end of the morphological spectrum and the unidirectional end of the scar pattern spectrum — a directional tendency that is stable across all specimens of this type.

To explore the factors driving variation in morphology and scar patterns, PERMANOVA tests were conducted for the three grouping variables. Core type had a significant effect on both morphological variation (R² = `r sdg$perm_morph_type_r2`, F = `r sdg$perm_morph_type_f`, p = `r sdg$perm_morph_type_p`) and scar pattern variation (R² = `r sdg$perm_scar_type_r2`, F = `r sdg$perm_scar_type_f`, p = `r sdg$perm_scar_type_p`), indicating systematic differences in both dimensions among core types. Raw material had a significant effect on morphology (R² = `r sdg$perm_morph_rawmat_r2`, F = `r sdg$perm_morph_rawmat_f`, p = `r sdg$perm_morph_rawmat_p`), though pairwise comparisons between chert and sandstone did not survive Holm correction (p = `r sdg$perm_rawmat_pairwise_p`). Raw material had no significant effect on scar direction variation (R² = `r sdg$perm_scar_rawmat_r2`, F = `r sdg$perm_scar_rawmat_f`, p = `r sdg$perm_scar_rawmat_p`), indicating that knappers applied consistent reduction strategies regardless of raw material — the physical properties of raw materials shaped core morphology outcomes but did not influence technological decision-making. Layer had no significant effect on either morphology (R² = `r sdg$perm_morph_layer_r2`, F = `r sdg$perm_morph_layer_f`, P = `r sdg$perm_morph_layer_p`) or scar pattern (R² = `r sdg$perm_scar_layer_r2`, F = `r sdg$perm_scar_layer_f`, p = `r sdg$perm_scar_layer_p`), suggesting continuity in reduction strategies across layers, with no clear evidence of technological change.

Taken together, the core assemblage from Sandinggai shows a systematic decoupling between core morphology and reduction strategy. This decoupling is not limited to any specific condition but appears as a general pattern across different time periods, raw materials, and core types. Morphological variation is primarily associated with raw material and core type, while scar direction variation shows no significant response to these factors. This suggests that reduction strategies may be relatively stable and largely independent of changes in raw material or core type — possibly reflecting a consistent technological tradition maintained by knappers throughout the core reduction process.

# Discussion

The present study extends the application of spherical harmonic methods from approximately spherical artifact categories, such as spheroids and polyhedral cores, to a broader range of core types exhibiting substantially greater morphological variability. Validation results demonstrate that the method performs effectively in both morphological reconstruction and scar-pattern reconstruction. Because spherical harmonic methods characterize morphology using complete surface information rather than discrete landmarks, the resulting coefficients can be used to reconstruct original three-dimensional forms while maintaining high morphological fidelity at relatively high harmonic orders. Compared with landmark-based approaches, this characteristic enables researchers to directly evaluate the representational performance of the method through visual inspection of reconstructed models. Reconstruction results indicate that the selected truncation order of 20 harmonics is sufficient to capture detailed aspects of core morphology. Furthermore, M-SPHARM successfully identified significant differences associated with both technological categories and raw material types within the SDG assemblage. Morphological reconstructions projected along the CoIA covariation axes further demonstrate that the shape variation captured by M-SPHARM possesses interpretable geometric meaning. Taken together, these results indicate that spherical harmonic methods can effectively characterize the morphology of diverse core types and that their applicability extends well beyond highly regular artifact forms such as spheroids.

With regard to scar-pattern organization, the validity of SP-SPHARM has been confirmed through visual evaluation of reconstructed ideal scar-pattern models. The reconstruction results of both experimental knapping cores and SDG specimens similarly exhibit clear and interpretable differences in scar-pattern organization. At the level of statistical inference, the PERMANOVA results obtained in the validation framework further demonstrate the ability of SP-SPHARM to discriminate between different scar-pattern categories. Although no statistically significant differences were detected between Discoidal and Bidirectional cores, nor between Discoidal and Multiplatform cores, these outcomes should be interpreted with caution. The former is likely attributable to limited statistical power resulting from small sample size; notably, LDA scatterplots still reveal a clear separation between these two groups in the linear discriminant space. The latter case may instead reflect the relatively high internal variability of the Multiplatform category, within which certain specimens exhibit scar patterns closely resembling those of Discoidal cores. However, this does not imply that these specimens should be reassigned to the Discoidal category. On the contrary, this pattern suggests that scar organization does not correspond in a simple one-to-one manner with existing typological classifications, but instead encodes additional technological dimensions. This observation further underscores the analytical value of treating scar patterns and morphology as separate yet complementary dimensions of core variability. Overall, the spherical harmonic framework provides a quantitatively robust representation of both morphological and scar-pattern variation across different core types. The resulting signals of variation are sufficiently structured and informative to support subsequent statistical analyses.

Compared with existing quantitative approaches to scar-pattern analysis—namely the SPI index proposed by Chris Clarkson (2006) and the fabric metrics developed by Lin et al. (2024)—SP-SPHARM demonstrates clear advantages, although it does not constitute a complete replacement for either method. Statistical testing of the SPI index shows that it is particularly effective in distinguishing unidirectional cores from other categories, consistent with the original conclusions of Clarkson (2006). In that study, the SPI was further transformed by applying an arccosine function to derive a scar-pattern angle, which was subsequently squared to enhance discrimination among core types. This transformation successfully separates conical and prismatic blade cores characterized by unidirectional parallel scar distributions, radial cores (including Levallois cores, discoidal cores, and Acheulean handaxes), and randomly reduced polyhedral cores. Compared with the raw SPI values, the squared scar-pattern angle improves discrimination of randomly reduced cores; however, even under this transformation, no effective separation is achieved among radial core subtypes (P = 0.907). From a conceptual perspective, the SPI index is essentially defined as the ratio between the resultant vector length and the sum of all scar vector lengths, making it conceptually analogous to the mean resultant length ($\bar{R}$) in directional statistics. The latter is computed by summing unit vectors and dividing by sample size to quantify the degree of directional concentration. The primary difference lies in weighting schemes: Clarkson’s SPI assigns weights based on actual scar length, meaning that longer scars contribute more strongly to the resultant vector, whereas $\bar{R}$ assigns equal weight to each scar, and is mathematically equivalent to the unit-vector SPI formulation used by Bretzke (2012). Although these weighting schemes may yield different sensitivities for specific core types, both approaches fundamentally aim to quantify the concentration and dispersion of directional data. This also explains why SPI is most sensitive to unidirectional patterns, while performing poorly in discriminating among radial core subtypes and polyhedral cores.

The fabric-based method of Lin et al. (2024) draws on taphonomic fabric analysis and treats each core as analogous to an archaeological assemblage, with each scar representing an analytical unit and the best-fit core plane serving as a reference “horizon.” This framework assesses orientational variability relative to this plane. By calculating tensor-derived metrics such as elongation and isotropy, the method characterizes scar organization along two orthogonal dimensions: dispersion within the plane and variation perpendicular to it. In the present evaluation, the fabric method identifies additional significant differences between bidirectional cores and radial cores (Levallois and discoidal), compared with SPI. This improvement arises from its explicit incorporation of vertical orientational information. However, the method fails to distinguish between unidirectional and bidirectional cores, a limitation that is directly derived from its mathematical formulation. This constraint is also reflected in the standardized Euclidean distance structure of the ideal model space, where fabric metrics cannot differentiate between parallel scar systems with opposite orientations. This indicates that fabric analysis should not be considered a replacement for SPI, but rather a complementary approach capturing a different dimension of scar organization.

In contrast, SP-SPHARM achieves the highest number of statistically significant pairwise distinctions among all tested methods. Of particular importance is its ability to separate subtypes within radial cores, notably the significant distinction between Levallois and discoidal cores. Although sample size limitations and significant PERMDISP results indicate that within-group dispersion contributes to PERMANOVA outcomes, LDA visualization nonetheless reveals a clear separation between these two categories. This discriminative power is not achieved by existing methods and likely arises from the fact that SP-SPHARM directly encodes the full spatial structure of scar organization rather than compressing it into a small set of summary variables. As a result, it is capable of capturing fine-grained phase-dependent differences in scar configuration. However, SP-SPHARM does not imply a complete replacement of existing methods. Unlike SPI and fabric metrics, SP-SPHARM allows spherical reconstruction of scar-pattern surfaces, yet these reconstructions cannot be straightforwardly mapped onto discrete scar-pattern categories. Even when idealized templates are used as references, the variability observed in real archaeological specimens makes direct categorical interpretation of reconstructed forms highly uncertain. By contrast, SPI and fabric-based indices retain stronger direct physical interpretability. Accordingly, the strength of SP-SPHARM lies in its ability to capture the global structure of scar organization with higher statistical sensitivity, making it particularly suitable for large-scale comparative analyses, such as cross-site or regional comparisons of reduction strategies. For more fine-grained archaeological interpretation, however, SPI and fabric methods remain essential complements. In practice, an effective analytical workflow is to first assess overall group differences using SP-SPHARM, and then integrate SPI and fabric-based analyses, together with typological frameworks, to refine archaeological interpretation of the observed patterns.

Building on the above validation, the present study further integrates spherical harmonic power spectra of both core morphology and scar-pattern organization into a CoIA framework to test their covariation at the assemblage level. Results from the experimental knapping cores indicate that neither the global Mantel statistic nor the RV coefficient reaches statistical significance, suggesting that no systematic covariation exists between core morphology and scar-pattern organization at the level of the entire assemblage. However, global tests may obscure structured patterns within subgroups. A more fine-grained interpretation based on the CoIA biplot reveals a more nuanced picture. Although no significant differences were detected in line lengths across all core types—indicating comparable degrees of technology–morphology decoupling among categories—the orientation of the connecting lines shows a significant non-random clustering pattern for Levallois, discoidal, and bidirectional cores. This suggests that, even though these core types do not exhibit stronger overall correspondence between morphology and scar organization than other categories, their covariation tends to unfold along internally consistent directional axes. In other words, when morphological variation occurs within these core types, scar-pattern organization tends to adjust in a predictable and structured manner along specific covariation trajectories. In the case of Levallois cores, this directional consistency may reflect intrinsic constraints linking morphology and flaking organization within the technological system. This interpretation is consistent with Éric Boëda’s concept of volumetric conception, which proposes that Levallois reduction sequences integrate volume management and flaking strategy within a unified operational logic. It should be noted, however, that the relatively small sample sizes for discoidal and bidirectional cores warrant caution in interpreting the statistical significance of these results.

Applying the above analytical framework to the SDG assemblage and comparing the results with previous studies provides a more direct evaluation of the method’s performance on real archaeological materials. Earlier work, following the core typological framework of Manuel de la Torre, proposed that lithic technology at the SDG site remained broadly continuous through time, with no major technological shifts. Two principal operational sequences were identified: (1) a small-flake reduction system primarily based on chert, and (2) a large-flake and large cutting tool (LCT) production system based on quartz sandstone. However, previous studies have not fully explored whether these two raw material systems involve distinct reduction strategies at a finer behavioral level. The results of the present spherical harmonic framework show no significant differences in either core morphology or scar-pattern organization between Layer 4 and Layer 3 at SDG, providing quantitative support for previous observational interpretations and suggesting continuity in knapping strategies across stratigraphic layers, at least at the level of reduction behavior. It should be noted that Layer 2 was excluded from statistical analysis due to insufficient sample size. At the same time, PERMANOVA results reveal a more nuanced pattern of raw material structuring in technological behavior. Cores made of chert and quartz sandstone differ significantly in morphology, but show no significant differences in scar-pattern organization. This indicates that knappers employed a consistent reduction strategy that did not systematically vary according to raw material type or size constraints. The observed morphological differences therefore likely reflect intrinsic differences in raw material properties: under a relatively uniform knapping strategy, different materials yield distinct core morphologies, without implying a corresponding change in technological organization. Importantly, despite the overall decoupling between technology and morphology, Co-inertia Analysis still reveals a significant concentration in the covariation direction for unifacial unidirectional cores. This suggests that this core type—often interpreted as a “expedient” or opportunistic reduction strategy—may, within the SDG assemblage, have been subject to specific structural constraints. The precise nature of these constraints, however, remains unclear and requires further investigation using additional analytical dimensions. Nevertheless, the present spherical harmonic framework demonstrates clear potential for identifying structured regularities that are difficult to detect using traditional analytical approaches.

At a more macro-analytical level, the methodological framework proposed in this study advances the development of three-dimensional quantitative approaches to core analysis and offers potential analytical pathways for addressing several unresolved issues in Paleolithic archaeology. Taking the debate on Levallois technology in China as an example, discussions surrounding the Guanyindong Site have included statistical analyses of cores and detached products based primarily on linear metric attributes. However, systematic three-dimensional quantitative evaluations remain lacking. Similarly, the evolutionary relationship between Levallois technology and Acheulean industries has long been a central topic in Lower–Middle Paleolithic research. Competing hypotheses regarding the origin of the Levallois concept include independent invention, derivation from Acheulean handaxe traditions, and development from earlier prepared-core technologies of the Lower Paleolithic. Recent studies employing landmark-based geometric morphometrics have compared Levallois cores, prepared cores, and Acheulean bifaces in a three-dimensional framework, suggesting that multiple regional trajectories of origin may exist. However, as discussed above, landmark configurations often vary between researchers, making cross-study integration difficult and ultimately limiting the potential for large-scale comparative analyses. The approach proposed in the present study offers a potential solution to this limitation. Although the current application focuses exclusively on cores and does not yet include large cutting tools (LCTs), the method is, in principle, fully applicable to such artifact classes. More importantly, it also has the potential to be integrated into emerging analytical paradigms. Recent work has incorporated elliptic Fourier series into phylogenetic analyses, although these applications have been restricted to two-dimensional morphologies of flaking products. As a three-dimensional generalization of Fourier methods, spherical harmonic analysis could be readily extended to phylogenetic analytical frameworks. In particular, the scar-pattern component developed in this study provides a direct line of evidence for reconstructing hominin flaking strategies. Future integration of this approach with phylogenetic modeling may therefore offer new opportunities to investigate technological transmission, population interaction, and processes of social learning among prehistoric human groups.

This study also has several limitations. As noted above, constraints in sample size mean that the current results should be interpreted primarily as exploratory, and they cannot yet be regarded as definitive empirical conclusions. In addition, at the methodological level, the present study focuses on core-level analyses, and therefore the patterns identified are relatively macroscopic in nature. The regularities revealed here should be further examined in future work through the integration of more fine-grained and targeted technological attributes. Nevertheless, the present study introduces an innovative analytical approach and a systematic research framework, which together provide an important foundation for future extensions and more detailed investigations in related research domains.

# Conclusion

# CRediT authorship contribution statement

# Data availability statement

# Funding

# Declaration of competing interest

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```
