---
title: "Characterizing core scar patterning and morphology within a unified spherical harmonic framework"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
  - Hao Li:
      correspondence: "yes"
      orcid: 
      institute:
        - ALPHA
        - UCAS
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - UCAS: University of Chinese Academy of Sciences, Beijing 100049, China
  - ALPHA: Group of Alpine Paleoecology and Human Adaptation (ALPHA), Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing 100101; 
  - TPESRE: State Key Laboratory of Tibetan Plateau Earth System, Resources and Environment (TPESRE), Beijing 100101
  - uw: Department of Anthropology, University of Washington, Seattle, WA, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
number-sections: true
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  
editor_options: 
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

```{r}
#| label: setup
#| include: false
library(targets)
library(tidyverse)
tar_config_set(store = "_targets")

sa          <- tar_read(spharm_analysis)
ph          <- sa$spharm_posthoc

im          <- tar_read(im_comparison)
dist_spi    <- im$dist_spi
dist_fabric <- im$dist_fabric
dist_spharm <- im$dist_spharm

cia         <- tar_read(exp_cia_analysis)
sdg         <- tar_read(sdg_cia_analysis)
```

# Introduction

Cores preserve the most direct and information-rich evidence of reduction strategies within lithic assemblages and therefore constitute critical evidence for understanding knapping behavior. Through core analysis, researchers can not only investigate hominin technological strategies, knapping skill, and the technological traditions represented within lithic assemblages, but also address broader archaeological questions concerning cognitive capacities, mobility patterns, site function, and mechanisms of social learning. @bakkerAppliedMultivariateStatistics @abrunhosa2019

Traditionally, core analysis has been grounded primarily in typological approaches, whereby individual cores are assigned to specific categories, sometimes in conjunction with the technological characteristics of detached products, in order to reconstruct reduction strategies. Different core types are generally defined by hierarchically organized sets of discrete attributes, including overall morphology, scar patterning, flaking angles, and the characteristics of flaking surfaces and platforms, thereby giving rise to categories such as Levallois, discoidal, and laminar cores. However, such typological frameworks are often far from standardized, and different types may rely on entirely different descriptive vocabularies. Even where terminology and technological definitions are relatively detailed, substantial uncertainty remains during practical observation, resulting in classification inconsistencies both between analysts and within the same analyst over time. The long-standing debate surrounding Levallois technology in China represents a typical example. Although researchers commonly identify Levallois cores according to the detailed criteria proposed by Éric Boëda, disagreements persist regarding whether certain cores from the Shiyu Site and Guanyindong Site should be classified as Levallois. To reduce such inconsistencies, several scholars have attempted to establish unified classificatory frameworks, including the “Unified Lithic Taxonomy” proposed by Nicholas Conard and the definitions of different volumetric structures of exploitation (VSE) within the Logic Analytical System. By employing shared terminology and standardized observational criteria, these approaches have partially reduced inter-observer variability in classification practice.

Nevertheless, regardless of the terminology or classificatory framework employed, most studies remain predominantly qualitative in both observation and analysis. In recent years, increasing demands for precision and reproducibility in lithic research have stimulated growing interest in quantitative approaches. Among these, 3D modeling technologies provide substantially higher-resolution data acquisition than traditional linear measurements and qualitative observations, and have already been widely applied to a range of lithic research questions and artifact categories. However, quantitative 3D studies specifically focused on cores remain comparatively limited. Existing 3D core analyses can generally be divided into two major approaches. The first involves extracting customized technological variables from 3D models, such as platform angles, cortex proportions, reduction intensity, volumetric ratios, and scar patterning; these variables are typically predefined by the researcher to capture specific technological dimensions of core reduction. The second approach focuses on the characterization and analysis of overall core morphology, most commonly through landmark-based geometric morphometrics, although recent studies have also introduced spherical harmonic methods into the study of stone balls and polyhedral cores. Importantly, these two approaches are not mutually exclusive but rather highly complementary, and several studies have achieved deeper insights through their integration.

Despite substantially expanding the methodological toolkit available for core analysis and overcoming some limitations of traditional approaches, these methods themselves still present several shortcomings. As noted by Antoine Muller, landmark-based approaches necessarily compress continuous morphological variation into a limited number of discrete points, inevitably resulting in information loss. Furthermore, landmark methods rely on assumptions of homology. Current landmark configurations for cores, which typically place landmarks along core edges and semi-landmarks across core surfaces, are mainly applicable to highly standardized technological systems such as Levallois reduction. In contrast, they are difficult to apply to more variable forms, such as polyhedral cores, or to comparative analyses across different core types. Even within Levallois core studies, landmark configurations vary considerably between researchers, making direct comparisons between morphospaces problematic. By contrast, spherical harmonic approaches do not depend on homology assumptions, are capable of capturing morphological information more comprehensively, and possess substantial potential for cross-type comparisons. In addition, power spectra derived from spherical harmonic coefficients are rotationally invariant, thereby eliminating the need for manual orientation of cores prior to analysis. To date, however, applications of this method have remained largely restricted to artifact classes with globally near-spherical morphologies, such as stone balls and polyhedral cores, while its applicability to more morphologically diverse core types remains insufficiently explored.

At the same time, quantitative approaches to scar pattern analysis remain relatively underdeveloped. To date, only Chris Clarkson and Lin have proposed relatively systematic quantitative frameworks. The former introduced the Scar Pattern Index (SPI), a univariate metric that measures directional concentration through the ratio between resultant vector length and total vector length. The latter adapted the fabric analysis framework from taphonomy, using eigenvalues of orientation tensors to characterize the dispersion of scar orientations across two orthogonal dimensions. Both approaches represent important advances in the quantification of scar organization. Fundamentally, however, they compress complex scar-vector configurations into one or only a few parameters, addressing primarily the question of “how dispersed” scar distributions are, rather than directly characterizing the structural properties of scar patterning itself. Such dimensional reduction inevitably entails information loss, and this limitation is reflected in their analytical performance. Both SPI and fabric-based approaches are effective at distinguishing broad technological organizations, such as cores characterized by unidirectional or bidirectional removals from single or opposed platforms, as well as Levallois and discoidal systems involving intersecting flaking surfaces. However, their ability to differentiate variability within these broader categories remains comparatively limited. These constraints indicate that substantial room for methodological development still exists.

In addition, the relationship between scar organization and core morphology has long constituted a central problem in core analysis. Previous studies have demonstrated that knappers actively structured flaking sequences to produce and maintain particular morphologies throughout reduction. At the same time, core morphology may also be strongly constrained by objective factors such as raw material properties and reduction intensity, rather than being determined solely by technological intention. Consequently, there remains a lack of quantitative methods capable of simultaneously characterizing technological organization and core morphology within a unified analytical framework while directly testing their covariation. This limitation has hindered efforts to evaluate the relative contributions of “technological intention” and “objective constraints” within core reduction sequences.

Against this background, the present study systematically explores the application of spherical harmonic methods to core analysis and develops an integrated quantitative framework linking technology and morphology. For morphological analysis, we evaluate the applicability of spherical harmonics across diverse core types rather than restricting the method to globally near-spherical forms. For scar pattern analysis, we likewise employ spherical harmonics to characterize overall scar organization on core surfaces. Because scar data on individual cores are typically sparse, we first transform discrete scar vectors into continuous spatial distributions using kernel density estimation in order to preserve their original spatial structure as fully as possible. These continuous distributions are then subjected to spherical harmonic expansion, from which descriptors including spherical harmonic coefficients and power spectra are derived. Through this workflow, technological and morphological information can be integrated within the same mathematical framework, thereby reducing subjective dependence on normalization procedures and weighting schemes when combining datasets with different dimensional properties. Building upon this framework, we further employ Co-inertia Analysis (CoIA) to project the dimensionality-reduced results of different datasets into a shared covariance space, allowing the technological–morphological covariation patterns of individual cores to be examined within a unified feature space. Combined with the reconstructive capacity of spherical harmonic functions, this approach additionally enables direct reconstruction of technological and morphological trajectories along CoIA covariation axes, thereby reconnecting statistical patterns to concrete archaeological interpretations. In this way, variation among cores is no longer treated merely as a static arrangement of typological categories, but can instead be understood and discussed as part of a continuous spectrum of technological and morphological variability.

# Material and Methods

**2.1 Materials**

The materials analyzed in this study derive from three datasets. The Ideal Digital Cores dataset was used to evaluate the sensitivity of spherical harmonic methods to standardized scar patterns and to visually demonstrate how different scar organizations are represented through spherical harmonic power spectra and coefficient-based reconstructions. The Experimental Knapping Cores dataset was primarily employed to test the scar-pattern analysis framework under conditions of greater morphological and technological variability, while also integrating morphological information to evaluate the statistical framework proposed in this study. Finally, the Archaeological Specimens dataset was analyzed within the validated analytical framework in order to demonstrate the practical archaeological significance of the spherical harmonic approach.

Ideal digital cores: One idealized digital model was included for each core type, all sourced from Lin. These standardized digital core models comprise eight idealized core categories, including cylindrical, conical, discoidal, Levallois (preferential, convergent, and laminar), bifacial, and polyhedral cores. Each category is associated with distinct scar organizations. For example, cylindrical cores include unipolar and bipolar scar configurations, whereas discoidal cores include both bifacial and unifacial scar patterns. Detailed descriptions of all scar-pattern categories are provided in the supplementary materials.

Experimental knapping cores: The experimental cores were produced by Chris Clarkson and comprise a total of 58 specimens, including 16 unidirectional cores, 6 bidirectional cores, 21 Levallois cores (10 preferential, 5 convergent, 2 Nubian, 2 recurrent, and 2 laminar), 5 discoidal cores, and 10 multiplatform cores. These specimens are currently curated at the School of Social Sciences at the University of Queensland. The cores were digitized by Sam Lin using a Polyga HDI Compact C210 structured-light scanner in combination with FlexiScan3D software. Post-processing of the scanned models was conducted in Geomagic Wrap, including the filling of minor holes and the correction of subtle mesh errors such as non-manifold edges and self-intersections.

Archaeological specimens: The archaeological specimens derive from the Sandinggai Site (hereafter SDG) in Hunan Province, China, previously reported by Li. The SDG sequence is divided into three cultural layers (L2, L3, and L4), with optically stimulated luminescence (OSL) dating indicating an age range of approximately 100–20 ka. According to Li, the lithic assemblage at SDG is characterized by two distinct operational sequences: (1) a small-flake system primarily based on chert exploitation, and (2) a large-flake and large cutting tool (LCT) system primarily based on quartz sandstone.

The present study focuses on the previously reported cores from the SDG assemblage. After excluding incomplete specimens, heavily abraded cores, and tested cores, 51 complete and well-preserved cores retaining clear technological information were selected for analysis. These specimens derive from Layers 2, 3, and 4; however, because only two cores originate from Layer 2, the statistical analyses presented here are restricted to Layers 3 and 4. Core classifications follow the criteria established in Li, and detailed operational procedures are described in the original publication. Information concerning core types and raw materials is provided in Table X.

All archaeological cores were digitized using a EinScan Pro XS handheld 3D scanner manufactured by SHINING 3D, in combination with a turntable and fixed camera stand to ensure stable scanning conditions. Model alignment and export were completed using the accompanying EinScan-Pro series_v3.1.0.4 software, while subsequent mesh processing was conducted in MeshLab and Geomagic Wrap. Processing procedures included the filling of minor holes and the correction of subtle mesh defects such as non-manifold edges and self-intersections.

**2.2 Methods**

**2.2.1 Data Extraction**

The protocol used for recording scar orientations follows the procedures established by Lin. Only scars larger than 5 mm were included in the analysis. Scar orientations were recorded as vectors defined by the three-dimensional coordinates of their start and end points, using Geomagic Wrap 2021 for data acquisition and export. In addition, following the procedures of Lin, a best-fit plane was calculated for each core model within Geomagic Wrap 2021, and the coordinates of both the plane centroid and its normal vector were exported for subsequent model alignment.

Two alignment procedures were employed in this study in order to eliminate initial phase differences among core models. The first follows the method proposed by Lin and aligns models based on their best-fit planes. The workflow consists of: (1) rotating the normal vector of the best-fit plane to coincide with the Z-axis of the three-dimensional coordinate system; and (2) identifying the longest scar and translating its starting point to the global origin. These procedures were implemented using custom R scripts, and interactive HTML outputs were generated to allow direct visual inspection of the alignment results for each model.

The second alignment procedure is proposed for the first time in the present study and relies exclusively on the spatial organization of scar data. The workflow consists of: (1) calculating unit vectors for all scars and applying singular value decomposition (SVD), with the third singular vector selected as the normal vector of the alignment plane, as it represents the direction least correlated with the overall scar orientation structure; (2) rotating this normal vector to align with the Z-axis; (3) calculating the global centroid of all scar coordinates and translating this centroid to the global origin; and (4) using the first singular vector derived from the SVD as the principal direction of scar organization and aligning it with the X-axis.

The original vectors, together with the vectors aligned using both procedures, were subsequently used to evaluate the rotational invariance of different quantitative indices of scar patterning.

**2.2.2 Spherical Harmonic Analysis**

Spherical harmonic analysis (SPHARM or SHA) represents the three-dimensional extension of elliptic Fourier analysis. This method characterizes the surfaces of closed genus-zero objects using a relatively small set of shape descriptors that are invariant to rotation, translation, and scaling. In addition, spherical harmonic methods allow the original three-dimensional morphology of an object to be reconstructed directly from these descriptors. The implementation of SPHARM is based on the surface parameterization framework proposed by Brechbühler, in which each surface vertex is uniquely mapped onto the unit sphere $S^2$ through the equivalent coordinate functions $x(\theta, \phi)$, $y(\theta, \phi)$ 和 $z(\theta, \phi)$.

Initially, spherical harmonic methods were developed in the field of medical imaging for the morphometric analysis of brain structures. With subsequent methodological advances, the approach evolved from applications based on voxel representations to analyses of triangular mesh models and has since been widely adopted across multiple disciplines for three-dimensional morphometric research, including medical imaging, biological morphometrics, and particle simulation and reconstruction in engineering mechanics. More recently, several studies have applied spherical harmonic methods to lithic analysis. Sholts used spherical harmonics to quantify three-dimensional asymmetry in Clovis points, while Antoine Muller and Ye applied the method to investigate the morphology of spheroids. Collectively, these studies demonstrate that spherical harmonic methods constitute a mature and practically valuable framework for three-dimensional morphometric analysis.

However, the value of spherical harmonic methods extends beyond morphology alone. Previous studies have shown that morphological data can be analyzed using directional statistical approaches and that mathematical relationships exist between morphological and directional datasets. More recently, spherical harmonic methods have been employed to characterize nodal force orientations within spatial truss structures in architectural design, further demonstrating their suitability for modeling directional information. In the present study, we apply spherical harmonic analysis to scar orientation data for the first time, performing spherical harmonic analysis on flaking scar organization.

Our analytical workflow is as follows. First, scar vectors are transformed using von Mises–Fisher spherical kernel density estimation (VMF-sKDE), converting discrete scar orientations into a continuous density distribution on the sphere:

$$f(\mathbf{g}) = \frac{1}{n} \sum_{i=1}^{n} \exp\left(\kappa \, \mathbf{g} \cdot \mathbf{x}_i\right)$$

where $\mathbf{g}$ is a unit vector on the spherical grid, $\mathbf{x}_i$ is the unit vector representing the orientation of the i-th scar, and $\mathbf{κ}$ is the concentration parameter of the von Mises–Fisher kernel, controlled by the bandwidth parameter $\mathbf{h}$. The relationship between $\mathbf{κ}$ and $\mathbf{h}$ is given by:

$$\kappa = \frac{1}{h^2}$$

The bandwidth is inversely related to $\mathbf{κ}$: smaller bandwidth values produce larger $\mathbf{κ}$ values and consequently sharper density distributions, whereas larger bandwidths generate smoother distributions. In the present study, a bandwidth value of 0.35 was selected, corresponding to $\mathbf{κ}$ ≈ 8.16, which provides a balanced compromise between oversmoothing and overfitting. If $\mathbf{κ}$ is excessively large, the kernel density approaches a Dirac delta function, requiring very high truncation orders in the spherical harmonic expansion to achieve sufficient approximation. Conversely, if $\mathbf{κ}$ is too small, density contributions from neighboring orientations overlap excessively, reducing directional resolution and obscuring differences between scar-pattern distributions. The suitability of this parameter choice was validated through reconstruction of the original distributions, in which distinct scar-pattern organizations remained directly observable within the reconstructed spherical harmonic morphologies, indicating that the selected bandwidth adequately captures directional variation.

The KDE distributions were subsequently interpolated onto a 64×128 Driscoll–Healy grid to satisfy the spherical rasterization requirements of pyshtools. Spherical harmonic expansion was then performed on the resulting spherical coordinates. Each shape $f(\theta, \phi)$ was first mapped onto the unit sphere and subsequently represented as a weighted combination of spherical harmonic basis functions $Y_l^m(\theta, \phi)$:

$$f(\theta, \phi) = \sum_{l=0}^{\infty} \sum_{m=-l}^{l} \hat{f}(l, m) \, Y_l^m(\theta, \phi)$$

where the coefficients $\hat{f}(l, m)$ quantify the contribution of each basis function to the overall shape. Here, θ and ϕ denote spherical coordinate angles, while l and m represent the order and degree, respectively. In this study, preliminary parameter screening resulted in the selection of a truncation threshold of $l_{\max} = 20$ in order to balance computational efficiency and descriptive accuracy. Previous studies have demonstrated that $l_{\max} = 20$ is sufficient to capture detailed lithic morphological variation. Under this truncation order, spherical harmonic coefficients were calculated. These coefficients retain phase information and can therefore be used to reconstruct original morphologies. However, phase information may introduce noise into comparative statistical analyses of shape variation. Accordingly, spherical harmonic power spectra were calculated from the coefficients in order to represent phase-independent spherical harmonic morphology. The resulting power spectrum consists of the relative contributions of different harmonic orders, with the total normalized to 1.

For core morphological analysis, the workflow follows the procedures established by Ye. Three-dimensional models (.stl) were first downsampled to 20,000 faces, after which Laplacian smoothing was applied to reduce local surface noise and regularize vertex distribution. Centroid normalization and principal-axis alignment were then performed to ensure that all vertex coordinates were positioned within a common Cartesian coordinate system, which was subsequently transformed into spherical coordinates. Finally, spherical harmonic expansion was conducted, and both spherical harmonic coefficients and power spectra were calculated under the condition of $l_{\max} = 20$. The complete analytical workflow is illustrated in Figure X。

**2.2.3 Data Analysis**

[Rotation Invariance Assessment.]{.underline} To evaluate whether different scar pattern metrics are rotation-invariant, we employed the Bland–Altman approach. The Bland–Altman analysis is widely used to assess agreement between two sets of measurements and provides a quantitative evaluation through the Bland–Altman plot. In this plot, the x-axis represents the mean of the two measurements, while the y-axis represents their difference. The mean difference is defined as the bias, and the 95% limits of agreement (LoA) are calculated based on this bias. A high level of agreement is inferred when the limits of agreement are narrow, approximately 95% of the data points fall within these limits, and the differences are randomly distributed without systematic trends.

In this study, the Bland–Altman framework was used to test the rotation invariance of three metrics: SPI, fabric analysis (Elongation and Isotropy), and the SP-SPHARM power spectrum. Specifically, each metric was computed under three conditions: (1) using the original scar-vector data, (2) using morphological plane aligned data, and (3) using technological plane aligned data. Pairwise Bland–Altman comparisons were then conducted among these conditions. High agreement between the paired measurements was interpreted as evidence of rotation invariance for the corresponding metric.

[Method comparison.]{.underline} To evaluate the effectiveness of different scar-pattern analysis methods, both the ideal digital core dataset and the experimental knapping core dataset were used for comparative testing. Because the current spherical harmonic power spectrum includes 20 harmonic orders, further dimensional screening was required prior to statistical analysis in order to reduce noise and avoid the curse of dimensionality in subsequent models. Harmonic orders were therefore evaluated based on a combination of their coefficients of variation and cumulative power contributions.

Because only one idealized model exists for each core type, analytical performance was assessed using pairwise Euclidean distances, specifically by measuring the degree of separation between different ideal core types within a standardized feature space. Greater pairwise distances indicate stronger differentiation between core types and therefore provide a measure of the discriminatory power of each analytical method. Results were visualized using heatmaps, and mean pairwise distances were calculated for each method.

For the experimental knapping cores, a significance-testing framework was employed to evaluate the practical performance of different quantitative scar-pattern metrics. Methods capable of identifying the largest number of statistically distinguishable core-type pairs were considered to possess stronger discriminatory performance. Because the Scar Pattern Index (SPI) is a univariate metric, differences between groups were evaluated using the Kruskal–Wallis test. In contrast, Fabric metrics and SP-SPHARM descriptors were analyzed using PERMANOVA and PERMDISP. Within this framework, PERMANOVA evaluates overall structural differences between groups, whereas PERMDISP tests differences in within-group dispersion. Significant PERMDISP results indicate that differences in dispersion may partially contribute to the observed multivariate separation, requiring PERMANOVA results to be interpreted with caution. To ensure statistical robustness, all permutation procedures were conducted using 9,999 permutations.

Where overall tests yielded significant results, post hoc comparisons were subsequently performed. P-values from post hoc analyses were corrected using the Holm–Bonferroni procedure in order to balance the risks of Type I and Type II errors.

[Techno-morphological statistical analytical framework.]{.underline} Based on the spherical harmonic power spectra derived from both core morphology and scar-pattern organization, we employed the following analytical framework to investigate technology–morphology relationships. First, we tested whether an overall covariation relationship exists between morphology and scar organization. Second, through joint dimensional reduction, we further explored the interaction patterns between morphological and technological variation.

*Isometric log-ratio transformation.* Because spherical harmonic power spectra possess the properties of compositional data, in which the total power is constrained to a fixed sum, variation in one harmonic order necessarily induces covariance among the remaining orders. This generates spurious correlations within the data space and violates assumptions of variable independence. To address this issue, the present study applied Isometric log-ratio transformation (ILR) to all power spectrum data. Based on compositional data analysis (CoDA) theory, ILR transformation projects compositional data into Euclidean space through log-ratio calculations among harmonic orders, thereby eliminating the statistical bias introduced by closure effects. All subsequent statistical analyses based on spherical harmonic power spectra were therefore conducted using ILR-transformed data.

*Mantel test and RV coefficient.* Overall covariation relationships were evaluated using both Mantel tests and permutation tests based on the RV coefficient. The Mantel test assesses the correlation between two distance matrices by calculating the correlation coefficient between corresponding matrix elements and evaluating significance through random permutation procedures. This approach is particularly suitable for testing structural correspondence between different variable systems. In the present study, Spearman’s rank correlation coefficient was used, with significance assessed using 9,999 permutations.

The RV coefficient is a multivariate statistical measure used to quantify the overall degree of association between two datasets. RV values range from 0 to 1, with values approaching 1 indicating greater similarity in overall dataset structure. In this study, the significance of RV coefficients was likewise evaluated using permutation tests with 9,999 permutations. Because the Mantel test primarily emphasizes correlations between distance matrices, whereas the RV coefficient focuses more directly on covariance in multivariate structure, the combined use of both methods allows dataset correspondence to be evaluated from complementary statistical perspectives, thereby improving the robustness and interpretive reliability of the results.

*Co-inertia analysis.* Co-inertia Analysis (CoIA) is a multivariate statistical method designed to evaluate structural relationships between two datasets. Its core principle is to identify shared structures in which the directions of variation within the two datasets are maximally concordant while simultaneously maximizing covariance between them. Typically, principal component analysis (PCA) is first performed separately on each dataset, after which projections are used to identify correspondences both among samples and among variables across the two data spaces. Compared with simple correlation analyses, CoIA not only determines whether two datasets are related, but also reveals which samples or variables primarily drive these relationships. Consequently, the method is particularly well suited for comparing datasets that differ in physical meaning but share identical sampling units.

In CoIA ordination plots, the axes represent the principal covariation axes extracted by the analysis and therefore correspond to the major directions of shared variation between the two datasets. Each sample is typically represented by two points corresponding to its projected position within each dataset, connected by a line segment. The length of the connecting line reflects the degree of discrepancy between the positions of the same sample in the two data spaces: shorter lines indicate stronger correspondence and greater structural consistency between datasets, whereas longer lines indicate larger discrepancies in the representation of that sample. The direction of the connecting line reflects the corresponding trend of variation between the two datasets within the covariation space. If line directions are broadly consistent across samples, this indicates a stable and coherent covariation structure between the datasets; conversely, highly dispersed directions suggest weaker structural correspondence or the presence of localized variation.

To further evaluate the statistical properties of sample correspondences in CoIA space, the present study employed Kruskal–Wallis tests to compare line lengths among groups, thereby assessing whether structural consistency between datasets differs significantly across sample groups. In addition, Rayleigh tests were applied to line directions in order to evaluate whether line orientations exhibit significant directional clustering. The former identifies group-level differences in the degree of divergence between datasets, whereas the latter assesses whether line directions deviate significantly from random orientation patterns. Importantly, because line directions in CoIA fundamentally represent axial correspondences between two data spaces rather than unidirectional vectors with fixed origins and endpoints, they were treated here as axial data, such that directions differing by 180° were considered equivalent. Under this framework, Rayleigh tests provide an effective means of determining whether line orientations exhibit significant non-random clustering. The combined use of line-length and line-direction analyses therefore enables simultaneous evaluation of coupling strength and structural consistency between datasets from both distance-based and directional perspectives, thereby improving the completeness and reliability of CoIA interpretation.

**2.2.4 Reproducibility and open source materials**

To facilitate the re-use of our materials and to enhance the reproducibility and transparency of this study, we provide the complete R and Python scripts used for all analyses and visualizations presented in this paper in an online repository at XX. This version-controlled repository also includes the raw datasets underlying all reported analyses, together with an R package implementing the analytical workflows proposed in this study for application to other datasets. All figures, tables, and statistical results presented here can therefore be independently reproduced using the code and data provided in the repository.

# Results

**3.1 Validity Analysis**

**3.1.1 Rotational Invariance of Scar Pattern Analysis Measures**

```{r}
#| label: fig-rotation-validity
#| fig-cap: "Bland-Altman analysis of rotational invariance. **(A-C)** SPI across three coordinate conditions. **(D-F)** Fabric elongation ratio across three coordinate conditions. **(G-I)** Fabric isotropy ratio across three coordinate conditions. **(J)** SP-SPHARM power spectra bias across three coordinate conditions and random Z-axis perturbation."
#| fig-width: 12
#| fig-height: 14
#| echo: false
#| dpi: 150
tar_read(p_rotational_invariance_validity)
```

To evaluate rotational invariance, SPI, fabric metrics, and SP-SPHARM descriptors were computed in three coordinate systems: the original system, a system aligned to the morphological fitted plane, and one aligned to the technological fitted plane. Bland-Altman analysis was used to assess agreement across coordinates. SPI and fabric metrics showed near-perfect agreement across all three conditions, with biases and 95% limits of agreement (LoA) on the order of 10⁻¹⁶ — well within the range of floating-point numerical precision (@fig-rotation-validity) — confirming that both measures are strictly rotationally invariant. Consequently, coordinate alignment of the core model is not required prior to analysis.

The SP-SPHARM descriptors exhibited limited variation across alignment methods. Bland-Altman analysis revealed small systematic biases in low-order power (*l* = 1–2; 95% LoA = ±0.02), which diminished rapidly with increasing degree; no substantive differences were observed for *l* ≥ 3 (@fig-rotation-validity). To further evaluate this, we applied random rotational perturbations around the Z-axis to technological aligned data and recomputed the power spectra. Differences between perturbed and unperturbed spectra remained within machine precision, indicating that directional uncertainty within the coordinate frame — given a fixed Z-axis — has no effect on the power spectrum.

These results suggest that the SP-SPHARM descriptors is theoretically rotationally invariant by design, and that the systematic bias observed in low-order terms more likely originates from upstream computational steps such as KDE-based interpolation and coordinate mapping. Crucially, when a consistent alignment method is applied uniformly across all specimens, this systematic error affects all specimens equally and does not alter relative differences among them, leaving subsequent statistical comparisons unaffected. All subsequent analyses therefore employ the technological fitted plane for coordinate alignment.

**3.1.2 Validity of the SPHARM Method for Scar Pattern Analysis**

```{r}
#| label: fig-im-comparison
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| fig-cap: !expr >
#|   paste0("Pairwise standardised Euclidean distances between ideal core 
#|   models across three methods. Mean distances: SPI = ", dist_spi,
#|   ", Fabric = ", dist_fabric, ", SPHARM = ", dist_spharm, ".")

im$p_heatmap
```

**Test of Ideal models** Using ideal core models, we computed SPI, fabric metrics, and SP-SPHARM descriptors and assessed how well each method separates different patterns in feature space. Heatmaps revealed a progressive increase in inter-type separation from SPI to fabric metrics to SP-SPHARM descriptors (@fig-im-comparison). This ordering was corroborated by mean standardised Euclidean distance: the SP-SPHARM descriptors yielded the greatest inter-type distance ($\bar{d}_{\mathrm{E,std}}$ = `r dist_spharm`), substantially exceeding those of fabric metrics ($\bar{d}_{\mathrm{E,std}}$ = `r dist_fabric`) and SPI ($\bar{d}_{\mathrm{E, std}}$ = `r dist_spi`). These results indicate that SP-SPHARM descriptors produce the clearest separation among idealized scar patterns.\`

```{r}
#| label: fig-exp-method-compare
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 12
#| echo: false

tar_read(spharm_analysis)$p_exp_method_combined
```

**Test of experimental cores** To evaluate the sensitivity of each method on empirical data, we computed all three measures for experimental cores and conducted statistical comparisons by core type. All three methods detected significant inter-type differences (@fig-exp-method-compare). Post-hoc comparisons with Holm-Bonferroni correction are presented in @fig-exp-method-compare. The SP-SPHARM descriptors achieved the highest overall discrimination, identifying the greatest number of significant pairwise differences. Fabric metrics distinguished the majority of type pairs but failed to differentiate unidirectional from bidirectional cores. SPI was the most limited, detecting differences only where unidirectional cores were involved.

These results suggest that the three methods are complementary. SP-SPHARM descriptors capture the overall structure of scar patterning for most pairwise comparisons, though separation is reduced for patterns with bipolar symmetry, such as discoidal-bidirectional pairs ($P_{\text{adj}}$ = `r ph$p_dir_disc_bi`). Fabric metrics effectively distinguishes differences between unidirectional and bidirectional cores from Levallois and discoidal cores, but cannot differentiate unidirectional from bidirectional cores ($P_{\text{adj}}$ = `r ph$p_fab_uni_bi`), nor Levallois from discoidal cores ($P_{\text{adj}}$ = `r ph$p_fab_lev_disc`). SPI is most sensitive to the unidirectional type but shows limited discriminative capacity for other types. Notably, although SP-SPHARM descriptors failed to separete discoidal from bidirectional cores due to their similarity structures in SPHARM power spectrum (@fig-im-comparison), this distinction is successfully recovered by fabric metrics ($P_{\text{adj}}$ = `r ph$p_fab_disc_bi`). PERMDISP tests further revealed that discoidal and multiplatform cores exhibit greater within-group dispersion than other types, suggesting that these core types may be operationally more variable. It should also be noted that bidirectional (n = 6) and discoidal (n = 5) are represented by considerably fewer specimens than other types, which limits statistical power for comparisons involving these groups.

**3.2 Joint Analysis of Scar Patterns and Core Morphology in Experimental Knapping Specimens**

The analyses above demonstrate that SP-SPHARM descriptors exhibit broadly valid and significant differences across core types. To explore morphological variation among types, we further performed PERMANOVA tests using M-SPHARM descriptors. No significant differences in overall core morphology were detected across types ($R^2$ = `r sa$perm_morph_r2`, $P$ = `r sa$perm_morph_p`), suggesting that core morphology and reduction strategies may vary along largely decoupled axes at the assemblage level. To test this hypothesis, we integrated morphological and SP-SPHARM descriptors using Mantel tests and Co-Inertia Analysis (CoIA). The global Mantel test revealed no significant correlation between the two descriptor sets (Mantel's $r$ = `r cia$mantel_global_r`, $P$ = `r cia$mantel_global_p`), and the RV coefficient derived from CoIA likewise did not reach significance ($RV$ = `r cia$rv`, $P$ = `r cia$rv_p`).

The CoIA results provide further detail on the covariation between morphology and scar patterns. The first two CoIA axes together account for 88.8% of total covariation (Axis 1 = 70.3%, Axis 2 = 18.5%; @fig-exp-cia). Axis 1 primarily contrasts cores with elongated morphologies and longitudinally scars against those with flatter shape and more centripetal scar arrangements. Axis 2 captures residual variation in scar asymmetry and lateral organisation: on the scar pattern side, it distinguishes bidirectional removals from unidirectional flaking pattern; on the morphology side, it reflects differences in core thickness relative to width — that is, how plump or flat the core is in cross-section (@fig-exp-cia).

In the CoIA biplot, each specimen's morphological and scar pattern SPHARM spectra are mapped into the same feature space; the length of the line connecting the two projections reflects the degree of coupling between morphology and scar pattern, with shorter lines indicating closer correspondence. A Kruskal-Wallis test revealed no significant difference in line lengths across core types (@fig-exp-cia), suggesting that the overall degree of morphology–scar coupling is broadly comparable across reduction strategies. The orientation of lines reflects the direction of techno-morphological covariation within each type: concentrated orientations imply a consistent axis of joint variation, potentially indicative of an underlying operational schema. Rayleigh tests identified significant directional concentration in three core types — bidirectional (mean direction = `r cia$mean_dir_bi`°, $P$ = `r cia$rayleigh_bi_p`), discoidal (mean direction = `r cia$mean_dir_disc`°, $P$ = `r cia$rayleigh_disc_p`), and Levallois (mean direction = `r cia$mean_dir_lev`°, $P$ = `r cia$rayleigh_lev_p`) — while unidirectional and multiplatform cores showed no significant concentration (@fig-exp-cia). Watson two-sample tests further revealed that Levallois line orientations differ significantly from those of bidirectional, discoidal, and multiplatform cores (all $P$ \< 0.05), confirming that Levallois cores occupy a distinct techno-morphological trajectory in CoIA space. Given the interpretation of the CoIA axes, the Levallois mean direction of 299° projects primarily along the negative end of Axis 1, indicating that Levallois technology is consistently associated with elongated core morphology and longitudinally organised scar patterns. The convergence of bidirectional and discoidal mean directions (\~112–126°) along the positive region of Axis 2 suggests that these two types share a similar techno-morphological trajectory despite their typological differences.

```{r}
#| label: fig-exp-cia
#| echo: false
#| fig-width: 10
#| fig-height: 14
#| fig-cap: "CoIA biplot, line length, and directional distribution of techno-morphological coupling in experimental cores."

cia$p_final
```

**3.3 Archaeological Case Study: Core Analysis from Sandinggai**

Based on the analytical framework outlined in 3.2, we conducted a joint analysis of core morphology and scar patterns on 51 cores excavated from Sandinggai.

The global Mantel test revealed no significant pairwise distance correlation between the morphological spectrum and the scar pattern spectrum (r = `r sdg$mantel_global_r`, p = `r sdg$mantel_global_p`), indicating that core morphology and scar patterns are not significantly similar at the level of specimen pairs. Cross-frequency Mantel tests remained non-significant after Holm-Bonferroni correction, confirming that this independence holds consistently across all frequency bands. The RV coefficient from CoIA likewise failed to reach significance (RV = `r sdg$rv`, p = `r sdg$rv_p`), indicating that no systematic covariation exists between the two dimensions at the level of overall multivariate structure either.

The CoIA dimensionality reduction results provide further detail on the covariation between morphology and scar patterns. The first two CoIA axes together account for 90.6% of total covariation (Axis 1 = 57.6%, Axis 2 = 33.0%; @fig-coia-composite). Axis 1 primarily contrasts cores with flattened morphologies and unidirectional removals against those with more rounded forms and birectional removals. Axis 2 captures additional variation in both scar pattern structure and morphological detail. On the scar pattern side, it distinguishes unidirectional and bidirectional removals from more centripetal patterns. On the morphology side, it primarily reflects mid-frequency shape variation associated with three-fold and four-fold rotational symmetry, likely corresponding to differences in core cross-sectional geometry (@fig-coia-composite).

```{r}
#| label: fig-coia-composite
#| echo: false
#| fig-width: 12
#| fig-height: 18
#| fig-cap: "CoIA biplot, line length, and line direction by layer, raw material, and core type."

sdg$fig_coia_composite
```

To examine how line features derived from CoIA vary across grouping conditions, we analyzed line length and direction separately. Kruskal–Wallis tests showed no significant differences in line length among layers (χ² = `r sdg$kw_arrow_layer_chi2`, df = 1, p = `r sdg$kw_arrow_layer_p`), raw materials (χ² = `r sdg$kw_arrow_rawmat_chi2`, df = 1, p = `r sdg$kw_arrow_rawmat_p`), or core types (χ² = `r sdg$kw_arrow_type_chi2`, df = 5, p = `r sdg$kw_arrow_type_p`; @fig-coia-composite), indicating that the decoupling between morphology and scar patterns is consistent in magnitude across all grouping conditions. Rayleigh tests on line direction showed random distributions for most groups and core types, with the only exception of Unifacial unidirectional cores, which exhibited significantly concentrated directionality (U = `r sdg$rayleigh_uni_U`, p = `r sdg$rayleigh_uni_p`) with a mean direction of approximately `r sdg$mean_dir_uni`°, indicating that their techno–morphology relationship is specifically structured along Axis 1. More specifically, regardless of how individual specimens vary in morphology and scar pattern space, Unifacial unidirectional cores consistently shift toward the flattened end of the morphological spectrum and the unidirectional end of the scar pattern spectrum — a directional tendency that is stable across all specimens of this type.

To explore the factors driving variation in morphology and scar patterns, PERMANOVA tests were conducted for the three grouping variables. Core type had a significant effect on both morphological variation (R² = `r sdg$perm_morph_type_r2`, F = `r sdg$perm_morph_type_f`, p = `r sdg$perm_morph_type_p`) and scar pattern variation (R² = `r sdg$perm_scar_type_r2`, F = `r sdg$perm_scar_type_f`, p = `r sdg$perm_scar_type_p`), indicating systematic differences in both dimensions among core types. Raw material had a significant effect on morphology (R² = `r sdg$perm_morph_rawmat_r2`, F = `r sdg$perm_morph_rawmat_f`, p = `r sdg$perm_morph_rawmat_p`), though pairwise comparisons between chert and sandstone did not survive Holm correction (p = `r sdg$perm_rawmat_pairwise_p`). Raw material had no significant effect on scar direction variation (R² = `r sdg$perm_scar_rawmat_r2`, F = `r sdg$perm_scar_rawmat_f`, p = `r sdg$perm_scar_rawmat_p`), indicating that knappers applied consistent reduction strategies regardless of raw material — the physical properties of raw materials shaped core morphology outcomes but did not influence technological decision-making. Layer had no significant effect on either morphology (R² = `r sdg$perm_morph_layer_r2`, F = `r sdg$perm_morph_layer_f`, P = `r sdg$perm_morph_layer_p`) or scar pattern (R² = `r sdg$perm_scar_layer_r2`, F = `r sdg$perm_scar_layer_f`, p = `r sdg$perm_scar_layer_p`), suggesting continuity in reduction strategies across layers, with no clear evidence of technological change.

Taken together, the core assemblage from Sandinggai shows a systematic decoupling between core morphology and reduction strategy. This decoupling is not limited to any specific condition but appears as a general pattern across different time periods, raw materials, and core types. Morphological variation is primarily associated with raw material and core type, while scar direction variation shows no significant response to these factors. This suggests that reduction strategies may be relatively stable and largely independent of changes in raw material or core type — possibly reflecting a consistent technological tradition maintained by knappers throughout the core reduction process.

# Discussion

The present study extends the application of spherical harmonic methods from approximately spherical artifact categories, such as spheroids and polyhedral cores, to a broader range of core types exhibiting substantially greater morphological variability. Validation results demonstrate that the method performs effectively in both morphological reconstruction and scar-pattern reconstruction. Because spherical harmonic methods characterize morphology using complete surface information rather than discrete landmarks, the resulting coefficients can be used to reconstruct original three-dimensional forms while maintaining high morphological fidelity at relatively high harmonic orders. Compared with landmark-based approaches, this characteristic enables researchers to directly evaluate the representational performance of the method through visual inspection of reconstructed models. Reconstruction results indicate that the selected truncation order of 20 harmonics is sufficient to capture detailed aspects of core morphology. Furthermore, M-SPHARM successfully identified significant differences associated with both technological categories and raw material types within the SDG assemblage. Morphological reconstructions projected along the CoIA covariation axes further demonstrate that the shape variation captured by M-SPHARM possesses interpretable geometric meaning. Taken together, these results indicate that spherical harmonic methods can effectively characterize the morphology of diverse core types and that their applicability extends well beyond highly regular artifact forms such as spheroids.

With regard to scar-pattern organization, the validity of SP-SPHARM has been confirmed through visual evaluation of reconstructed ideal scar-pattern models. The reconstruction results of both experimental knapping cores and SDG specimens similarly exhibit clear and interpretable differences in scar-pattern organization. At the level of statistical inference, the PERMANOVA results obtained in the validation framework further demonstrate the ability of SP-SPHARM to discriminate between different scar-pattern categories. Although no statistically significant differences were detected between Discoidal and Bidirectional cores, nor between Discoidal and Multiplatform cores, these outcomes should be interpreted with caution. The former is likely attributable to limited statistical power resulting from small sample size; notably, LDA scatterplots still reveal a clear separation between these two groups in the linear discriminant space. The latter case may instead reflect the relatively high internal variability of the Multiplatform category, within which certain specimens exhibit scar patterns closely resembling those of Discoidal cores. However, this does not imply that these specimens should be reassigned to the Discoidal category. On the contrary, this pattern suggests that scar organization does not correspond in a simple one-to-one manner with existing typological classifications, but instead encodes additional technological dimensions. This observation further underscores the analytical value of treating scar patterns and morphology as separate yet complementary dimensions of core variability. Overall, the spherical harmonic framework provides a quantitatively robust representation of both morphological and scar-pattern variation across different core types. The resulting signals of variation are sufficiently structured and informative to support subsequent statistical analyses.

Compared with existing quantitative approaches to scar-pattern analysis—namely the SPI index proposed by Chris Clarkson (2006) and the fabric metrics developed by Lin et al. (2024)—SP-SPHARM demonstrates clear advantages, although it does not constitute a complete replacement for either method. Statistical testing of the SPI index shows that it is particularly effective in distinguishing unidirectional cores from other categories, consistent with the original conclusions of Clarkson (2006). In that study, the SPI was further transformed by applying an arccosine function to derive a scar-pattern angle, which was subsequently squared to enhance discrimination among core types. This transformation successfully separates conical and prismatic blade cores characterized by unidirectional parallel scar distributions, radial cores (including Levallois cores, discoidal cores, and Acheulean handaxes), and randomly reduced polyhedral cores. Compared with the raw SPI values, the squared scar-pattern angle improves discrimination of randomly reduced cores; however, even under this transformation, no effective separation is achieved among radial core subtypes (P = 0.907). From a conceptual perspective, the SPI index is essentially defined as the ratio between the resultant vector length and the sum of all scar vector lengths, making it conceptually analogous to the mean resultant length ($\bar{R}$) in directional statistics. The latter is computed by summing unit vectors and dividing by sample size to quantify the degree of directional concentration. The primary difference lies in weighting schemes: Clarkson’s SPI assigns weights based on actual scar length, meaning that longer scars contribute more strongly to the resultant vector, whereas $\bar{R}$ assigns equal weight to each scar, and is mathematically equivalent to the unit-vector SPI formulation used by Bretzke (2012). Although these weighting schemes may yield different sensitivities for specific core types, both approaches fundamentally aim to quantify the concentration and dispersion of directional data. This also explains why SPI is most sensitive to unidirectional patterns, while performing poorly in discriminating among radial core subtypes and polyhedral cores.

The fabric-based method of Lin et al. (2024) draws on taphonomic fabric analysis and treats each core as analogous to an archaeological assemblage, with each scar representing an analytical unit and the best-fit core plane serving as a reference “horizon.” This framework assesses orientational variability relative to this plane. By calculating tensor-derived metrics such as elongation and isotropy, the method characterizes scar organization along two orthogonal dimensions: dispersion within the plane and variation perpendicular to it. In the present evaluation, the fabric method identifies additional significant differences between bidirectional cores and radial cores (Levallois and discoidal), compared with SPI. This improvement arises from its explicit incorporation of vertical orientational information. However, the method fails to distinguish between unidirectional and bidirectional cores, a limitation that is directly derived from its mathematical formulation. This constraint is also reflected in the standardized Euclidean distance structure of the ideal model space, where fabric metrics cannot differentiate between parallel scar systems with opposite orientations. This indicates that fabric analysis should not be considered a replacement for SPI, but rather a complementary approach capturing a different dimension of scar organization.

In contrast, SP-SPHARM achieves the highest number of statistically significant pairwise distinctions among all tested methods. Of particular importance is its ability to separate subtypes within radial cores, notably the significant distinction between Levallois and discoidal cores. Although sample size limitations and significant PERMDISP results indicate that within-group dispersion contributes to PERMANOVA outcomes, LDA visualization nonetheless reveals a clear separation between these two categories. This discriminative power is not achieved by existing methods and likely arises from the fact that SP-SPHARM directly encodes the full spatial structure of scar organization rather than compressing it into a small set of summary variables. As a result, it is capable of capturing fine-grained phase-dependent differences in scar configuration. However, SP-SPHARM does not imply a complete replacement of existing methods. Unlike SPI and fabric metrics, SP-SPHARM allows spherical reconstruction of scar-pattern surfaces, yet these reconstructions cannot be straightforwardly mapped onto discrete scar-pattern categories. Even when idealized templates are used as references, the variability observed in real archaeological specimens makes direct categorical interpretation of reconstructed forms highly uncertain. By contrast, SPI and fabric-based indices retain stronger direct physical interpretability. Accordingly, the strength of SP-SPHARM lies in its ability to capture the global structure of scar organization with higher statistical sensitivity, making it particularly suitable for large-scale comparative analyses, such as cross-site or regional comparisons of reduction strategies. For more fine-grained archaeological interpretation, however, SPI and fabric methods remain essential complements. In practice, an effective analytical workflow is to first assess overall group differences using SP-SPHARM, and then integrate SPI and fabric-based analyses, together with typological frameworks, to refine archaeological interpretation of the observed patterns.

Building on the above validation, the present study further integrates spherical harmonic power spectra of both core morphology and scar-pattern organization into a CoIA framework to test their covariation at the assemblage level. Results from the experimental knapping cores indicate that neither the global Mantel statistic nor the RV coefficient reaches statistical significance, suggesting that no systematic covariation exists between core morphology and scar-pattern organization at the level of the entire assemblage. However, global tests may obscure structured patterns within subgroups. A more fine-grained interpretation based on the CoIA biplot reveals a more nuanced picture. Although no significant differences were detected in line lengths across all core types—indicating comparable degrees of technology–morphology decoupling among categories—the orientation of the connecting lines shows a significant non-random clustering pattern for Levallois, discoidal, and bidirectional cores. This suggests that, even though these core types do not exhibit stronger overall correspondence between morphology and scar organization than other categories, their covariation tends to unfold along internally consistent directional axes. In other words, when morphological variation occurs within these core types, scar-pattern organization tends to adjust in a predictable and structured manner along specific covariation trajectories. In the case of Levallois cores, this directional consistency may reflect intrinsic constraints linking morphology and flaking organization within the technological system. This interpretation is consistent with Éric Boëda’s concept of volumetric conception, which proposes that Levallois reduction sequences integrate volume management and flaking strategy within a unified operational logic. It should be noted, however, that the relatively small sample sizes for discoidal and bidirectional cores warrant caution in interpreting the statistical significance of these results.

Applying the above analytical framework to the SDG assemblage and comparing the results with previous studies provides a more direct evaluation of the method’s performance on real archaeological materials. Earlier work, following the core typological framework of Manuel de la Torre, proposed that lithic technology at the SDG site remained broadly continuous through time, with no major technological shifts. Two principal operational sequences were identified: (1) a small-flake reduction system primarily based on chert, and (2) a large-flake and large cutting tool (LCT) production system based on quartz sandstone. However, previous studies have not fully explored whether these two raw material systems involve distinct reduction strategies at a finer behavioral level. The results of the present spherical harmonic framework show no significant differences in either core morphology or scar-pattern organization between Layer 4 and Layer 3 at SDG, providing quantitative support for previous observational interpretations and suggesting continuity in knapping strategies across stratigraphic layers, at least at the level of reduction behavior. It should be noted that Layer 2 was excluded from statistical analysis due to insufficient sample size. At the same time, PERMANOVA results reveal a more nuanced pattern of raw material structuring in technological behavior. Cores made of chert and quartz sandstone differ significantly in morphology, but show no significant differences in scar-pattern organization. This indicates that knappers employed a consistent reduction strategy that did not systematically vary according to raw material type or size constraints. The observed morphological differences therefore likely reflect intrinsic differences in raw material properties: under a relatively uniform knapping strategy, different materials yield distinct core morphologies, without implying a corresponding change in technological organization. Importantly, despite the overall decoupling between technology and morphology, Co-inertia Analysis still reveals a significant concentration in the covariation direction for unifacial unidirectional cores. This suggests that this core type—often interpreted as a “expedient” or opportunistic reduction strategy—may, within the SDG assemblage, have been subject to specific structural constraints. The precise nature of these constraints, however, remains unclear and requires further investigation using additional analytical dimensions. Nevertheless, the present spherical harmonic framework demonstrates clear potential for identifying structured regularities that are difficult to detect using traditional analytical approaches.

At a more macro-analytical level, the methodological framework proposed in this study advances the development of three-dimensional quantitative approaches to core analysis and offers potential analytical pathways for addressing several unresolved issues in Paleolithic archaeology. Taking the debate on Levallois technology in China as an example, discussions surrounding the Guanyindong Site have included statistical analyses of cores and detached products based primarily on linear metric attributes. However, systematic three-dimensional quantitative evaluations remain lacking. Similarly, the evolutionary relationship between Levallois technology and Acheulean industries has long been a central topic in Lower–Middle Paleolithic research. Competing hypotheses regarding the origin of the Levallois concept include independent invention, derivation from Acheulean handaxe traditions, and development from earlier prepared-core technologies of the Lower Paleolithic. Recent studies employing landmark-based geometric morphometrics have compared Levallois cores, prepared cores, and Acheulean bifaces in a three-dimensional framework, suggesting that multiple regional trajectories of origin may exist. However, as discussed above, landmark configurations often vary between researchers, making cross-study integration difficult and ultimately limiting the potential for large-scale comparative analyses. The approach proposed in the present study offers a potential solution to this limitation. Although the current application focuses exclusively on cores and does not yet include large cutting tools (LCTs), the method is, in principle, fully applicable to such artifact classes. More importantly, it also has the potential to be integrated into emerging analytical paradigms. Recent work has incorporated elliptic Fourier series into phylogenetic analyses, although these applications have been restricted to two-dimensional morphologies of flaking products. As a three-dimensional generalization of Fourier methods, spherical harmonic analysis could be readily extended to phylogenetic analytical frameworks. In particular, the scar-pattern component developed in this study provides a direct line of evidence for reconstructing hominin flaking strategies. Future integration of this approach with phylogenetic modeling may therefore offer new opportunities to investigate technological transmission, population interaction, and processes of social learning among prehistoric human groups.

This study also has several limitations. As noted above, constraints in sample size mean that the current results should be interpreted primarily as exploratory, and they cannot yet be regarded as definitive empirical conclusions. In addition, at the methodological level, the present study focuses on core-level analyses, and therefore the patterns identified are relatively macroscopic in nature. The regularities revealed here should be further examined in future work through the integration of more fine-grained and targeted technological attributes. Nevertheless, the present study introduces an innovative analytical approach and a systematic research framework, which together provide an important foundation for future extensions and more detailed investigations in related research domains.

# Conclusion

# CRediT authorship contribution statement

# Data availability statement

# Funding

# Declaration of competing interest

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```
