---
title: "New methods based on SPHARM in lithic analysis"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
  - Hao Li:
      correspondence: "yes"
      orcid: 
      institute:
        - ALPHA
        - UCAS
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - UCAS: University of Chinese Academy of Sciences, Beijing 100049, China
  - ALPHA: Group of Alpine Paleoecology and Human Adaptation (ALPHA), Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing 100101; 
  - TPESRE: State Key Laboratory of Tibetan Plateau Earth System, Resources and Environment (TPESRE), Beijing 100101
  - uw: Department of Anthropology, University of Washington, Seattle, WA, USA
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
number-sections: true
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  
editor_options: 
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction

Cores preserve the most complete record of reduction strategies among all lithic artifact types, making them indispensable for understanding hominin technological behavior. Traditional analyses have relied primarily on the *chaîne opératoire* framework, employing qualitative observations of scar patterns and core morphology to infer reduction strategies. Over the past two decades, lithic studies have witnessed a marked shift toward quantitative approaches, substantially enhancing the scientific rigor and reproducibility of lithic analysis. These advances include the application of multivariate statistical methods, 2D geometric morphometric analyses based on landmarks or Fourier series, and morphological analyses derived from 3D models.

Despite these developments, existing quantitative studies have largely focused on morphological description, leaving the systematic analysis of flaking direction comparatively underdeveloped. To date, only two studies have addressed this issue directly — @clarkson2006 and @lin2024 — yet both rely on one or two variables to characterize overall scar patterns, inevitably entailing a loss of information. This limitation is evident in their capacity to discriminate certain scar configurations: the SPI method struggles to differentiate discoidal cores from Levallois cores, while the Lin's method has difficulty distinguishing unipolar from bipolar reduction patterns. More broadly, most existing approaches are methodologically fragmented, with variables selected *a priori* to match predefined technological categories. This constrains their applicability across different core types and assemblages from different periods, and ultimately limits their generalizability.

Against this background, the present study proposes a new quantitative analytical framework for cores, designed to address two interrelated problems: (1) to improve methods for analyzing flaking pattern on cores; and (2) to integrate scar pattern analysis and morphological analysis within a unified mathematical framework, supported by multivariate statistical methods. By presenting this comprehensive analytical framework, the study aims to offer a methodological contribution to the quantitative study of core analysis.

# Material and Methods

**2.1 Materials**

Ideal core models: one ideal model per core type, sourced from @lin2024.

Experimental knapping models by Clarkson: 1–22 specimens per core type, 59 specimens in total, sourced from Lin et al. (2024) (awaiting email confirmation).

Archaeological specimens from the Sandinggai site: after excluding incomplete cores and tested cores, 55 specimens remain. Specimens derive from Layers 2, 3, and 4 of the site; however, as only two specimens come from Layer 2, statistical analyses in this study are restricted to Layers 3 and 4. Data sourced from Li et al. (2022).

**2.2 Methods**

**2.2.1 Data Extraction**

Scar orientations were digitized in Geomagic Wrap 2021, and their vector coordinates were exported directly.

**2.2.2 Spherical Harmonic Analysis**

Spherical harmonic analysis represents the three-dimensional extension of elliptic Fourier analysis. Its mathematical foundations... Several studies have already applied spherical harmonic methods to lithic analysis: @sholts2017 used spherical harmonics to quantify three-dimensional asymmetry in Clovis points, while @muller2023 and Ye et al. (2026, in reviewed) applied the method to investigate the morphology of spheroids.

In the present study, we apply this method to scar orientation data for the first time, performing spherical harmonic analysis on core scars. For morphological data... For orientation data, a continuous density surface is first fitted using von Mises–Fisher spherical kernel density estimation based on directional statistics, which is then decomposed into spherical harmonics. The spherical harmonic variance and power spectrum are subsequently computed, and an appropriate spectral degree is selected by inspecting the variance contribution across degrees.

**2.2.3 Data Analysis**

*Rotation Invariance Assessment.* To evaluate whether different scar pattern metrics are rotation-invariant, we employed the Bland–Altman approach. The Bland–Altman analysis is widely used to assess agreement between two sets of measurements and provides a quantitative evaluation through the Bland–Altman plot. In this plot, the x-axis represents the mean of the two measurements, while the y-axis represents their difference. The mean difference is defined as the bias, and the 95% limits of agreement (LoA) are calculated based on this bias. A high level of agreement is inferred when the limits of agreement are narrow, approximately 95% of the data points fall within these limits, and the differences are randomly distributed without systematic trends.

In this study, the Bland–Altman framework was used to test the rotation invariance of three metrics: SPI, fabric analysis (Elongation and Isotropy), and the SPHARM power spectrum. Specifically, each metric was computed under three conditions: (1) using the original scar-vector data, (2) using morphological plane aligned data, and (3) using technological plane aligned data. Pairwise Bland–Altman comparisons were then conducted among these conditions. High agreement between the paired measurements was interpreted as evidence of rotation invariance for the corresponding metric.

*Method comparison.* To evaluate the effectiveness of spherical harmonic analysis for characterizing scar patterns, we compare the classificatory performance of three methods — SPI, Lin's fabric analysis, and spherical harmonics — using the ideal models. Since only one ideal model exists per type, we assess performance using pairwise Euclidean distances: specifically, how far apart different ideal model types are in a standardized feature space. Greater pairwise distances indicate stronger differentiation between types, thereby serving as a measure of each method's discriminatory power. Results are visualized as heatmaps, and mean pairwise distances are calculated for each method.

To evaluate the differences in information captured by each method, we applied Multiple Regression on Distance Matrices (MRM). Conceptually, MRM is equivalent to standard multiple linear regression; however, it operates on the lower-triangular elements of distance matrices as observational units and employs permutation tests instead of parametric inference to account for the non-independence among matrix entries.

The analytical procedure consisted of four steps. First, distance matrices for the three scar pattern metrics were constructed using standardized Euclidean distance. Second, each method was treated in turn as the dependent variable, with the remaining two distance matrices used as predictors, resulting in three separate MRM models. The coefficient of determination (R²) from each model represents the proportion of variance in the focal method’s distance structure explained jointly by the other two methods, whereas 1 − R² reflects the proportion of method-specific (unique) information.

Third, with the SPHARM power spectrum distance matrix as the dependent variable, we constructed a full model including both SPI and Fabric as predictors, along with two reduced models each containing a single predictor. Based on standard variance partitioning, the total variance in SPHARM was decomposed into four components: the unique contribution of SPI, the unique contribution of Fabric, the shared contribution of both predictors, and the unexplained variance. The latter component represents the proportion of information uniquely captured by SPHARM relative to SPI and Fabric.

Finally, the residual matrix was extracted from the full model. Positive residuals indicate specimen pairs for which dissimilarity in SPHARM space exceeds the joint prediction from SPI and Fabric, suggesting that SPHARM captures additional directional distribution differences for these pairs; negative residuals indicate the opposite pattern. By ranking standardized residuals, specimen pairs that most strongly reflect the unique discriminative capacity of SPHARM were identified and subsequently visualized.

*Dimensionality reduction.* Because spherical harmonics describe a nonlinear manifold, linear methods such as PCA are ill-suited to recovering the global structure of spherical harmonic data. We therefore apply UMAP to reduce the dimensionality of the spherical harmonic power spectra.

*Correlation analysis and Mantel test.* Spearman rank correlation coefficients and Mantel tests are used to explore whether systematic associations exist between the technological and morphological characteristics of cores.

*Co-inertia analysis.* This method performs joint dimensionality reduction on two data matrices simultaneously, and is used here to interpret the overall covariation between technological and morphological patterns across cores.

# Results

**3.1 Validity Analysis**

**3.1.1 Rotational Invariance of Scar Pattern Analysis Metrics**

To examine the rotational invariance of scar pattern analysis methods, we computed each metric under three coordinate conditions: the original coordinate system, a coordinate system aligned to the morphology fitted plane, and one aligned to the scar pattern fitted plane of scar orientations. Bland-Altman analysis was applied to assess agreement across coordinate conditions. SPI and Fabric values showed near-perfect agreement across all three conditions, with biases and 95% limits of agreement (LoA) on the order of 10⁻¹⁶—well within the expected range of floating-point numerical precision (Fig. X). This confirms that both SPI and Fabric are strictly rotationally invariant, and that consistent results can be obtained without coordinate alignment of the core model prior to analysis.

The SPHARM power spectrum exhibited limited variation across alignment methods. Bland-Altman analysis revealed small systematic biases in low-order power (degrees l = 1–2; 95% LoA = ±0.02), which converged rapidly with increasing degree; no substantive differences were observed among comparison groups for l ≥ 3 (Fig. X). To further evaluate this pattern, we applied random rotational perturbations about the Z-axis to scar pattern aligned data and recomputed the SPHARM power spectra. Bland-Altman analysis showed that differences between perturbed and unperturbed spectra remained within machine precision, indicating that directional uncertainty within the coordinate frame, given a fixed Z-axis, has no effect on the power spectrum.

These results suggest that the SPHARM power spectrum is theoretically rotationally invariant by design, and that the systematic bias observed in low-order terms more likely originates from upstream computational steps such as KDE interpolation and coordinate mapping. Crucially, when a consistent alignment method is applied uniformly across all specimens, this systematic error affects all specimens equally and does not alter relative differences among them, leaving the validity of subsequent statistical analyses unaffected. Accordingly, all subsequent analyses in this study employ the scar pattern fitted plane for coordinate alignment.

**3.1.2 Validity of the SPHARM Method for Scar Pattern Analysis**

```{r}
#| label: setup-im-comparison
#| include: false

library(here)
library(tidyverse)
library(readxl)

# ---- 公共函数 ----
compute_R <- function(dx, dy, dz) {
  resultant_magnitude <- sqrt(sum(dx)^2 + sum(dy)^2 + sum(dz)^2)
  total_length        <- sum(sqrt(dx^2 + dy^2 + dz^2))
  resultant_magnitude / total_length
}

compute_EI <- function(ux, uy, uz) {
  n      <- length(ux)
  U      <- cbind(ux, uy, uz)
  T_mat  <- (t(U) %*% U) / n
  eig    <- eigen(T_mat, symmetric = TRUE)
  lambda <- sort(eig$values, decreasing = TRUE)
  lambda <- pmax(lambda, 0)
  list(
    E = ifelse(lambda[1] > 1e-10, 1 - lambda[2] / lambda[1], NA_real_),
    I = ifelse(lambda[1] > 1e-10,     lambda[3] / lambda[1], NA_real_)
  )
}

# ---- 读入数据 ----
raw <- read_excel(
  here("analysis/data/raw_data/Scar_orientation_data.xlsx"), sheet = 1)
SPHARM_direction <- read_csv(
  here("analysis/data/derived_data/SPHARM_direction.csv"),
  show_col_types = FALSE)

# ---- IM 标本 SPHARM ----
SPHARM_IM <- SPHARM_direction %>%
  filter(str_starts(ID, "IM_")) %>%
  select(ID, Typology, spectral_entropy, power_l1:power_l4)

# ---- IM 标本 Fabric（R/E/I）----
fabric_IM <- raw %>%
  filter(str_starts(ID, "IM_")) %>%
  mutate(
    dx  = End_X - Start_X,
    dy  = End_Y - Start_Y,
    dz  = End_Z - Start_Z,
    len = sqrt(dx^2 + dy^2 + dz^2)
  ) %>%
  filter(len > 1e-10) %>%
  mutate(ux = dx / len, uy = dy / len, uz = dz / len) %>%
  group_by(ID) %>%
  summarise(
    R = compute_R(ux, uy, uz),
    E = compute_EI(ux, uy, uz)$E,
    I = compute_EI(ux, uy, uz)$I,
    .groups = "drop"
  )

# ---- 合并 ----
df_im <- SPHARM_IM %>%
  left_join(fabric_IM, by = "ID") %>%
  mutate(
    label = ID %>%
      str_remove("^IM_") %>%
      str_replace_all("_", " ") %>%
      str_to_sentence()
  )

# ---- 距离矩阵 ----
id_order <- c(
  "Cylindrical unipolar cortical", "Cylindrical unipolar scarred",
  "Cylindrical bipolar",
  "Conical unipolar cortical",     "Conical unipolar scarred",
  "Discoid",                       "Discoid unifacial",
  "Levallois preferential",        "Levallois convergent",
  "Levallois laminar",
  "Biface",                        "Multiplatform"
)

make_dist_df <- function(X, labels, method_name) {
  d <- as.matrix(dist(scale(X), method = "euclidean"))
  rownames(d) <- colnames(d) <- labels
  as.data.frame(d) %>%
    rownames_to_column("From") %>%
    pivot_longer(-From, names_to = "To", values_to = "distance") %>%
    mutate(method = method_name)
}

labs <- df_im$label

dist_all <- bind_rows(
  make_dist_df(df_im %>% select(R)                 %>% as.matrix(), labs, "SPI"),
  make_dist_df(df_im %>% select(E, I)              %>% as.matrix(), labs, "Fabric"),
  make_dist_df(df_im %>% select(power_l1:power_l4) %>% as.matrix(), labs, "SPHARM")
) %>%
  mutate(
    From   = factor(From,   levels = id_order),
    To     = factor(To,     levels = id_order),
    method = factor(method, levels = c("SPI", "Fabric", "SPHARM"))
  )

# ---- 平均距离 ----
avg_dist <- dist_all %>%
  filter(From != To) %>%
  group_by(method) %>%
  summarise(mean_dist = round(mean(distance), 2), .groups = "drop")

dist_spi    <- avg_dist$mean_dist[avg_dist$method == "SPI"]
dist_fabric <- avg_dist$mean_dist[avg_dist$method == "Fabric"]
dist_spharm <- avg_dist$mean_dist[avg_dist$method == "SPHARM"]
```

```{r}
#| label: fig-im-comparison
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| fig-cap: !expr >
#|   paste0("Pairwise standardised Euclidean distances between ideal core 
#|   models across three methods. Mean distances: SPI = ", dist_spi,
#|   ", Fabric = ", dist_fabric, ", SPHARM = ", dist_spharm, ".")

ggplot(dist_all, aes(x = To, y = From, fill = distance)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(
    aes(label = ifelse(From != To, sprintf("%.1f", distance), "")),
    size = 2.2
  ) +
  facet_wrap(~ method, ncol = 3) +
  scale_fill_gradient(
    low  = "white", high = "#E84D89",
    name = "Euclidean\ndistance\n(standardised)"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  scale_y_discrete(limits = rev) +
  theme_bw(base_size = 8) +
  labs(x = NULL, y = NULL) +
  theme(
    strip.text       = element_text(face = "bold", size = 9),
    strip.background = element_rect(fill = "grey80", color = "grey80"),
    axis.text        = element_text(size = 6.5)
  )
```

**Ideal model tests** Using idealized geometric models, we computed SPI, Fabric, and SPHARM metrics and evaluated their discriminative performance. Heatmaps revealed a progressive increase in discriminative capacity from SPI to Fabric to SPHARM (@fig-im-comparison). This ordering was corroborated by mean Euclidean distance values: the SPHARM method yielded the highest mean inter-group distance (mean distance = `r dist_spharm`), substantially exceeding those of SPI (mean distance = `r dist_spi`) and Fabric (mean distance = `r dist_fabric`). These results demonstrate that, under idealized patterns, SPHARM most effectively distinguishes among different scar distribution structures.

**Experimental core tests** To evaluate the performance of the three methods on empirical data, we computed each metric for experimental knapped cores and conducted statistical tests by core type: a Kruskal-Wallis test for SPI, and PERMANOVA for the Fabric bivariate distribution and the SPHARM scar pattern spectrum. All three methods demonstrated a capacity to detect inter-type differences (Fig. X). Post-hoc pairwise comparisons with Holm-Bonferroni correction are presented in Fig. X. The SPHARM scar pattern spectrum achieved the highest overall discriminative power, successfully identifying the greatest number of significant pairwise differences; only the Discoid–Unidirectional and Discoid–Multiplatform pairs failed to reach significance. Fabric distinguished the majority of type pairs but could not differentiate Unidirectional cores from Bidirectional cores. SPI detected only those differences involving Unidirectional cores.

These results indicate that the three methods exhibit a complementary relationship in type recognition. SPHARM captures the full structure of scar patterning and is particularly sensitive to differences between Discoid and Levallois cores, though its discriminative power is limited for patterns with bipolar symmetry. Fabric effectively extracts differences between Unidirectional and Bidirectional types relative to Levallois and Discoid cores, but cannot distinguish between Unidirectional and Bidirectional cores, nor between Levallois and Discoid cores. SPI is most sensitive to the Unidirectional type but shows limited discriminative capacity for other types. Notably, SPHARM failed to distinguish Discoid from Multiplatform cores, as both exhibit symmetric scar distributions and consequently produce similar power spectrum structures (@fig-im-comparison), a distinction that Fabric was able to recover.

**3.2 Joint Analysis of Scar Patterns and Core Morphology in Experimental Knapping Specimens**

The analyses above demonstrate that scar pattern SPHARM spectra exhibit broadly valid and significant differences across core types. To explore morphological variation among types, we further applied PERMANOVA tests using morphological SPHARM spectra. No significant differences in overall core morphology were detected across types, suggesting that core morphology and reduction strategies may constitute two independent dimensions of variation. To test this hypothesis, we integrated morphological and scar pattern SPHARM spectra using Mantel tests and Co-Inertia Analysis (CoIA). The global Mantel test revealed no significant correlation between morphological and scar pattern SPHARM spectra; however, the RV coefficient derived from CoIA was statistically significant (RV = 0.140, P = 0.004). Taken together, these results suggest that core morphology and scar pattern represent largely independent dimensions of variation, while still exhibiting a modest degree of covariation.

To rule out the possibility that global structure might obscure local signals, we conducted within-group Mantel tests stratified by core type. No significant correlations were found in any type (Table X). Nevertheless, several types displayed notably large effect sizes, including Discoid (R = 0.333), Bidirectional (R = −0.361), and Multiplatform (R = −0.411) cores, suggesting potential associative trends between morphology and scar patterning within these groups, whose non-significance may reflect limited sample size.

The CoIA ordination provided more detailed insight into the covariation structure between morphology and scar pattern. The first two CoIA axes jointly explained 91.7% of total covariance (Fig. X). The first axis, accounting for 78.9% of covariance, captures the covariation between overall regularity of core morphology and scar pattern. Interpreted in conjunction with the PCA results for each spectrum independently, the positive end of this axis represents specimens whose power is concentrated in low-order frequencies—smooth and regular in morphology, with highly concentrated removals, corresponding to unidirectional cores; the negative end represents specimens with more complex morphology and dispersed scar patterns, trending toward multiplatform reduction (Fig. X). The second axis, explaining 12.8% of covariance, captures covariation along a symmetry dimension. Its positive end represents specimens in which first-order spherical harmonic power predominates over second-order, morphologically approximating a uniform sphere, with non-symmetric scar distributions, as in Unidirectional and Unifacial Discoid types; the negative end represents specimens with a pronounced long axis (ellipsoidal or flatten morphology) and symmetrically distributed scars, as in Bifacial Discoid types (Fig. X).

In the CoIA biplot, each specimen's morphological and scar pattern SPHARM spectra are mapped into the same feature space; the length of the line connecting the two projections reflects the degree of coupling between morphology and scar pattern, with shorter lines indicating closer correspondence. A Kruskal-Wallis test revealed no significant difference in line lengths across core types (Fig. X). Discoid cores showed a relatively high median line length, suggesting a greater degree of decoupling between morphology and scar patterning, though this trend requires verification with larger samples. The spatial orientation of the lines reveals the direction of techno-morphological covariation within each type: when the lines of a given type are concentrated, this implies a consistent axis of techno-morphological variation, potentially indicative of an underlying mental template. Rayleigh tests showed that the line orientations for Levallois cores were significantly concentrated, with a mean direction of 122° (Fig. X). Given the archaeological interpretation of the CoIA axes, this covariation trajectory extends primarily along the positive direction of the second axis, indicating that Levallois technology is consistently associated with concentrated scar directionality and morphological uniformity. A secondary component along the negative first axis further suggests that Levallois cores also tend toward greater surface complexity and more structured scar pattern.

To quantify the structural complexity of the power spectra, we computed spectral entropy for each specimen. Kruskal-Wallis tests showed that the scar pattern spectral entropy of Levallois cores was significantly lower than that of all other types except Unidirectional cores, indicating that Levallois scar pattern exhibits the highest structural regularity. Read in conjunction with the CoIA results, this suggests that while Levallois reduction sequences are organizationally complex, they conform to a highly structured set of operational rules. By contrast, morphological spectral entropy did not differ significantly across types, indicating no systematic differences in morphological complexity among core types (Fig. X).

**3.3 Archaeological Case Study: Core Analysis from Sandinggai**

Based on the analytical framework established in Section 3.2, we conducted a joint analysis of core morphology and scar patterns on 55 cores excavated from Sandinggai.

At the global level, no significant systematic covariation was observed between core morphology and scar patterns (Mantel r = 0.009, P = 0.375; RV = 0.047, P = 0.500), indicating that variations in morphology and scar patterns are largely independent. Cross-frequency Mantel tests also yielded non-significant results after FDR correction, suggesting that this independence is consistent across all frequency bands rather than being masked by specific frequency signals.

To assess whether this independence holds under different conditions, the dataset was stratified by layer, raw material, and core type, and Mantel tests were conducted within each subgroup. Results showed that none of the subgroups exhibited significant covariation after FDR correction, indicating that the independence between morphology and scar patterns is robust across different time periods, raw materials, and core types. Notably, the Unifacial unidirectional cores exhibited the highest Mantel r value (r = 0.192) among the subgroups; although non-significant due to a small sample size (n = 8), this may suggest a potential morphology–technology covariation that warrants further investigation with larger samples.

To further explore the factors driving variation in morphology and scar patterns, PERMANOVA tests were conducted for the three grouping variables. The results revealed a differential pattern of influence across the two dimensions. Raw material significantly affected morphology (R² = 0.036, P \< 0.001) but not scar pattern (R² = 0.016, P = 0.966), indicating systematic three-dimensional differences between chert and sandstone cores, while the reduction strategies employed by knappers remained consistent across materials. This suggests that the physical properties of raw materials constrain core morphology, whereas knappers’ reduction strategies are stable and independent of raw material characteristics.

Core type also had a significant effect on morphology (R² = 0.135, P = 0.003), but its influence on scar patterns did not reach significance (R² = 0.119, P = 0.198). Although the effect size for scar patterns is not negligible and is close to that observed for morphology, the multiple core type groups and limited sample sizes in some categories may reduce statistical power, and thus differences in scar pattern among core types cannot yet be fully excluded. Layer had no significant effect on either morphology (R² = 0.017, P = 0.951) or scar patterns (R² = 0.019, P = 0.309), indicating continuity in reduction strategies between Layers 3 and 4 with no apparent technological shifts.

Further, to investigate variation in arrow features derived from CoIA dimensionality reduction under different grouping conditions, we examined arrow length and direction separately. Kruskal–Wallis tests on arrow length showed no significant differences among core types, raw materials, or layers (core type: χ² = 4.7392, df = 6, P = 0.5777; raw material: χ² = 0.1978, df = 1, P = 0.6565; layer: χ² = 0.9104, df = 1, P = 0.3400), indicating overall consistency in arrow length across groups.

Analysis of arrow direction using Rayleigh and Watson tests revealed no significant differences between raw materials or layers, and most core types also showed no directional variation. The exception was the Unifacial unidirectional cores, which exhibited a significantly concentrated directionality (Rayleigh P = 0.0036), suggesting more focused variation within this type. Notably, this core type also displayed a relatively high Mantel effect size (r = 0.192), it indicates a potential morphology–technology covariation that warrants verification in larger samples.

Overall, the core assemblage from Sandinggai demonstrates a systematic decoupling between morphology and reduction strategy. This decoupling is not restricted to specific conditions but appears to be a general pattern across different time periods, raw materials, and core types. Morphological variation is primarily associated with raw material and core type, whereas variation in scar pattern does not show a significant response to these factors. These results suggest that reduction strategies may be relatively stable and partly independent of raw material or core type variation, potentially reflecting a consistent technological strategy pattern adopted by knappers during core reduction.

# Discussion

# Conclusion

# CRediT authorship contribution statement

# Data availability statement

# Funding

# Declaration of competing interest

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```
