---
title: "Reproduction Report"
subtitle: "Every statistical result printed by `targets::tar_make()`. The manuscript reports only a subset for reasons of space; this is the complete record, for review and verification."
date: now
date-format: long
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    embed-resources: true
    grid:
      body-width: 1100px
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
.module { border: 1px solid #e1e4e8; border-radius: 8px; margin: 1rem 0 1.5rem;
          background: #fff; box-shadow: 0 1px 2px rgba(27,31,35,.04); }
.module > summary { cursor: pointer; list-style: none; font-weight: 600;
          padding: .5rem .9rem; color: #24467a; background: #f6f8fa;
          border-radius: 8px; }
.module[open] > summary { border-bottom: 1px solid #e1e4e8; border-radius: 8px 8px 0 0; }
.module > summary::-webkit-details-marker { display: none; }
.module > summary::before { content: "▸ "; }
.module[open] > summary::before { content: "▾ "; }
pre.results { margin: 0; max-height: 460px; overflow: auto;
          background: #fbfcfd; border: none; border-radius: 0 0 8px 8px;
          padding: .8rem 1rem; font-size: .80rem; line-height: 1.5;
          white-space: pre; }
p.what { color: #57606a; font-style: italic; margin: .25rem 0 .6rem; }
h2 { margin-top: 2rem; padding-top: .4rem; }
</style>
```

```{r}
#| label: setup
#| include: false

# Reproduce the pipeline's package environment so the analysis scripts run
# exactly as they do inside tar_make(). All package / conflict chatter is
# suppressed here so it never reaches the report.
suppressWarnings(suppressMessages(suppressPackageStartupMessages({
  library(targets)
  library(tidyverse)
  library(patchwork)
  library(here)
  library(readxl)
  library(vegan)
  library(FSA)
  library(RVAideMemoire)
  library(ggrepel)
  library(conflicted)
  library(linkET)
  library(compositions)
  library(ade4)
  library(circular)
  library(rsvg)
  library(png)
  library(grid)
  library(htmltools)
  conflicts_prefer(dplyr::filter)
  conflicts_prefer(dplyr::select)
  conflicts_prefer(dplyr::lag)
  conflicts_prefer(stats::sd)
  conflicts_prefer(stats::var)
  conflicts_prefer(stats::dist)
  conflicts_prefer(stats::cor)
  conflicts_prefer(stats::cov)
  conflicts_prefer(stats::anova)
  conflicts_prefer(graphics::segments)
  conflicts_prefer(base::norm)
  conflicts_prefer(base::scale)
  conflicts_prefer(base::`%*%`)
  conflicts_prefer(ggplot2::aes)
  conflicts_prefer(ggplot2::theme_bw)
  conflicts_prefer(ggplot2::ggsave)
  conflicts_prefer(ggplot2::annotate)
  conflicts_prefer(ggplot2::theme_classic)
  conflicts_prefer(ggplot2::theme_minimal)
  conflicts_prefer(ggplot2::theme_void)
  conflicts_prefer(ggplot2::theme_grey)
  conflicts_prefer(ggplot2::theme_gray)
  conflicts_prefer(ggplot2::theme_light)
  conflicts_prefer(ggplot2::theme_dark)
  conflicts_prefer(ggplot2::theme_linedraw)
})))

# Print full tables (no "# i N more rows/variables" truncation) and no ANSI
# colour codes (which would show as garbage inside <pre>).
options(width = 110, max.print = 1e5,
        crayon.enabled = FALSE, cli.num_colors = 1,
        tibble.print_max = Inf, tibble.print_min = Inf,
        pillar.advice = FALSE, pillar.max_footer_lines = 0,
        dplyr.summarise.inform = FALSE)

knitr::opts_chunk$set(comment = "", message = FALSE, warning = FALSE)

SCRIPT <- "/project/_targets.R"
STORE  <- "/project/analysis/paper/_targets"

for (.d in c("analysis/data/derived_data", "analysis/output/figures", "analysis/output/html")) {
  dir.create(here::here(.d), recursive = TRUE, showWarnings = FALSE)
}

# Run one analysis script (the same one tar_make() sources for that target),
# reading the already-computed derived_data CSVs — no Python is re-run — and
# capture its statistical console output as clean text. Plots go to a null
# device; conflict/truncation chatter is stripped.
run_capture <- function(rel) {
  grDevices::pdf(base::nullfile())
  on.exit(grDevices::dev.off(), add = TRUE)
  out <- tryCatch(
    capture.output(suppressWarnings(suppressMessages(
      local(source(here::here(rel), local = TRUE))
    ))),
    error = function(e) paste("[Could not reproduce this module]", conditionMessage(e))
  )
  out <- out[!grepl("Will prefer|\\[conflicted\\]|ℹ", out)]   # conflicted + tibble "i" footers
  # trim leading / trailing blank lines and collapse runs of >2 blanks
  out <- out[cumsum(nzchar(trimws(out))) > 0]
  out <- rev(rev(out)[cumsum(nzchar(trimws(rev(out)))) > 0])
  paste(out, collapse = "\n")
}

# Emit a module as a heading (-> appears in the TOC) + one-line description +
# a collapsible, scrollable panel of its statistical output.
emit_module <- function(title, what, rel) {
  body <- htmltools::htmlEscape(run_capture(rel))
  cat(sprintf("\n\n## %s\n\n", title))
  cat(sprintf("<p class=\"what\">%s</p>\n", htmltools::htmlEscape(what)))
  cat(sprintf(
    "<details class=\"module\" open>\n<summary>Statistical output</summary>\n<pre class=\"results\">%s</pre>\n</details>\n\n",
    body))
}
```

This report gathers the **complete statistical output** that scrolls past in the
console during `targets::tar_make()`. The article presents only selected
results; here every test, table, and summary is reproduced in full. Each module
below re-runs the exact R script `tar_make()` uses for that analysis (reading the
already-computed power-spectrum tables — no Python is re-run) and shows its
output in a collapsible panel.

# Pipeline map

`tar_make()` executes in dependency order. The interactive graph and the table
below show how the modules relate.

```{r}
#| label: visnetwork
invisible(capture.output(suppressWarnings(suppressMessages(
  vn <- targets::tar_visnetwork(targets_only = TRUE, label = "description",
                                script = SCRIPT, store = STORE,
                                callr_function = NULL)
))))
vn
```

```{r}
#| label: tbl-steps
invisible(capture.output(suppressWarnings(suppressMessages(
  m <- tryCatch(
    tar_manifest(fields = c("name", "description"), script = SCRIPT, store = STORE,
                 callr_function = NULL),
    error = function(e) NULL)
))))
if (!is.null(m)) {
  knitr::kable(m, col.names = c("Target", "Description"))
}
```

# Inputs (data preparation)

Two preparatory steps produce the inputs to the statistics below; they write
data files rather than printing statistical results, so their console output is
not reproduced here:

- **Scar-direction alignment** writes the aligned scar vectors to
  `analysis/data/derived_data/directions_aligned_svd.csv` and
  `directions_aligned_lin2024.csv`.
- **Python SPHARM** writes the morphology and scar-direction power spectra to
  `analysis/data/derived_data/SPHARM_morphology.csv` and
  `SPHARM_direction.csv`.

# Statistical analyses

```{r}
#| label: modules
#| output: asis
#| fig-show: hide

emit_module(
  "Method comparison and PERMANOVA (experimental cores)",
  "Compares SPI, fabric and SPHARM descriptors and tests group differences with PERMANOVA.",
  "analysis/scripts/r_spharm/spharm_analysis.R")

emit_module(
  "Comparing SPI / fabric / SPHARM on idealized models",
  "Quantifies how completely each descriptor separates the idealized core types.",
  "analysis/scripts/r_validation/methods_comparison_IM.R")

emit_module(
  "Rotational-invariance validation",
  "Checks that SP-SPHARM power spectra are invariant to core orientation across alignments.",
  "analysis/scripts/r_validation/validate_rotation_all.R")

emit_module(
  "Experimental cores: co-inertia, Mantel and circular statistics",
  "Co-inertia of morphology versus scar pattern, Mantel tests, and directional (circular) statistics.",
  "analysis/scripts/r_statistics/exp_cores_statistics.R")

emit_module(
  "Sandinggai (SDG) cores: co-inertia, PERMANOVA and circular statistics",
  "The same battery applied to the archaeological assemblage, by core type, raw material and layer.",
  "analysis/scripts/r_statistics/SDG_cores_statistics.R")

emit_module(
  "Degree-selection diagnostic",
  "Per-degree across-specimen CV and cumulative power justifying the harmonic truncation order.",
  "analysis/scripts/r_spharm/power_degree_selection.R")
```

---

*Generated from the `targets` pipeline. To inspect the dependency graph
interactively, run `targets::tar_visnetwork()` in the project.*
