---
title: "Supplementary Information"
subtitle: "Characterizing core morphology and scar patterning within a unified spherical harmonic framework"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - uw
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      orcid: 0009-0008-1504-9401
      institute:
        - ALPHA
        - MOE
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China;
  - MOE: MOE Key Laboratory of Western China's Environmental System, Center for Excellence in Archaeological Science and Cultural Heritage, Lanzhou University, Lanzhou, China
  - uw: Department of Anthropology, University of Washington, Seattle, WA, USA
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
# Separate doc → numbering restarts at 1; the "S" prefix gives Figure S1 / Table S1.
# Sections are manually labelled S1–S3, so section auto-numbering is off.
crossref:
  fig-title: "Figure S"
  fig-prefix: "Figure S"
  tbl-title: "Table S"
  tbl-prefix: "Table S"
number-sections: false
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 800
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-method-and-theory.csl"
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(targets)
library(tidyverse)
library(patchwork)

# Two independent target stores (see _targets.yaml).
si_store   <- "/project/analysis/robustness/_targets_store"   # this SI pipeline
main_store <- "/project/analysis/paper/_targets"              # main-text pipeline
tar_config_set(store = si_store)

# Robustness targets (built by: TAR_PROJECT=supplementary tar_make()).
bw   <- tar_read(robustness_bandwidth)
scar <- tar_read(robustness_scar_threshold)
mesh <- tar_read(robustness_mesh_decimation)

# Curated metric tables, rendered live from the targets so they always agree with
# the figures (and, for S1/S2, with the deterministic values quoted in the text).
pm <- function(r, p) sprintf("%.3f (%.3f)", r, p)   # "estimate (p)"
```

# Sensitivity and robustness analyses

We assessed whether the headline results depend on the principal free parameters of the two
descriptors. For the scar descriptor (SP-SPHARM) we varied (S1) the von Mises–Fisher kernel
bandwidth $h$ and (S2) the scar minimum-size inclusion threshold. For the morphology
descriptor (M-SPHARM) we varied (S3) the mesh decimation target and the amount of Laplacian
smoothing, and we verified that the morphology pipeline is deterministic. In every sweep all
other settings were held at their production values, and at the production setting the
recomputed descriptors reproduce the cached values, anchoring each sweep to the main
analysis. As in the main text, results are interpreted on the magnitude of effects rather
than on significance thresholds; the analyses below repeatedly show that what moves under
parameter perturbation is the *significance* of small-to-modest effects near the 0.05 line,
not their *magnitude*.

## S1. Bandwidth of the von Mises–Fisher kernel ($h$) — scar descriptor

**Methods.** The spherical kernel-density bandwidth was fixed at $h = 0.35$ (concentration
$\kappa = 1/h^2 = 8.16$) in the main analysis. We recomputed the SP-SPHARM power spectra over
$h \in \{0.20, 0.25, 0.30, 0.35, 0.40, 0.45, 0.50\}$ ($\kappa = 4.0$–$25.0$), holding
everything else fixed (alignment, the spherical evaluation grid and Driscoll–Healy transform,
$l_{\max} = 20$, normalisation, and seeds). Morphology (M-SPHARM) is independent of $h$ and
was reused unchanged. At $h = 0.35$ the recomputed spectra reproduce the cached values.

**Order selection (truncation at $l = 1$–$6$).** Across all $h$, SP-SPHARM power saturates
early: cumulative power through $l = 6$ is 96.71–99.99% (experimental assemblage, EXP), and
the cross-specimen coefficient of variation (CV) first exceeds 100% at degree $l = 7$–$14$.
The $l = 1$–$6$ truncation is robust to $h$; it is most complete at moderate-to-large $h$ (at
the smallest $h$ a little reliable signal persists beyond $l = 6$).

```{r}
#| label: fig-si-bandwidth-order
#| fig-cap: "SP-SPHARM power saturation (cumulative power, top) and cross-specimen CV (bottom) by spherical-harmonic degree $l$, overlaid across bandwidths $h$, for the experimental (EXP) and archaeological (SDG) assemblages. Dashed lines mark the 95%/99% power and 100% CV references."
#| fig-width: 9
#| fig-height: 8
bw$fig_orderselection
```

**Ideal-core discriminability.** The pairwise standardised-distance structure is highly
stable: each $h$'s distance matrix correlates with the $h = 0.35$ matrix at Pearson
$r \ge 0.961$, and key separations persist at every $h$ (e.g. discoidal vs Levallois,
bifacial vs unifacial discoidal).

```{r}
#| label: fig-si-bandwidth-im
#| fig-cap: "Ideal-core standardised distance matrices (power $l = 1$–$4$) at $h = 0.20$, $0.35$ and $0.50$; the key type separations persist across bandwidths."
#| fig-width: 11
#| fig-height: 4.5
bw$fig_im_heatmaps
```

**Experimental cores (PERMANOVA).** The scar-pattern PERMANOVA by core type stays significant
across $h$ ($R^2 = 0.259$–$0.374$; $p < 0.05$ throughout). The monotonic rise of $R^2$ with
$h$ is a dimensionality effect — stronger smoothing concentrates spectral energy into fewer
effective degrees, inflating the between-group share of total variance — rather than improved
discrimination, consistent with the caveat in the main text that $R^2$ is comparable only
within a fixed effective dimensionality. The pairwise resolution profile is consistent across
the central range: 8 of 10 typology pairs are resolved at $h = 0.30$–$0.45$ (including the
chosen $h = 0.35$), dropping to 6/10 at the two smallest bandwidths and 7/10 at the largest.

**Decoupling (Mantel + RV).** The association between morphology and scar patterning is
negligible in magnitude at every bandwidth: Mantel $r \approx 0$ throughout (EXP $-0.062$ to
$-0.008$; SDG $0.001$ to $0.017$) and RV $\approx 0.10$–$0.12$ (EXP) and $\approx 0.09$ (SDG).
For SDG both measures are non-significant across the entire range. For the experimental
assemblage the Mantel test is non-significant at every $h$, and the RV coefficient is
non-significant from $h = 0.20$ to $0.40$, becomes marginal at $h = 0.45$ ($p = 0.078$), and
crosses $p < 0.05$ only at the most over-smoothed setting ($h = 0.50$: RV $= 0.117$,
$p = 0.044$) — where the association is still negligible and is not corroborated by the Mantel
test ($r = -0.008$, $p = 0.54$). Read on effect size, the decoupling is robust across the full
range, and the chosen $h = 0.35$ sits well inside the non-significant regime. The SDG
scar–core-type PERMANOVA remains significant throughout ($R^2 = 0.166$–$0.177$).

```{r}
#| label: fig-si-bandwidth-summary
#| fig-cap: "Downstream metrics versus bandwidth $h$ (dashed vertical = production $h = 0.35$): EXP/SDG PERMANOVA $R^2$, Mantel $r$, RV, and the ideal-core distance-matrix correlation against the $h = 0.35$ structure."
#| fig-width: 9
#| fig-height: 8
bw$fig_summary
```

```{r}
#| label: tbl-si-bandwidth
#| tbl-cap: "Per-bandwidth sensitivity metrics (`*` = production setting, $h = 0.35$). Cumulative power, the CV-crossing degree, the ideal-core (IM) correlation, $R^2$, and the Mantel/RV estimates are deterministic; permutation $p$-values use a fixed seed."
bw$metrics |>
  arrange(h) |>
  transmute(
    `h`                = ifelse(abs(h - 0.35) < 1e-9, paste0(format(h, nsmall = 2), "*"), format(h, nsmall = 2)),
    `κ`           = sprintf("%.2f", kappa),
    `EXP cum%@l6`      = sprintf("%.2f", exp_cumpower_l6),
    `CV>100% at l`     = exp_cv_cross_l,
    `IM corr vs .35`   = sprintf("%.3f", im_corr_vs_ref),
    `EXP R2`           = sprintf("%.3f", exp_perm_R2),
    `EXP p`            = sprintf("%.3f", exp_perm_p),
    `EXP sig pairs`    = paste0(exp_perm_nsig, "/", exp_perm_npairs),
    `SDG type R2`      = sprintf("%.3f", sdg_perm_scar_coretype_R2),
    `SDG p`            = sprintf("%.3f", sdg_perm_scar_coretype_p),
    `EXP Mantel r (p)` = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`       = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)` = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`       = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## S2. Scar minimum-size inclusion threshold — scar descriptor

**Methods.** The main pipeline applies no scar-size filter in code, so the analysed set
contains every recorded scar (smallest $\approx 2$ mm). To test whether the scar-patterning
results depend on this inclusion rule — and in particular whether they are driven by small,
less reliably measured scars — we recomputed the SP-SPHARM power spectra after dropping scars
at or below increasing cutoffs $T \in \{0, 5, 10\}$ mm (retaining length $> T$), holding
everything else fixed (alignment, $h = 0.35$, the evaluation and transform grids,
$l_{\max} = 20$, normalisation, seeds). Morphology was reused unchanged. $T = 0$ (all recorded
scars) is the production anchor and reproduces the cached values. Relative to the full set,
the $>5$ mm rule retains 95.9% (EXP) and 99.2% (SDG) of scars and $>10$ mm retains 80.7% and
88.5%; no EXP or SDG specimen falls below the three-scar floor. The threshold is not applied
to the synthetic ideal cores, whose scar lengths are fixed non-physical values, so their
discriminability is unchanged by construction.

```{r}
#| label: fig-si-scar-counts
#| fig-cap: "Per-specimen scar retention by minimum-size threshold $T$; the dashed line is the three-scar advisory floor, below which no EXP/SDG specimen falls."
#| fig-width: 9
#| fig-height: 4.5
scar$fig_scarcounts
```

**Order selection (truncation at $l = 1$–$6$).** SP-SPHARM power saturates early at every
threshold: cumulative power through $l = 6$ is 99.91–99.92% (EXP), and the cross-specimen CV
first exceeds 100% at degree $l = 9$ for all $T$. The $l = 1$–$6$ truncation is robust to the
inclusion rule.

```{r}
#| label: fig-si-scar-order
#| fig-cap: "SP-SPHARM power saturation and cross-specimen CV by degree across scar minimum-size thresholds $T$, for EXP and SDG."
#| fig-width: 9
#| fig-height: 8
scar$fig_orderselection
```

**Experimental cores (PERMANOVA).** The scar-pattern PERMANOVA by core type is significant at
every threshold ($R^2 = 0.307$–$0.360$; $p < 0.05$ throughout). The pairwise resolution
profile is 8/10 typology pairs at the production baseline ($T = 0$), 6/10 at $>5$ mm, and
10/10 at $>10$ mm. Restricting the analysis to larger, more reliably measured scars therefore
does not weaken type discrimination and, at $>10$ mm, sharpens it; the experimental core-type
signal is carried by the larger scars and is not an artefact of small scars.

**Decoupling (Mantel + RV).** Morphology and scar patterning remain negligibly associated and
non-significant at every threshold: Mantel $r \approx 0$ (EXP $-0.061$ to $0.004$; SDG
$-0.047$ to $0.012$) and RV $= 0.101$–$0.109$ (EXP) and $0.057$–$0.091$ (SDG), all
$p > 0.05$. The SDG scar–core-type PERMANOVA is significant throughout ($R^2 = 0.168$–$0.171$).

```{r}
#| label: fig-si-scar-summary
#| fig-cap: "Downstream metrics versus scar minimum-size threshold $T$ (dashed = production $T = 0$, all scars)."
#| fig-width: 9
#| fig-height: 7
scar$fig_summary
```

```{r}
#| label: tbl-si-scar
#| tbl-cap: "Per-threshold sensitivity metrics (`*` = production baseline, all recorded scars). EXP/SDG scar columns are retained scar totals. Ideal cores held at production values (threshold not applicable). Permutation $p$-values use a fixed seed."
scar$metrics |>
  arrange(threshold_mm) |>
  transmute(
    `T (mm)`           = ifelse(abs(threshold_mm) < 1e-9, "0*", as.character(as.integer(threshold_mm))),
    `EXP scars`        = scars_kept_EXP,
    `SDG scars`        = scars_kept_SDG,
    `EXP cum%@l6`      = sprintf("%.2f", exp_cumpower_l6),
    `CV>100% at l`     = exp_cv_cross_l,
    `EXP R2`           = sprintf("%.3f", exp_perm_R2),
    `EXP p`            = sprintf("%.3f", exp_perm_p),
    `EXP sig pairs`    = paste0(exp_perm_nsig, "/", exp_perm_npairs),
    `SDG type R2`      = sprintf("%.3f", sdg_perm_scar_coretype_R2),
    `SDG p`            = sprintf("%.3f", sdg_perm_scar_coretype_p),
    `EXP Mantel r (p)` = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`       = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)` = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`       = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## S3. Mesh decimation and Laplacian smoothing — morphology descriptor

**Methods.** Core morphology is decimated by quadric edge collapse to a fixed target face
count (production = 20000) and Laplacian-smoothed (production = 3 iterations) before the
spherical-harmonic expansion. We recomputed the M-SPHARM power spectra over two
one-dimensional sweeps through the production setting: decimation target
$\in \{10000, 20000, 50000\}$ faces (at production smoothing) and smoothing
$\in \{0, 3, 6\}$ iterations (at the production target), holding everything else fixed (the
quadric-decimation algorithm, mesh normalisation, deterministic area-weighted PCA alignment,
the spherical sampling grid, $l_{\max} = 20$, normalisation, seeds). The scar descriptor is
not derived from these meshes and was reused unchanged. At the production setting the
recomputed spectra reproduce the cached values.

```{r}
#| label: fig-si-mesh-inventory
#| fig-cap: "Raw mesh face counts (log scale) for the IM, SDG and EXP groups against the decimation targets (solid = production 20k; dashed = 10k/50k). EXP and SDG meshes are always decimated; the low-resolution ideal (IM) meshes lie at or below the targets."
knitr::include_graphics(here::here(
  "analysis/robustness/mesh_decimation_sensitivity/figures/fig_S_mesh_inventory.png"))
```

**Determinism.** With the decimation algorithm and parameters fixed, re-running the same
meshes through the pipeline reproduces them exactly (126/126 meshes identical on re-run,
maximum spectral difference $0.0$). The previously reported non-reproducibility of the
morphology results arose from inconsistent, manual decimation; the fixed quadric-decimation
pipeline is deterministic.

**Order selection (truncation at $l = 1$–$8$).** M-SPHARM power concentrates by $l = 8$
(cumulative power 97.98–98.58% EXP, 97.75–98.59% SDG) and the cross-specimen CV stays below
the 90% reliability line through $l = 8$ (maximum CV 82.5–88.6% EXP, 54.6–69.6% SDG). The
$l = 1$–$8$ truncation is robust to decimation and smoothing.

```{r}
#| label: fig-si-mesh-order
#| fig-cap: "M-SPHARM power concentration and cross-specimen CV by degree across mesh-preprocessing settings, for EXP and SDG; dashed line = 90% reliability."
#| fig-width: 9
#| fig-height: 8
mesh$fig_orderselection
```

**Ideal-core discriminability.** The ideal-model meshes are low-resolution
($\le \sim$20k faces), at or below the production decimation target, so the face-count sweep
leaves them essentially unchanged (distance-matrix Pearson $r \approx 1.000$ vs production).
Smoothing has a larger effect on these synthetic meshes: without smoothing the distance
structure is more weakly correlated with the production (smoothed) structure ($r = 0.794$),
reflecting residual faceting artefacts that smoothing removes. The production setting applies
smoothing and the real assemblages are robust (below).

**Archaeological cores (SDG).** Core type is the dominant, significant morphology term across
all settings ($R^2 = 0.178$–$0.216$; $p < 0.05$ throughout), while raw material is small
($R^2 = 0.013$–$0.051$; significant only at the heaviest smoothing) and layer is negligible
($R^2 = 0.007$–$0.030$; non-significant throughout). The "core type dominant, technological
continuity across layers and raw materials" picture holds across all settings.

**Experimental cores.** The experimental morphology PERMANOVA by core type is weak and its
significance is not robust ($R^2 = 0.107$–$0.154$; non-significant at the production setting,
$p = 0.082$; 0 of 10 typology pairs resolved). In the experimental assemblage, type
information therefore resides in the scar-pattern descriptor (which resolves 8/10 pairs; S2)
rather than in morphology — evidence that the two descriptors are complementary rather than
redundant.

**Decoupling (Mantel + RV).** With the perturbed morphology and the fixed scar descriptor,
the association between morphology and scar patterning is negligible in magnitude at every
setting: Mantel $r \approx 0$ (EXP $-0.106$ to $-0.048$; SDG $-0.058$ to $0.102$) and RV
$\approx 0.10$ (EXP $0.095$–$0.116$; SDG $0.072$–$0.093$). The Mantel test is non-significant
throughout (EXP and SDG), and the RV is non-significant at every setting except the unsmoothed
mesh (0 smoothing iterations: EXP RV $= 0.116$, $p = 0.043$) — where, as for the most
over-smoothed bandwidth in S1, the association is still negligible and is not corroborated by
the Mantel test ($r = -0.069$, $p = 0.83$). Read on effect size, the decoupling holds under
perturbation of the morphology descriptor as well as the scar descriptor (S1–S2).

```{r}
#| label: fig-si-mesh-summary
#| fig-cap: "Morphology metrics versus decimation target (left, at production smoothing) and smoothing iterations (right, at production faces): EXP/SDG core-type $R^2$, the ideal-core distance correlation, and EXP/SDG Mantel $r$ and RV."
#| fig-width: 9
#| fig-height: 10
mesh$fig_summary
```

```{r}
#| label: tbl-si-mesh
#| tbl-cap: "Per-setting sensitivity metrics (`*` = production setting, 20000 faces / 3 smoothing iterations). Ideal-model meshes are low-resolution, so the face-count sweep barely changes them. Permutation $p$-values use a fixed seed."
mesh$metrics |>
  arrange(desc(smooth_iters == 3), face_target, smooth_iters) |>
  transmute(
    `faces`               = ifelse(is_production, paste0(face_target, "*"), as.character(face_target)),
    `smooth`              = smooth_iters,
    `EXP cum%@l8`         = sprintf("%.2f", exp_morph_cumpower_l8),
    `EXP maxCV(l1-8)`     = sprintf("%.1f", exp_morph_maxcv_l8),
    `SDG cum%@l8`         = sprintf("%.2f", sdg_morph_cumpower_l8),
    `IM corr vs prod`     = sprintf("%.3f", im_morph_corr_vs_prod),
    `EXP morph R2 (p)`    = pm(exp_morph_perm_R2, exp_morph_perm_p),
    `SDG type R2 (p)`     = pm(sdg_morph_coretype_R2, sdg_morph_coretype_p),
    `SDG rawmat R2 (p)`   = pm(sdg_morph_rawmat_R2, sdg_morph_rawmat_p),
    `SDG layer R2 (p)`    = pm(sdg_morph_layer_R2, sdg_morph_layer_p),
    `EXP Mantel r (p)`    = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`          = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)`    = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`          = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Summary

Across both descriptors, every conclusion the analysis relies on is robust to its principal
free parameters: early truncation ($l = 1$–$6$ for the scar descriptor, $l = 1$–$8$ for
morphology); ideal-core discrimination; the dominance of core type over raw material and layer
in the archaeological (SDG) morphology and scar-pattern signals; and the decoupling of
morphology and scar patterning, which holds under perturbation of either descriptor. The
morphology pipeline is deterministic. The quantities that shift under perturbation are the
*significance* — not the magnitude — of effects that are negligible-to-modest and sit near the
0.05 line: the experimental RV at the most extreme bandwidth (S1) and at the unsmoothed mesh
(S3), the experimental morphology core-type term (S3), and the SDG raw-material term (S3).
This is the empirical reason the main text reads effects on magnitude rather than on threshold
crossings.
