---
title: 'Supplementary Information for "Characterizing core morphology and scar patterning within a unified spherical harmonic framework"'
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      orcid: 0009-0008-1504-9401
      institute:
        - ALPHA
        - MOE
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - MOE: MOE Key Laboratory of Western China's Environmental System, Center for Excellence in Archaeological Science and Cultural Heritage, Lanzhou University, Lanzhou, China
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
    output-file: "ESM_1"
crossref:
  fig-title: "Figure S"
  fig-prefix: "Figure S"
  tbl-title: "Table S"
  tbl-prefix: "Table S"
number-sections: false
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 800
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-method-and-theory.csl"
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false
library(targets)
library(tidyverse)
library(patchwork)

si_store   <- "/project/analysis/robustness/_targets_store"   
main_store <- "/project/analysis/paper/_targets"             
tar_config_set(store = si_store)
bw   <- tar_read(robustness_bandwidth)
scar <- tar_read(robustness_scar_threshold)
mesh <- tar_read(robustness_mesh_decimation)
pm <- function(r, p) sprintf("%.3f (%.3f)", r, p)  
```

# Sensitivity analysis

To evaluate the stability of the spherical harmonic framework, we
conducted a parameter sensitivity robustness analysis. The purpose of
this analysis was to determine whether the main computational results
depend on a single selected parameter setting, or whether they remain
stable across a reasonable range of values for key control parameters.
This is an important step in assessing algorithmic performance, because
the outputs of a computational method may be shaped not only by the
empirical structure of the data, but also by analyst-defined parameters
[@hambyReviewTechniquesParameter1994;
@saltelliSensitivityAnalysisImportance2002].

In computational modelling, sensitivity analysis is commonly used to
examine how changes in model inputs or assumptions affect model outputs
and the resulting inferences
[@saltelliSensitivityAnalysisImportance2002]. In our study, the tested
parameters may influence how the spherical harmonic descriptors
represent core morphology and scar patterning. Therefore, if the main
patterns of variation, group separation, effect size ranking, and
covariation results remain stable when these parameters are varied
within reasonable ranges, the method and the selected main analysis
settings can be considered robust and justified. Conversely, if the
analytical conclusions appear only under a specific parameter setting,
those conclusions would be parameter dependent, which would also limit
the broader generalizability of the analytical method.

We therefore assessed whether the main results depend on the principal
free parameters of the two descriptors. For the scar-pattern descriptor
(SP-SPHARM), we varied the von Mises–Fisher kernel bandwidth ($h$) and
the scar minimum-size inclusion threshold. For the core morphology
descriptor (M-SPHARM), we varied the mesh decimation target and the
number of Laplacian smoothing iterations. In each parameter sweep, all
settings other than the focal parameter were held at the values used in
the main analysis. At the main analysis setting, the recomputed
descriptors reproduced the previously saved descriptor values,
confirming that each sensitivity sweep was anchored to the results
reported in the main analysis. As in the main text, we interpret the
results primarily in terms of effect magnitudes rather than significance
thresholds. The analyses below show repeatedly that parameter
perturbation affects the significance of small-to-modest effects near
the 0.05 threshold, but not the magnitude of the effects themselves.

## Bandwidth of the von Mises–Fisher kernel

The spherical kernel density bandwidth was fixed at $h$ = 0.35
(concentration $\kappa$ = 1/$h^2$ = 8.16) in the main analysis. We
recomputed the SP-SPHARM power spectra over $h$ ∈ {0.20, 0.25, 0.30,
0.35, 0.40, 0.45, 0.50} ($\kappa$ = 4.0–25.0), holding everything else
fixed (alignment, the spherical evaluation grid and Driscoll–Healy
transform, $l_{\max}$ = 20, normalisation, and seeds). Morphology
(M-SPHARM) is independent of $h$ and was reused unchanged. At $h$ = 0.35
the recomputed spectra reproduce the cached values.

Across all $h$, SP-SPHARM power saturates early: cumulative power
through $l$ = 6 is 96.71–99.99%, and the cross specimen coefficient of
variation (CV) first exceeds 100% at degree $l$ = 7–14. The $l$ = 1–6
truncation is robust to $h$; it is most complete at moderate-to-large
$h$ (at the smallest $h$ a little reliable signal persists beyond $l$ =
6).

```{r}
#| label: fig-si-bandwidth-order
#| fig-cap: "SP-SPHARM cumulative power (top) and cross specimen CV (bottom) by spherical harmonic degree $l$, overlaid across bandwidths $h$, for the experimental (EXP) and archaeological (SDG) assemblages. Dashed lines mark the 95%/99% power and 100% CV references."
#| fig-width: 9
#| fig-height: 8
bw$fig_orderselection
```

The pairwise standardised Euclidean distance structure is highly stable:
each $h$'s distance matrix correlates with the $h$ = 0.35 matrix at
Pearson $r$ ≥ 0.961, and key separations persist at every $h$ (e.g.
discoidal vs Levallois).

```{r}
#| label: fig-si-bandwidth-im
#| fig-cap: "Idealized core standardised Euclidean distance matrices at $h$ = 0.20, 0.35 and 0.50."
#| fig-width: 11
#| fig-height: 4.5
bw$fig_im_heatmaps
```

The scar pattern PERMANOVA by experimental core type stays significant
across $h$ ($R^2$ = 0.298–0.329; $p$ \< 0.05 throughout). The pairwise resolution profile is stable: 7 of 10 typology pairs are resolved at the three smallest bandwidths ($h$ = 0.20–0.30) and 8 of 10 from $h$ = 0.35 to 0.50, including the chosen $h$
= 0.35), dropping to 6/10 at the two smallest bandwidths and 7/10 at the
largest. The association between morphology and scar patterning is
negligible in magnitude at every bandwidth: Mantel $r$ ≈ 0 throughout
(EXP -0.062 to -0.008; SDG 0.001 to 0.017) and RV ≈ 0.10–0.12 (EXP) and
≈ 0.09 (SDG). For SDG both measures are non-significant across the
entire range. For the experimental assemblage the Mantel test is
non-significant at every $h$, and the RV coefficient is non-significant
from $h$ = 0.20 to 0.40, becomes marginal at $h$ = 0.45 ($p$ = 0.078),
and crosses $p$ \< 0.05 only at the most over smoothed setting ($h$ =
0.50: RV = 0.117, $p$ = 0.044), where the association is still
negligible and is not corroborated by the Mantel test ($r$ = -0.008, $p$
= 0.54). Read on effect size, the decoupling is robust across the full
range, and the chosen $h$ = 0.35 sits well inside the non-significant
regime. The SDG scar–core type PERMANOVA remains significant throughout
($R^2$ = 0.166–0.177; $p$ \< 0.05).

```{r}
#| label: tbl-si-bandwidth
#| tbl-cap: "Per bandwidth sensitivity metrics. The asterisk marks the main analysis setting ($h$ = 0.35). Reported metrics include cumulative power, coefficient of variation (CV), idealized-model (IM) correlation, $R^2$, and Mantel/RV estimates."
bw$metrics |>
  arrange(h) |>
  transmute(
    `h`                = ifelse(abs(h - 0.35) < 1e-9, paste0(format(h, nsmall = 2), "*"), format(h, nsmall = 2)),
    `κ`           = sprintf("%.2f", kappa),
    `EXP cum% (l=6)`   = sprintf("%.2f", exp_cumpower_l6),
    `CV>100% at l`     = exp_cv_cross_l,
    `IM corr vs .35`   = sprintf("%.3f", im_corr_vs_ref),
    `EXP R2`           = sprintf("%.3f", exp_perm_R2),
    `EXP p`            = sprintf("%.3f", exp_perm_p),
    `EXP sig pairs`    = paste0(exp_perm_nsig, "/", exp_perm_npairs),
    `SDG type R2`      = sprintf("%.3f", sdg_perm_scar_coretype_R2),
    `SDG p`            = sprintf("%.3f", sdg_perm_scar_coretype_p),
    `EXP Mantel r (p)` = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`       = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)` = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`       = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Flaking scar size threshold

To test whether the results depend on flaking scar size, and in
particular whether they are driven by small, less reliably measured
scars, we recomputed the SP-SPHARM power spectra after dropping scars at
or below increasing cutoffs $T$ ∈ {2, 5, 10} mm (retaining length \>
$T$), holding everything else fixed (alignment, $h$ = 0.35, the
evaluation and transform grids, $l_{\max}$ = 20, normalisation, seeds).
The cutoff $T$ = 2 was used as the main analysis, at which the
recomputed descriptors reproduced the saved descriptor values used in
the main analysis. The \>5 mm rule retains 95.9% (EXP) and 99.2% (SDG)
of scars and \>10 mm retains 80.7% and 88.5%. The threshold is not
applied to the idealized cores, whose scar lengths are fixed values, so
their discriminability is unchanged by construction.

```{r}
#| label: fig-si-scar-counts
#| fig-cap: "Per specimen scar retention by size threshold $T$; the dashed line is the three-scar advisory floor, below which no specimen falls."
#| fig-width: 9
#| fig-height: 4.5
scar$fig_scarcounts
```

SP-SPHARM power saturates early at every threshold: cumulative power
through $l$ = 6 is 99.91–99.92%, and the cross-specimen CV first exceeds
100% at degree $l$ = 9 for all cutoff value. The $l$ = 1–6 truncation is
robust to the inclusion rule.

```{r}
#| label: fig-si-scar-order
#| fig-cap: "SP-SPHARM power saturation and CV by degree across scar size thresholds $T$, for experimentally knapped cores and SDG cores."
#| fig-width: 9
#| fig-height: 8
scar$fig_orderselection
```

The scar pattern PERMANOVA by core type is significant at every
threshold ($R^2$ = 0.282–0.322; $p$ \< 0.05 throughout). The pairwise resolution profile is 8/10 typology pairs at the main analysis baseline, 7/10 at \>5 mm, and 8/10 at \>10 mm. Restricting the analysis to larger, more reliably measured scars therefore does not weaken type discrimination; the experimental core-type signal is carried by the larger scars and is not an artifact of small scars.

Morphology and scar patterning remain negligibly associated and
non-significant at every threshold: Mantel $r$ ≈ 0 (EXP -0.061 to 0.004;
SDG -0.047 to 0.012) and RV = 0.101–0.109 (EXP) and 0.057–0.091 (SDG),
all $p$ \> 0.05. The SDG scar–core type PERMANOVA is significant
throughout ($R^2$ = 0.168–0.171).

```{r}
#| label: tbl-si-scar
#| tbl-cap: "Per threshold sensitivity metrics. The asterisk marks the main analysis baseline, in which > 2 mm scars are retained. The EXP and SDG scar columns report the total number of retained scars."
scar$metrics |>
  arrange(threshold_mm) |>
  transmute(
    `T (mm)`           = ifelse(abs(threshold_mm) < 1e-9, "0*", as.character(as.integer(threshold_mm))),
    `EXP scars`        = scars_kept_EXP,
    `SDG scars`        = scars_kept_SDG,
    `EXP cum% (l=6)`   = sprintf("%.2f", exp_cumpower_l6),
    `CV>100% at l`     = exp_cv_cross_l,
    `EXP R2`           = sprintf("%.3f", exp_perm_R2),
    `EXP p`            = sprintf("%.3f", exp_perm_p),
    `EXP sig pairs`    = paste0(exp_perm_nsig, "/", exp_perm_npairs),
    `SDG type R2`      = sprintf("%.3f", sdg_perm_scar_coretype_R2),
    `SDG p`            = sprintf("%.3f", sdg_perm_scar_coretype_p),
    `EXP Mantel r (p)` = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`       = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)` = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`       = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Mesh decimation and Laplacian smoothing

Core morphology is decimated by quadric edge collapse to a fixed target
face count (main analysis = 20,000) and Laplacian-smoothed (main
analysis = 3 iterations) before the spherical harmonic expansion. We
recomputed the M-SPHARM power spectra over two one-dimensional sweeps
through the main analysis setting: decimation target ∈ {10000, 20000,
50000} faces (at production smoothing) and smoothing ∈ {0, 3, 6}
iterations (at the production target), holding everything else fixed
(the quadric-decimation algorithm, mesh normalisation, deterministic
area-weighted PCA alignment, the spherical sampling grid, $l_{\max}$ =
20, normalisation, seeds).

M-SPHARM power concentrates by $l$ = 8 (cumulative power 97.98–98.58%
EXP, 97.75–98.59% SDG) and the cross-specimen CV stays below the 90%
reliability line through $l$ = 8 (maximum CV 82.5–88.6% EXP, 54.6–69.6%
SDG). The $l$ = 1–8 truncation is robust to decimation and smoothing.

```{r}
#| label: fig-si-mesh-order
#| fig-cap: "M-SPHARM power concentration and CV by degree across mesh-preprocessing settings, for EXP and SDG; dashed line = 90% reliability."
#| fig-width: 9
#| fig-height: 8
mesh$fig_orderselection
```

The experimental morphology PERMANOVA by core type is weak but consistently significant ($R^2$ = 0.121–0.149; 0 of 10 typology pairs resolved). In the experimental assemblage, type information is therefore carried far more strongly by the scar pattern descriptor (which resolves 8/10 pairs; S2) than by morphology, evidence that the two descriptors are complementary rather than redundant. Core type is the dominant, significant morphology term
across all settings ($R^2$ = 0.178–0.216; $p$ \< 0.05 throughout), while
raw material is small ($R^2$ = 0.013–0.051; significant only at the
heaviest smoothing) and layer is negligible ($R^2$ = 0.007–0.030;
non-significant throughout). The "core type dominant, technological
continuity across layers and raw materials" picture holds across all
settings.

With the perturbed morphology and the fixed scar descriptor, the
association between morphology and scar patterning is negligible in
magnitude at every setting: Mantel $r$ ≈ 0 (EXP -0.106 to -0.048; SDG
-0.058 to 0.102) and RV ≈ 0.10 (EXP 0.095–0.116; SDG 0.072–0.093). The
Mantel test is non-significant throughout (EXP and SDG), and the RV is
non-significant at every setting except the unsmoothed mesh (0 smoothing
iterations: EXP RV = 0.116, $p$ = 0.043) — where, as for the most over
smoothed bandwidth in S1, the association is still negligible and is not
corroborated by the Mantel test ($r$ = -0.069, $p$ = 0.83). Read on
effect size, the decoupling holds under perturbation of the morphology
descriptor as well as the scar descriptor (S1–S2).

```{r}
#| label: tbl-si-mesh
#| tbl-cap: "Per setting sensitivity metrics. The asterisk marks the main analysis setting, defined as 20,000 faces and three Laplacian smoothing iterations."
mesh$metrics |>
  arrange(desc(smooth_iters == 3), face_target, smooth_iters) |>
  transmute(
    `faces`               = ifelse(is_production, paste0(face_target, "*"), as.character(face_target)),
    `smooth`              = smooth_iters,
    `EXP cum% (l=8)`         = sprintf("%.2f", exp_morph_cumpower_l8),
    `EXP maxCV(l1-8)`     = sprintf("%.1f", exp_morph_maxcv_l8),
    `SDG cum% (l=8)`         = sprintf("%.2f", sdg_morph_cumpower_l8),
    `EXP morph R2 (p)`    = pm(exp_morph_perm_R2, exp_morph_perm_p),
    `SDG type R2 (p)`     = pm(sdg_morph_coretype_R2, sdg_morph_coretype_p),
    `SDG rawmat R2 (p)`   = pm(sdg_morph_rawmat_R2, sdg_morph_rawmat_p),
    `SDG layer R2 (p)`    = pm(sdg_morph_layer_R2, sdg_morph_layer_p),
    `EXP Mantel r (p)`    = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`          = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)`    = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`          = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Summary

Across both descriptors, every conclusion the analysis relies on is
robust to its principal free parameters: early truncation ($l$ = 1–6 for
the scar descriptor, $l$ = 1–8 for morphology); idealized core
discrimination; the dominance of core type over raw material and layer
in the archaeological (SDG) morphology and scar pattern signals; and the
decoupling of morphology and scar patterning, which holds under
perturbation of either descriptor.

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::
