---
title: 'Supplementary Information for "Characterizing core morphology and scar patterning within a unified spherical harmonic framework"'
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      orcid: 0009-0008-1504-9401
      institute:
        - ALPHA
        - MOE
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - MOE: MOE Key Laboratory of Western China's Environmental System, Center for Excellence in Archaeological Science and Cultural Heritage, Lanzhou University, Lanzhou, China
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
    output-file: "ESM_1"
crossref:
  fig-title: "Figure S"
  fig-prefix: "Figure S"
  tbl-title: "Table S"
  tbl-prefix: "Table S"
number-sections: false
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 800
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-method-and-theory.csl"
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false
library(targets)
library(tidyverse)
library(patchwork)

si_store   <- "/project/analysis/robustness/_targets_store"   
main_store <- "/project/analysis/paper/_targets"             
tar_config_set(store = si_store)
bw   <- tar_read(robustness_bandwidth)
scar <- tar_read(robustness_scar_threshold)
mesh <- tar_read(robustness_mesh_decimation)
pwr  <- tar_read(robustness_power)
trn  <- tar_read(robustness_truncation)
ann  <- tar_read(robustness_annotation)
pm <- function(r, p) sprintf("%.3f (%.3f)", r, p)
# inline accessor: a formatted power/sensitivity metric for one assemblage
pv <- function(asm, col, d = 3) sprintf(paste0("%.", d, "f"),
  pwr$metrics[[col]][pwr$metrics$assemblage == asm])

# ---- truncation sweep accessors (scan = "SP" | "M") -------------------------
trow <- function(scan_, asm, L)
  trn$summary[trn$summary$scan == scan_ &
              trn$summary$assemblage == asm &
              trn$summary$varied_lmax == L, ]
# one metric, formatted to d decimals
tv   <- function(scan_, asm, L, col, d = 3)
  sprintf(paste0("%.", d, "f"), trow(scan_, asm, L)[[col]])
# one integer-valued metric
tint <- function(scan_, asm, L, col) as.character(trow(scan_, asm, L)[[col]])
# min-max of a metric across a sweep; sep is en dash for strictly positive
# ranges and " to " where the range spans zero, matching the sections above
trng <- function(scan_, asm, col, d = 3, sep = "–") {
  x <- trn$summary[[col]][trn$summary$scan == scan_ &
                          trn$summary$assemblage == asm]
  paste0(sprintf(paste0("%.", d, "f"), min(x)), sep,
         sprintf(paste0("%.", d, "f"), max(x)))
}
# stable truncation interval for one scan x assemblage, e.g. "1-6 to 1-7"
tstab <- function(scan_, asm) {
  r <- trn$stable[trn$stable$scan == scan_ & trn$stable$assemblage == asm, ]
  sprintf("1–%d to 1–%d", r$lo, r$hi)
}
# full swept span for a scan, and the anchor / held-fixed degrees
tspan   <- function(scan_) {
  x <- trn$summary$varied_lmax[trn$summary$scan == scan_]
  sprintf("1–%d to 1–%d", min(x), max(x))
}
tanchor <- function(scan_) unique(trn$summary$varied_lmax[
  trn$summary$scan == scan_ & trn$summary$is_anchor])
thold   <- function(scan_) {
  d <- trn$summary[trn$summary$scan == scan_, ]
  unique(if (scan_ == "SP") d$lmax_M else d$lmax_SP)
}
# typology pair(s) resolved at the anchor but lost at truncation L; the leading
# type name is lower-cased to read as running prose ("unidirectional-Levallois")
tlost <- function(scan_, asm, L) {
  pre <- if (scan_ == "SP") "SP_" else "M_"
  g <- function(LL) trimws(strsplit(
         trow(scan_, asm, LL)[[paste0(pre, "sig_pairs")]], ";")[[1]])
  lost <- setdiff(g(tanchor(scan_)), g(L))
  paste(sub("^([A-Za-z]+)-", "\\L\\1–", lost, perl = TRUE), collapse = ", ")
}

# ---- annotation-perturbation accessors (kind = "baseline" | "polarity" |
# "angle"; level is the flip fraction or the jitter s.d. in degrees) ----------
arow <- function(kind, lvl)
  ann$summary[ann$summary$perturbation == kind &
              abs(ann$summary$level - lvl) < 1e-9, ]
# one metric from the wide per-level summary, formatted to d decimals
av   <- function(kind, lvl, col, d = 3)
  sprintf(paste0("%.", d, "f"), arow(kind, lvl)[[col]])
# a resolved-pair count: a median over replicates, integral at every level here
aint <- function(kind, lvl, col) sprintf("%.0f", arow(kind, lvl)[[col]])
# the same metric for one of the three methods, from the tidy three-method table
amet <- function(kind, meth, lvl, col, d = 3) {
  x <- ann$three
  sprintf(paste0("%.", d, "f"),
          x[[col]][x$perturbation == kind & x$method == meth &
                   abs(x$level - lvl) < 1e-9])
}
# the perturbed levels only; the unperturbed anchor is quoted separately, as in
# the sweeps above
apert <- function() ann$summary[ann$summary$perturbation != "baseline", ]
# widest excursion of a summary metric across every perturbed level, as a
# magnitude, and its min-max range
aabs <- function(col, d = 3) sprintf(paste0("%.", d, "f"), max(abs(apert()[[col]])))
arng <- function(col, d = 3, sep = "–") {
  x <- apert()[[col]]
  paste0(sprintf(paste0("%.", d, "f"), min(x)), sep,
         sprintf(paste0("%.", d, "f"), max(x)))
}
# the swept levels of one perturbation, as a set for the running text
alev <- function(kind, mult = 1, d = 0)
  paste(sprintf(paste0("%.", d, "f"),
                mult * sort(apert()$level[apert()$perturbation == kind])),
        collapse = ", ")
# total replicates run for one perturbation, across all of its levels
arep <- function(kind) as.character(sum(apert()$R[apert()$perturbation == kind]))
# a flip fraction as a percentage, for the interpolated crossing levels
apct <- function(x, d = 1) sprintf(paste0("%.", d, "f"), 100 * x)
# scars actually reversed at a flip fraction, and how many specimens they fall in
aflip  <- function(f) as.character(ann$scars$flips[[as.character(f)]])
atouch <- function(f) as.character(ann$scars$n_touch[[as.character(f)]])
```

# Sensitivity analysis

To evaluate the stability of the spherical harmonic framework, we
conducted a parameter sensitivity robustness analysis. The purpose of
this analysis was to determine whether the main computational results
depend on a single selected parameter setting, or whether they remain
stable across a reasonable range of values for key control parameters.
This is an important step in assessing algorithmic performance, because
the outputs of a computational method may be shaped not only by the
empirical structure of the data, but also by analyst-defined parameters
[@hambyReviewTechniquesParameter1994;
@saltelliSensitivityAnalysisImportance2002].

In computational modelling, sensitivity analysis is commonly used to
examine how changes in model inputs or assumptions affect model outputs
and the resulting inferences
[@saltelliSensitivityAnalysisImportance2002]. In our study, the tested
parameters may influence how the spherical harmonic descriptors
represent core morphology and scar patterning. Therefore, if the main
patterns of variation, group separation, effect size ranking, and
covariation results remain stable when these parameters are varied
within reasonable ranges, the method and the selected main analysis
settings can be considered robust and justified. Conversely, if the
analytical conclusions appear only under a specific parameter setting,
those conclusions would be parameter dependent, which would also limit
the broader generalizability of the analytical method.

We therefore assessed whether the main results depend on the principal
free parameters of the two descriptors. For the scar pattern descriptor
(SP-SPHARM), we varied the von Mises–Fisher kernel bandwidth ($h$) and
the scar minimum-size inclusion threshold. For the core morphology
descriptor (M-SPHARM), we varied the mesh decimation target and the
number of Laplacian smoothing iterations. We additionally varied the
truncation degree at which the power spectra enter the statistical
comparison. This last parameter differs in kind from the other four: it
does not change how either descriptor is computed, since coefficients
were calculated to the same maximum degree throughout, but only which
degrees are retained for the statistical comparison. In each parameter
sweep, all settings other than the focal parameter were held at the
values used in the main analysis. At the main analysis setting, the recomputed
descriptors reproduced the previously saved descriptor values,
confirming that each sensitivity sweep was anchored to the results
reported in the main analysis. As in the main text, we interpret the
results primarily in terms of effect magnitudes rather than significance
thresholds. The analyses below show repeatedly that parameter
perturbation affects the significance of small-to-modest effects near
the 0.05 threshold, but not the magnitude of the effects themselves.

## Bandwidth of the von Mises–Fisher kernel

The spherical kernel density bandwidth was fixed at $h$ = 0.35
(concentration $\kappa$ = 1/$h^2$ = 8.16) in the main analysis. We
recomputed the SP-SPHARM power spectra over $h$ ∈ {0.20, 0.25, 0.30,
0.35, 0.40, 0.45, 0.50} ($\kappa$ = 4.0–25.0), holding everything else
fixed (alignment, the spherical evaluation grid and Driscoll–Healy
transform, $l_{\max}$ = 20, normalisation, and seeds). Morphology
(M-SPHARM) is independent of $h$ and was reused unchanged. At $h$ = 0.35
the recomputed spectra reproduce the cached values.

Across all $h$, SP-SPHARM power saturates early: cumulative power
through $l$ = 6 is 96.71–99.99%, and the cross specimen coefficient of
variation (CV) first exceeds 100% at degree $l$ = 7–14. The $l$ = 1–6
truncation is robust to $h$; it is most complete at moderate-to-large
$h$ (at the smallest $h$ a little reliable signal persists beyond $l$ =
6).

```{r}
#| label: fig-si-bandwidth-order
#| fig-cap: "SP-SPHARM cumulative power (top) and cross specimen CV (bottom) by spherical harmonic degree $l$, overlaid across bandwidths $h$, for the experimental (EXP) and archaeological (SDG) assemblages. Dashed lines mark the 95%/99% power and 100% CV references."
#| fig-width: 9
#| fig-height: 8
bw$fig_orderselection
```

The pairwise standardised Euclidean distance structure is highly stable:
each $h$'s distance matrix correlates with the $h$ = 0.35 matrix at
Pearson $r$ ≥ 0.961, and key separations persist at every $h$ (e.g.
discoidal vs Levallois).

```{r}
#| label: fig-si-bandwidth-im
#| fig-cap: "Idealized core standardised Euclidean distance matrices at $h$ = 0.20, 0.35 and 0.50."
#| fig-width: 11
#| fig-height: 4.5
bw$fig_im_heatmaps
```

The scar pattern PERMANOVA by experimental core type stays significant
across $h$ ($R^2$ = 0.298–0.329; $p$ \< 0.05 throughout). The pairwise
resolution profile is stable: 7 of 10 typology pairs are resolved at the
three smallest bandwidths ($h$ = 0.20–0.30) and 8 of 10 from $h$ = 0.35
to 0.50, including the chosen $h$ = 0.35), dropping to 6/10 at the two
smallest bandwidths and 7/10 at the largest. The association between
morphology and scar patterning is negligible in magnitude at every
bandwidth: Mantel $r$ ≈ 0 throughout (EXP -0.062 to -0.008; SDG 0.001 to
0.017) and RV ≈ 0.10–0.12 (EXP) and ≈ 0.09 (SDG). For SDG both measures
are non-significant across the entire range. For the experimental
assemblage the Mantel test is non-significant at every $h$, and the RV
coefficient is non-significant from $h$ = 0.20 to 0.40, becomes marginal
at $h$ = 0.45 ($p$ = 0.078), and crosses $p$ \< 0.05 only at the most
over smoothed setting ($h$ = 0.50: RV = 0.117, $p$ = 0.044), where the
association is still negligible and is not corroborated by the Mantel
test ($r$ = -0.008, $p$ = 0.54). Read on effect size, the decoupling is
robust across the full range, and the chosen $h$ = 0.35 sits well inside
the non-significant regime. The SDG scar–core type PERMANOVA remains
significant throughout ($R^2$ = 0.166–0.177; $p$ \< 0.05).

```{r}
#| label: tbl-si-bandwidth
#| tbl-cap: "Per bandwidth sensitivity metrics. The asterisk marks the main analysis setting ($h$ = 0.35). Reported metrics include cumulative power, coefficient of variation (CV), idealized-model (IM) correlation, $R^2$, and Mantel/RV estimates."
bw$metrics |>
  arrange(h) |>
  transmute(
    `h`                = ifelse(abs(h - 0.35) < 1e-9, paste0(format(h, nsmall = 2), "*"), format(h, nsmall = 2)),
    `κ`           = sprintf("%.2f", kappa),
    `EXP cum% (l=6)`   = sprintf("%.2f", exp_cumpower_l6),
    `CV>100% at l`     = exp_cv_cross_l,
    `IM corr vs .35`   = sprintf("%.3f", im_corr_vs_ref),
    `EXP R2`           = sprintf("%.3f", exp_perm_R2),
    `EXP p`            = sprintf("%.3f", exp_perm_p),
    `EXP sig pairs`    = paste0(exp_perm_nsig, "/", exp_perm_npairs),
    `SDG type R2`      = sprintf("%.3f", sdg_perm_scar_coretype_R2),
    `SDG p`            = sprintf("%.3f", sdg_perm_scar_coretype_p),
    `EXP Mantel r (p)` = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`       = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)` = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`       = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Flaking scar size threshold

To test whether the results depend on flaking scar size, and in
particular whether they are driven by small, less reliably measured
scars, we recomputed the SP-SPHARM power spectra after dropping scars at
or below increasing cutoffs $T$ ∈ {2, 5, 10} mm (retaining length \>
$T$), holding everything else fixed (alignment, $h$ = 0.35, the
evaluation and transform grids, $l_{\max}$ = 20, normalisation, seeds).
The cutoff $T$ = 2 was used as the main analysis, at which the
recomputed descriptors reproduced the saved descriptor values used in
the main analysis. The \>5 mm rule retains 95.9% (EXP) and 99.2% (SDG)
of scars and \>10 mm retains 80.7% and 88.5%. The threshold is not
applied to the idealized cores, whose scar lengths are fixed values, so
their discriminability is unchanged by construction.

```{r}
#| label: fig-si-scar-counts
#| fig-cap: "Per specimen scar retention by size threshold $T$; the dashed line is the three-scar advisory floor, below which no specimen falls."
#| fig-width: 9
#| fig-height: 4.5
scar$fig_scarcounts
```

SP-SPHARM power saturates early at every threshold: cumulative power
through $l$ = 6 is 99.91–99.92%, and the cross-specimen CV first exceeds
100% at degree $l$ = 9 for all cutoff value. The $l$ = 1–6 truncation is
robust to the inclusion rule.

```{r}
#| label: fig-si-scar-order
#| fig-cap: "SP-SPHARM power saturation and CV by degree across scar size thresholds $T$, for experimentally knapped cores and SDG cores."
#| fig-width: 9
#| fig-height: 8
scar$fig_orderselection
```

The scar pattern PERMANOVA by core type is significant at every
threshold ($R^2$ = 0.282–0.322; $p$ \< 0.05 throughout). The pairwise
resolution profile is 8/10 typology pairs at the main analysis baseline,
7/10 at \>5 mm, and 8/10 at \>10 mm. Restricting the analysis to larger,
more reliably measured scars therefore does not weaken type
discrimination; the experimental core-type signal is carried by the
larger scars and is not an artifact of small scars.

Morphology and scar patterning remain negligibly associated and
non-significant at every threshold: Mantel $r$ ≈ 0 (EXP -0.061 to 0.004;
SDG -0.047 to 0.012) and RV = 0.101–0.109 (EXP) and 0.057–0.091 (SDG),
all $p$ \> 0.05. The SDG scar–core type PERMANOVA is significant
throughout ($R^2$ = 0.168–0.171).

```{r}
#| label: tbl-si-scar
#| tbl-cap: "Per threshold sensitivity metrics. The asterisk marks the main analysis baseline, in which > 2 mm scars are retained. The EXP and SDG scar columns report the total number of retained scars."
scar$metrics |>
  arrange(threshold_mm) |>
  transmute(
    `T (mm)`           = ifelse(abs(threshold_mm) < 1e-9, "0*", as.character(as.integer(threshold_mm))),
    `EXP scars`        = scars_kept_EXP,
    `SDG scars`        = scars_kept_SDG,
    `EXP cum% (l=6)`   = sprintf("%.2f", exp_cumpower_l6),
    `CV>100% at l`     = exp_cv_cross_l,
    `EXP R2`           = sprintf("%.3f", exp_perm_R2),
    `EXP p`            = sprintf("%.3f", exp_perm_p),
    `EXP sig pairs`    = paste0(exp_perm_nsig, "/", exp_perm_npairs),
    `SDG type R2`      = sprintf("%.3f", sdg_perm_scar_coretype_R2),
    `SDG p`            = sprintf("%.3f", sdg_perm_scar_coretype_p),
    `EXP Mantel r (p)` = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`       = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)` = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`       = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Scar direction annotation error

The parameters swept above are analytical settings; this subsection varies the
quality of the input data instead. The preceding sweep asks which scars enter
the analysis, and this one asks how accurately those scars are recorded: manual
placement of the direction landmarks that define each scar vector is an
analyst-defined operation, and is
equally error-prone. We therefore
perturbed the scar unit vectors and re-ran the entire workflow
per replicate — vMF spherical kernel density estimation ($h$ = 0.35, $\kappa$ =
8.16), interpolation onto the Driscoll–Healy grid, spherical harmonic expansion
to $l_{\max}$ = 20, normalisation, the ILR transform and the statistical
comparison at $l$ = 1–6 — rather than adding noise to the power spectra
directly. Two error types were simulated. Polarity flips reverse a random
fraction $f$ of the scar vectors, interchanging the proximal and distal ends, at
$f$ ∈ {`r alev("polarity", 100)`}%. Angular jitter rotates each vector
isotropically by a small random angle of standard deviation $\sigma$ ∈
{`r alev("angle")`}°. Each level was repeated `r aint("polarity", 0.02, "R")`
times under fixed seeds, and the unperturbed condition was run once as the
anchor; at that setting the recomputed descriptors reproduced the saved
descriptor values used in the main analysis, confirming that this sweep is
anchored to the reported results as the sweeps above are. Polarity is treated as
the primary case because the case for SP-SPHARM over fabric rests on its
retaining the polarity of the scar vectors, which makes polarity at once its
source of information and its most exposed assumption.

Fabric (elongation and isotropy) and the SPI were recomputed from the same
perturbed vectors, replicate by replicate, so that all three methods see exactly
the same perturbed data at every level. Each method keeps the test it is given in
the main analysis — SP-SPHARM and fabric by PERMANOVA with Holm-corrected
pairwise contrasts, the SPI by Kruskal–Wallis with Dunn post-hoc tests — and
nothing was harmonised in order to draw them together. Two outcome measures are
reported for each, because neither substitutes for the other: the count of
resolved typology pairs is bounded by the significance threshold and is
insensitive to changes in margin, while the median $-\log_{10}$ of the
uncorrected pairwise $p$ values is continuous but insensitive to the loss of
individual marginal pairs. The truncation sweep met the same gap from the other
side, where the median statistic moved only slightly between $l$ = 1–5 and $l$ =
1–6 and yet a typology pair was lost (@tbl-si-truncation).

```{r}
#| label: fig-si-perturbation
#| fig-cap: "Degradation of SP-SPHARM, fabric (elongation and isotropy) and the SPI under simulated annotation error, in the experimental assemblage. Columns are the two error types and rows the two outcome measures, the count of resolved typology pairs of 10 and the median $-\\log_{10}$ of the uncorrected pairwise $p$ values. Each method keeps the test used in the main analysis: SP-SPHARM and fabric by PERMANOVA (adonis2) with Holm-corrected pairwise contrasts, the SPI by Kruskal–Wallis with Holm-corrected Dunn post-hoc tests; nothing was harmonised to draw them together. Every level is 100 replicates, lines are medians and bands are 2.5–97.5% percentile intervals. The fabric band has zero width throughout the polarity column because the orientation tensor is unchanged by $u$ → $-u$ [@markAnalysisAxialOrientation1973], so all polarity replicates return bit-identical fabric results; this is an identity, not a plotting fault. No ranking of the three methods is claimed in the angular jitter column: the two outcome measures order them oppositely and every percentile band spans the method's own baseline at every level."
#| out-width: "100%"
knitr::include_graphics(ann$fig_png)
```

SP-SPHARM degrades as the flip fraction rises, from
`r aint("baseline", 0, "n_sig_05_med")` of 10 resolved typology pairs at the
anchor to `r aint("polarity", 0.1, "n_sig_05_med")` at $f$ = 10% and
`r aint("polarity", 0.2, "n_sig_05_med")` at $f$ = 20%, with the continuous
statistic falling from `r av("baseline", 0, "med_neglog10_p_med", 2)` to
`r av("polarity", 0.2, "med_neglog10_p_med", 2)`. The SPI degrades likewise,
from `r amet("polarity", "SPI", 0, "n_sig_05_med", 0)` pairs to
`r amet("polarity", "SPI", 0.2, "n_sig_05_med", 0)`. Fabric does not degrade at
all: all `r arep("polarity")` polarity replicates return bit-identical results,
holding at `r amet("polarity", "Fabric (E, I)", 0.2, "n_sig_05_med", 0)` of 10
pairs and a median $-\log_{10}$ $p$ of
`r amet("polarity", "Fabric (E, I)", 0.2, "med_neglog10_med", 2)` at every level.
This is a mathematical necessity and not a computational or plotting error:
reversing a vector leaves the orientation tensor $\sum u u^{\top}$ unchanged, so
indices built on its eigenvalues cannot respond to polarity at all
[@markAnalysisAxialOrientation1973]. The zero-width percentile band on the fabric
line in the polarity column of @fig-si-perturbation is exactly this identity.

A descriptor can only degrade when information it encodes is destroyed, so
SP-SPHARM's sensitivity to polarity error is itself evidence that it encodes what
fabric discards. This turns the theoretical argument the main text makes from
@markAnalysisAxialOrientation1973 into a measured result. The consequence for
practice is visible where the curves cross: SP-SPHARM falls below fabric between
the 5% and 10% flip levels, the crossing interpolating to $f$ =
`r apct(ann$cross_count)`% on the resolved-pair count and
`r apct(ann$cross_med)`% on the continuous statistic. Fabric is the more reliable
descriptor where polarity annotation is poor and SP-SPHARM the stronger one where
it is good, which states the condition under which the recommendation to combine
the three methods, rather than to replace one of them with another, applies.

Under angular jitter none of the three methods degrades materially at $\sigma$ ≤
20°, and we make no ranking of them there. The two outcome measures order them
oppositely: SP-SPHARM's continuous statistic is unchanged at $\sigma$ = 20°
(`r av("angle", 20, "med_neglog10_p_med", 2)`, against
`r av("baseline", 0, "med_neglog10_p_med", 2)` at the anchor) while its resolved
count falls from `r aint("baseline", 0, "n_sig_05_med")` to
`r aint("angle", 20, "n_sig_05_med")`, whereas fabric gives up only one pair
(`r amet("angle", "Fabric (E, I)", 0, "n_sig_05_med", 0)` to
`r amet("angle", "Fabric (E, I)", 20, "n_sig_05_med", 0)`) but loses far more of
its continuous statistic (`r amet("angle", "Fabric (E, I)", 0, "med_neglog10_med", 2)`
to `r amet("angle", "Fabric (E, I)", 20, "med_neglog10_med", 2)`). Every
percentile band also spans the method's own baseline at every level, so these
differences are not resolvable at the precision of the measurement.

Across both error types and every level, the association between morphology and
scar patterning remains negligible. The median Mantel $r$ stays within |$r$| ≤
`r aabs("decoup_mantel_med")` and the median RV within `r arng("decoup_rv_med")`,
against anchor values of `r av("baseline", 0, "decoup_mantel_med")` and
`r av("baseline", 0, "decoup_rv_med")`. The decoupling result is therefore immune
to annotation error, as it is to the parameter sweeps above (@tbl-si-bandwidth,
@tbl-si-scar, @tbl-si-mesh).

For practice this places quality control where it does the most good. The
proximal–distal judgement is what has to be right: polarity error should be held
below the crossing identified above, and a rate under 5% keeps SP-SPHARM
resolving more pairs than either of the other two methods. Angular precision is a
much weaker requirement, since jitter within 10° changes little. The type of
annotation error therefore matters more than the choice of method. To give that
tolerance an absolute scale, the analysed experimental assemblage carries
`r ann$scars$n_scar` scars over `r ann$scars$n_spec` specimens, a median of
`r ann$scars$median` scars per specimen (range `r ann$scars$min`–`r ann$scars$max`).
Because the flip fraction is applied within each specimen as the rounded product
$f n_i$, $f$ = 2% reverses `r aflip(0.02)` scars in total, falling in
`r atouch(0.02)` of the `r ann$scars$n_spec` specimens — those carrying enough
scars for that product to reach one. Readers can weigh their own proximal–distal
error rate against these counts.

```{r}
#| label: tbl-si-perturbation
#| tbl-cap: "Per level annotation-error sensitivity metrics for all three scar pattern methods. The asterisk marks the unperturbed anchor, which is the main analysis setting and is shared by both sweeps. Each cell gives the median over the 100 replicates of that level with the 2.5–97.5% percentile interval in brackets; pairs are the Holm-corrected typology pairs resolved of 10, and med -log10 p is the median $-\\log_{10}$ of the uncorrected pairwise $p$ values. Each method keeps its main analysis test (SP-SPHARM and fabric, PERMANOVA; SPI, Kruskal–Wallis with Dunn post-hoc). The fabric columns are constant across the polarity sweep by the orientation tensor identity (see text). Mantel $r$ and RV are the morphology and scar patterning association, recomputed against the unperturbed morphology block."
key <- function(pert, lvl) sprintf("%s_%.4f", pert, lvl)
dec <- ann$summary |>
  transmute(
    k        = ifelse(perturbation == "baseline", "anchor", key(perturbation, level)),
    `EXP Mantel r` = sprintf("%.3f [%.3f, %.3f]",
                             decoup_mantel_med, decoup_mantel_lo, decoup_mantel_hi),
    `EXP RV`       = sprintf("%.3f [%.3f, %.3f]",
                             decoup_rv_med, decoup_rv_lo, decoup_rv_hi))
ann$three |>
  filter(!(perturbation == "angle" & level == 0)) |>
  transmute(
    k     = ifelse(level == 0, "anchor", key(perturbation, level)),
    ord   = ifelse(level == 0, 0, ifelse(perturbation == "polarity", 1, 2)),
    level, perturbation,
    method = recode(method, "SP-SPHARM" = "SP", "Fabric (E, I)" = "fabric"),
    pairs = sprintf("%.0f/10 [%.0f, %.0f]", n_sig_05_med, n_sig_05_lo, n_sig_05_hi),
    med   = sprintf("%.2f [%.2f, %.2f]", med_neglog10_med,
                    med_neglog10_lo, med_neglog10_hi)) |>
  pivot_wider(names_from = method, values_from = c(pairs, med),
              names_glue = "{method} {.value}") |>
  left_join(dec, by = "k") |>
  arrange(ord, level) |>
  transmute(
    `sweep` = c("anchor", "polarity (f)", "angle (σ)")[ord + 1],
    `level` = ifelse(ord == 0, "0*",
              ifelse(ord == 1, sprintf("%.0f%%", 100 * level),
                                sprintf("%.0f°", level))),
    `SP pairs`, `SP med -log10 p` = `SP med`,
    `fabric pairs`, `fabric med -log10 p` = `fabric med`,
    `SPI pairs`, `SPI med -log10 p` = `SPI med`,
    `EXP Mantel r`, `EXP RV`) |>
  knitr::kable()
```

{{< pagebreak >}}

## Mesh decimation and Laplacian smoothing

Core morphology is decimated by quadric edge collapse to a fixed target
face count (main analysis = 20,000) and Laplacian-smoothed (main
analysis = 3 iterations) before the spherical harmonic expansion. We
recomputed the M-SPHARM power spectra over two one-dimensional sweeps
through the main analysis setting: decimation target ∈ {10000, 20000,
50000} faces (at production smoothing) and smoothing ∈ {0, 3, 6}
iterations (at the production target), holding everything else fixed
(the quadric-decimation algorithm, mesh normalisation, deterministic
area-weighted PCA alignment, the spherical sampling grid, $l_{\max}$ =
20, normalisation, seeds).

M-SPHARM power concentrates by $l$ = 8 (cumulative power 97.98–98.58%
EXP, 97.75–98.59% SDG) and the cross-specimen CV stays below the 90%
reliability line through $l$ = 8 (maximum CV 82.5–88.6% EXP, 54.6–69.6%
SDG). The $l$ = 1–8 truncation is robust to decimation and smoothing.

```{r}
#| label: fig-si-mesh-order
#| fig-cap: "M-SPHARM power concentration and CV by degree across mesh-preprocessing settings, for EXP and SDG; dashed line = 90% reliability."
#| fig-width: 9
#| fig-height: 8
mesh$fig_orderselection
```

The experimental morphology PERMANOVA by core type is weak but
consistently significant ($R^2$ = 0.121–0.149; 0 of 10 typology pairs
resolved). In the experimental assemblage, type information is therefore
carried far more strongly by the scar pattern descriptor (which resolves
8/10 pairs; @tbl-si-scar) than by morphology, evidence that the two
descriptors are complementary rather than redundant. Core type is the
dominant, significant morphology term across all settings ($R^2$ =
0.178–0.216; $p$ \< 0.05 throughout), while raw material is small ($R^2$
= 0.013–0.051; significant only at the heaviest smoothing) and layer is
negligible ($R^2$ = 0.007–0.030; non-significant throughout). The "core
type dominant, technological continuity across layers and raw materials"
picture holds across all settings.

With the perturbed morphology and the fixed scar descriptor, the
association between morphology and scar patterning is negligible in
magnitude at every setting: Mantel $r$ ≈ 0 (EXP -0.106 to -0.048; SDG
-0.058 to 0.102) and RV ≈ 0.10 (EXP 0.095–0.116; SDG 0.072–0.093). The
Mantel test is non-significant throughout (EXP and SDG), and the RV is
non-significant at every setting except the unsmoothed mesh (0 smoothing
iterations: EXP RV = 0.116, $p$ = 0.043) — where, as for the most over
smoothed bandwidth in @tbl-si-bandwidth, the association is still
negligible and is not corroborated by the Mantel test ($r$ = -0.069, $p$
= 0.83). Read on effect size, the decoupling holds under perturbation of
the morphology descriptor as well as the scar descriptor
(@tbl-si-bandwidth, @tbl-si-scar).

```{r}
#| label: tbl-si-mesh
#| tbl-cap: "Per setting sensitivity metrics. The asterisk marks the main analysis setting, defined as 20,000 faces and three Laplacian smoothing iterations."
mesh$metrics |>
  arrange(desc(smooth_iters == 3), face_target, smooth_iters) |>
  transmute(
    `faces`               = ifelse(is_production, paste0(face_target, "*"), as.character(face_target)),
    `smooth`              = smooth_iters,
    `EXP cum% (l=8)`         = sprintf("%.2f", exp_morph_cumpower_l8),
    `EXP maxCV(l1-8)`     = sprintf("%.1f", exp_morph_maxcv_l8),
    `SDG cum% (l=8)`         = sprintf("%.2f", sdg_morph_cumpower_l8),
    `EXP morph R2 (p)`    = pm(exp_morph_perm_R2, exp_morph_perm_p),
    `SDG type R2 (p)`     = pm(sdg_morph_coretype_R2, sdg_morph_coretype_p),
    `SDG rawmat R2 (p)`   = pm(sdg_morph_rawmat_R2, sdg_morph_rawmat_p),
    `SDG layer R2 (p)`    = pm(sdg_morph_layer_R2, sdg_morph_layer_p),
    `EXP Mantel r (p)`    = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`          = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)`    = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`          = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Truncation degree

The truncation degree differs from the parameters above in that it does not
alter how the descriptors are computed: coefficients were calculated to
$l_{\max}$ = 20 in every case, and this sweep varies only which degrees are
retained for statistical comparison. We recomputed all analyses over SP-SPHARM
truncations $l$ = `r tspan("SP")` (holding M-SPHARM at $l$ = 1–`r thold("SP")`)
and over M-SPHARM truncations $l$ = `r tspan("M")` (holding SP-SPHARM at $l$ =
1–`r thold("M")`), perturbing one descriptor at a time; varying both
simultaneously would confound the two effects and would additionally change the
relative dimensionality of the two blocks. At the main analysis settings the
recomputed statistics reproduced the values reported in the main text.

Because the ILR transform maps $D$ parts onto $D - 1$ coordinates, the retained
dimensionality changes with the truncation degree and the resulting PERMANOVA
models differ in their degrees of freedom. Comparisons across truncation
settings are therefore not nested, and $R^2$ is not directly comparable between
them: lower truncations return mechanically higher $R^2$ through the reduction
in dimensionality alone, which is why the lowest truncations return $R^2$ values
at or above the main analysis setting in both assemblages. We accordingly judge
stability on the Holm-corrected pairwise resolution count, the median
$-\log_{10}$ of the uncorrected pairwise $p$ values, and the Mantel and RV
estimates, and report $R^2$ only for completeness.

For SP-SPHARM, cumulative power is already above 99% at the fifth degree
(`r tv("SP","EXP",5,"SP_cumpower_at_lmax",2)`% EXP,
`r tv("SP","SDG",5,"SP_cumpower_at_lmax",2)`% SDG) and reaches
`r tv("SP","EXP",6,"SP_cumpower_at_lmax",2)`% and
`r tv("SP","SDG",6,"SP_cumpower_at_lmax",2)`% at the sixth, so the power
criterion alone does not discriminate between these two settings; the CV
criterion places the onset of noise-dominated degrees at $l$ ≈
`r tint("SP","EXP",6,"SP_cv_cross_l")`. The two diagnostics therefore delimit an
acceptable interval of $l$ = 5–8 rather than a single value, and the sweep
resolves the choice within it. Truncating at $l$ = 1–5 costs discrimination: the
experimental pairwise resolution falls from `r tint("SP","EXP",6,"SP_nsig05")`
of `r tint("SP","EXP",6,"n_pairs")` to `r tint("SP","EXP",5,"SP_nsig05")` of
`r tint("SP","EXP",5,"n_pairs")` typology pairs, the pair lost being
`r tlost("SP","EXP",5)`, and the SDG core type PERMANOVA rises from $p$ =
`r tv("SP","SDG",6,"SP_p")` to $p$ = `r tv("SP","SDG",5,"SP_p")`, crossing the
conventional threshold. Extending to $l$ = 1–8 reverses the decoupling result in
the experimental assemblage (RV = `r tv("SP","EXP",8,"RV")`, $p$ =
`r tv("SP","EXP",8,"RV_p")`), with $l$ = 1–7 already marginal ($p$ =
`r tv("SP","EXP",7,"RV_p")`); this reversal does not appear in the SDG
assemblage. Given that the CV diagnostic identifies the high degrees as
noise-dominated and that the reversal is confined to one of the two assemblages,
we interpret it as shared computational noise in the high-degree coefficients
rather than as a technological signal, and note that it is precisely why an
upper-bound criterion is required alongside a power criterion. The conclusions
are stable over $l$ = `r tstab("SP","EXP")` in the experimental assemblage and
$l$ = `r tstab("SP","SDG")` in SDG.

For M-SPHARM the descriptor is insensitive to the truncation degree across the
range tested, with the core type effect ($R^2$ = `r trng("M","EXP","M_R2")` EXP;
`r trng("M","SDG","M_R2")` SDG) and the morphology–scar association (Mantel $r$
≈ 0: EXP `r trng("M","EXP","mantel_r", sep = " to ")`; SDG
`r trng("M","SDG","mantel_r", sep = " to ")`. RV ≈ 0.10: EXP
`r trng("M","EXP","RV")`; SDG `r trng("M","SDG","RV")`) essentially unchanged
from $l$ = 1–`r min(trn$summary$varied_lmax[trn$summary$scan == "M"])` to $l$ =
1–`r max(trn$summary$varied_lmax[trn$summary$scan == "M"])`. The conclusions are
stable over $l$ = `r tstab("M","EXP")` in the experimental assemblage and $l$ =
`r tstab("M","SDG")` in SDG.

The main analysis settings, $l$ = 1–`r tanchor("SP")` for SP-SPHARM and $l$ =
1–`r tanchor("M")` for M-SPHARM, both fall within these stable intervals. For
SP-SPHARM, $l$ = 1–`r tanchor("SP")` is the lowest-dimensional setting in the
interval common to both assemblages, consistent with the principle applied to
M-SPHARM, where reliable but very low-energy higher degrees were excluded to
avoid inflating dimensionality without materially improving discrimination.

```{r}
#| label: tbl-si-truncation
#| tbl-cap: "Per truncation sensitivity metrics for both sweeps. The asterisk marks the main analysis setting of each sweep ($l$ = 1–6 for SP-SPHARM, $l$ = 1–8 for M-SPHARM). Because the ILR transform maps the retained degrees onto one fewer coordinate, the number of retained coordinates (n coords) is given: models are not nested across settings and $R^2$ is therefore not directly comparable between rows (see text)."
trn$summary |>
  transmute(
    scan, assemblage, varied_lmax, is_anchor, n_coords, n_pairs,
    R2     = ifelse(scan == "SP", SP_R2, M_R2),
    p      = ifelse(scan == "SP", SP_p, M_p),
    nsig   = ifelse(scan == "SP", SP_nsig05, M_nsig05),
    med    = ifelse(scan == "SP", SP_med_neglog10_p, M_med_neglog10_p),
    mantel_r, mantel_p, RV, RV_p
  ) |>
  (\(d) left_join(filter(d, assemblage == "EXP"),
                  filter(d, assemblage == "SDG"),
                  by = c("scan", "varied_lmax"), suffix = c("_E", "_S")))() |>
  arrange(desc(scan), varied_lmax) |>
  transmute(
    `sweep`            = ifelse(scan == "SP", "SP-SPHARM", "M-SPHARM"),
    `truncation`       = ifelse(is_anchor_E, paste0("1-", varied_lmax, "*"),
                                             paste0("1-", varied_lmax)),
    `n coords`         = n_coords_E,
    `EXP R2`           = sprintf("%.3f", R2_E),
    `EXP p`            = sprintf("%.3f", p_E),
    `EXP sig pairs`    = paste0(nsig_E, "/", n_pairs_E),
    `EXP med -log10 p` = sprintf("%.2f", med_E),
    `SDG type R2`      = sprintf("%.3f", R2_S),
    `SDG p`            = sprintf("%.3f", p_S),
    `EXP Mantel r (p)` = pm(mantel_r_E, mantel_p_E),
    `EXP RV (p)`       = pm(RV_E, RV_p_E),
    `SDG Mantel r (p)` = pm(mantel_r_S, mantel_p_S),
    `SDG RV (p)`       = pm(RV_S, RV_p_S)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Summary

Across both descriptors, every conclusion the analysis relies on is
robust to its principal free parameters: early truncation ($l$ = 1–6 for
the scar descriptor, $l$ = 1–8 for morphology); idealized core
discrimination; the dominance of core type over raw material and layer
in the archaeological morphology and scar patterning; and the decoupling
of morphology and scar patterning, which holds under perturbation of
either descriptor throughout the bandwidth, scar threshold and mesh
sweeps, and across the stable truncation intervals of both descriptors.
The truncation sweep is the one qualification: at its extreme settings
($l$ = 1–8 for the scar descriptor, $l$ = 1–6 for morphology) the
experimental RV coefficient becomes nominally significant
(@tbl-si-truncation), although the association remains negligible in
magnitude and is not corroborated by the Mantel test in either case. The
main analysis settings sit inside the stable interval of both sweeps.

Robustness extends beyond the analytical parameters to the quality of the input
data, but not uniformly. Perturbing the scar direction landmarks themselves
leaves the decoupling result unchanged at every error type and level, and leaves
type discrimination intact under angular jitter to $\sigma$ = 20°. Polarity
annotation error is the one factor tested anywhere in this section that has a
material effect on a reported result, taking SP-SPHARM's pairwise resolution from
`r aint("baseline", 0, "n_sig_05_med")` of 10 typology pairs to
`r aint("polarity", 0.1, "n_sig_05_med")` at a 10% flip rate
(@tbl-si-perturbation). The conclusions are therefore robust to the analytical
parameters and to angular imprecision, and are conditional on proximal–distal
judgements being largely correct.

# Statistical power of the morphology and scar patterning covariation test

The tests of covariation between core morphology and scar patterning
returned null results in both assemblages (main text; @tbl-si-bandwidth,
@tbl-si-scar, @tbl-si-mesh). Because a non-significant result does not
by itself distinguish a genuine absence of association from insufficient
power to detect one, we assessed the informativeness of these nulls
directly. Rather than computing observed power from the realized effect,
which is uninformative [@hoenigAbusePowerPervasive2001], we followed the
recommendation to characterize the range of effect sizes a fixed design
can resolve [@lakensSampleSizeJustification2022]. Three quantities were
computed for each assemblage, reusing the same ILR transformed power
spectra as the main analysis. At the main analysis setting the
recomputed coefficients reproduced the reported raw RV and Mantel
statistics, confirming that the analysis was anchored to the values in
the main text.

We evaluated the null results in three complementary ways. First,
because raw RV coefficients can be upwardly biased when the number of
variables is large relative to sample size
[@adamsEvaluatingModularityMorphometric2016;
@smildeMatrixCorrelationsHighdimensional2009], we recalculated the
association using the bias-corrected modified RV
[@smildeMatrixCorrelationsHighdimensional2009]. Second, we used a
percentile bootstrap over specimens (2,000 resamples) to estimate how
large the association might plausibly be, while keeping each specimen's
morphology and scar pattern coordinates paired. Third, we used Monte
Carlo simulation to estimate how sensitive the Mantel and RV permutation
tests were at the realized sample sizes [@rohlfUseTwoBlockPartial2000;
@manlyRandomizationBootstrapMonte2007]. In these simulations, data were
generated with a known degree of association between the two blocks,
following established practice in morphometric studies of modularity and
integration [@adamsEvaluatingModularityMorphometric2016]. Variation
within each block was drawn from the empirical mean and covariance of
that block, and a shared component was added to create a controlled
level of between-block association. The strength of this component was
calibrated to a target population RV, by which we mean the underlying
between-block association imposed in the simulated population, rather
than the RV estimated from any single finite sample. Because the
simulated association was aligned with the leading axis of each block,
the resulting minimum detectable effects should be understood as
best-case values: a less favorably structured association would require
a larger effect to be detected [@lakensSampleSizeJustification2022].
Simulation settings were kept modest, and all procedures used fixed
seeds.

These checks suggest that the observed decoupling between morphology and
scar patterning is unlikely to be explained by sample size alone. After
bias correction, the modified RV fell from `r pv("EXP","raw_RV")` to
`r pv("EXP","RV2")` in the EXP assemblage and from
`r pv("SDG","raw_RV")` to `r pv("SDG","RV2")` in the SDG assemblage.
This indicates that much of the already small raw association was
attributable to small-sample inflation rather than to a strong
underlying signal. The bootstrap results also placed only modest upper
bounds on the possible association: approximately RV =
`r pv("EXP","RV2_CI_high",2)` for EXP and `r pv("SDG","RV2_CI_high",2)`
for SDG, and Mantel $r$ = `r pv("EXP","mantel_CI_high",2)` for EXP and
`r pv("SDG","mantel_CI_high",2)` for SDG. The simulation results showed
that the RV permutation test had sufficient power to detect a moderate
association at the realized sample sizes. It reached at least 80% power
at a population RV of `r pv("EXP","MDES_RV")` in EXP and
`r pv("SDG","MDES_RV")` in SDG, and exceeded 96% power by a population
RV of about 0.13. The Mantel test was less sensitive, reaching 80% power
only at a population RV of `r pv("EXP","MDES_mantel")` in EXP and
`r pv("SDG","MDES_mantel")` in SDG, consistent with the lower power of
Mantel tests relative to covariance-based statistics. Power curves are
shown in @fig-si-power, and all values are summarized in @tbl-si-power.

Taken together, these results support the interpretation that core
morphology and scar patterning are largely decoupled at the assemblage
level. If a moderate or stronger association had been present, it would
probably have been detected by the RV test. Instead, the bias-corrected
estimates were close to zero in both assemblages. A very weak residual
association, below a population RV of roughly
`r pv("EXP","MDES_RV",2)`–`r pv("SDG","MDES_RV",2)`, cannot be excluded.
The small per-type subsamples also remain underpowered, which is why the
within-type analyses are treated as exploratory in the main text.

```{r}
#| label: tbl-si-power
#| tbl-cap: "Power and sensitivity metrics for the morphology and scar patterning covariation test, by assemblage. Raw RV and Mantel $r$ are the values reported in the main text (anchor). RV2 is the bias-corrected modified RV coefficient. The 95% confidence intervals are percentile bootstrap intervals (2,000 resamples) and are interpreted as upper bounds on the association (see text). MDES is the minimum population effect detectable at 80% power, estimated by Monte Carlo simulation, and is a lower bound. Values in parentheses are permutation p values."
pwr$metrics |>
  transmute(
    Assemblage        = assemblage,
    n                 = n,
    `Raw RV (p)`      = pm(raw_RV, raw_RV_p),
    RV2               = sprintf("%.3f", RV2),
    `RV2 95% CI`      = sprintf("[%.3f, %.3f]", RV2_CI_low, RV2_CI_high),
    `Mantel r (p)`    = pm(mantel_r, mantel_p),
    `Mantel r 95% CI` = sprintf("[%.3f, %.3f]", mantel_CI_low, mantel_CI_high),
    `MDES RV`         = sprintf("%.3f", MDES_RV),
    `MDES Mantel`     = sprintf("%.3f", MDES_mantel)
  ) |>
  knitr::kable()
```

```{r}
#| label: fig-si-power
#| fig-cap: "Monte Carlo power of the RV permutation test and the Mantel test as a function of the population RV coefficient, at the realized sample sizes of the experimental and Sandinggai assemblages. The dashed horizontal line marks 80% power and the dotted vertical line marks the observed raw RV."
#| fig-width: 9
#| fig-height: 4.5
pwr$fig_power
```

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::
