---
title: 'Supplementary Information for "Characterizing core morphology and scar patterning within a unified spherical harmonic framework"'
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      orcid: 0009-0008-1504-9401
      institute:
        - ALPHA
        - MOE
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - MOE: MOE Key Laboratory of Western China's Environmental System, Center for Excellence in Archaeological Science and Cultural Heritage, Lanzhou University, Lanzhou, China
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
    output-file: "ESM_1"
crossref:
  fig-title: "Figure S"
  fig-prefix: "Figure S"
  tbl-title: "Table S"
  tbl-prefix: "Table S"
number-sections: false
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 800
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-method-and-theory.csl"
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false
library(targets)
library(tidyverse)
library(patchwork)

si_store   <- "/project/analysis/robustness/_targets_store"   
main_store <- "/project/analysis/paper/_targets"             
tar_config_set(store = si_store)
bw   <- tar_read(robustness_bandwidth)
scar <- tar_read(robustness_scar_threshold)
mesh <- tar_read(robustness_mesh_decimation)
pwr  <- tar_read(robustness_power)
trn  <- tar_read(robustness_truncation)
ann  <- tar_read(robustness_annotation)
pm <- function(r, p) sprintf("%.3f (%.3f)", r, p)
# inline accessor: a formatted power/sensitivity metric for one assemblage
pv <- function(asm, col, d = 3) sprintf(paste0("%.", d, "f"),
  pwr$metrics[[col]][pwr$metrics$assemblage == asm])

# ---- truncation sweep accessors (scan = "SP" | "M") -------------------------
trow <- function(scan_, asm, L)
  trn$summary[trn$summary$scan == scan_ &
              trn$summary$assemblage == asm &
              trn$summary$varied_lmax == L, ]
# one metric, formatted to d decimals
tv   <- function(scan_, asm, L, col, d = 3)
  sprintf(paste0("%.", d, "f"), trow(scan_, asm, L)[[col]])
# one integer-valued metric
tint <- function(scan_, asm, L, col) as.character(trow(scan_, asm, L)[[col]])
# min-max of a metric across a sweep; sep is en dash for strictly positive
# ranges and " to " where the range spans zero, matching the sections above
trng <- function(scan_, asm, col, d = 3, sep = "–") {
  x <- trn$summary[[col]][trn$summary$scan == scan_ &
                          trn$summary$assemblage == asm]
  paste0(sprintf(paste0("%.", d, "f"), min(x)), sep,
         sprintf(paste0("%.", d, "f"), max(x)))
}
# stable truncation interval for one scan x assemblage, e.g. "1-6 to 1-7"
tstab <- function(scan_, asm) {
  r <- trn$stable[trn$stable$scan == scan_ & trn$stable$assemblage == asm, ]
  sprintf("1–%d to 1–%d", r$lo, r$hi)
}
# full swept span for a scan, and the anchor / held-fixed degrees
tspan   <- function(scan_) {
  x <- trn$summary$varied_lmax[trn$summary$scan == scan_]
  sprintf("1–%d to 1–%d", min(x), max(x))
}
tanchor <- function(scan_) unique(trn$summary$varied_lmax[
  trn$summary$scan == scan_ & trn$summary$is_anchor])
thold   <- function(scan_) {
  d <- trn$summary[trn$summary$scan == scan_, ]
  unique(if (scan_ == "SP") d$lmax_M else d$lmax_SP)
}
# typology pair(s) resolved at the anchor but lost at truncation L; the leading
# type name is lower-cased to read as running prose ("unidirectional-Levallois")
tlost <- function(scan_, asm, L) {
  pre <- if (scan_ == "SP") "SP_" else "M_"
  g <- function(LL) trimws(strsplit(
         trow(scan_, asm, LL)[[paste0(pre, "sig_pairs")]], ";")[[1]])
  lost <- setdiff(g(tanchor(scan_)), g(L))
  paste(sub("^([A-Za-z]+)-", "\\L\\1–", lost, perl = TRUE), collapse = ", ")
}

# ---- annotation-perturbation accessors (kind = "baseline" | "polarity" |
# "angle"; level is the flip fraction or the jitter s.d. in degrees) ----------
arow <- function(kind, lvl)
  ann$summary[ann$summary$perturbation == kind &
              abs(ann$summary$level - lvl) < 1e-9, ]
# one metric from the wide per-level summary, formatted to d decimals
av   <- function(kind, lvl, col, d = 3)
  sprintf(paste0("%.", d, "f"), arow(kind, lvl)[[col]])
# a resolved-pair count: a median over replicates, integral at every level here
aint <- function(kind, lvl, col) sprintf("%.0f", arow(kind, lvl)[[col]])
# the same metric for one of the three methods, from the tidy three-method table
amet <- function(kind, meth, lvl, col, d = 3) {
  x <- ann$three
  sprintf(paste0("%.", d, "f"),
          x[[col]][x$perturbation == kind & x$method == meth &
                   abs(x$level - lvl) < 1e-9])
}
# the perturbed levels only; the unperturbed anchor is quoted separately, as in
# the sweeps above
apert <- function() ann$summary[ann$summary$perturbation != "baseline", ]
# widest excursion of a summary metric across every perturbed level, as a
# magnitude, and its min-max range
aabs <- function(col, d = 3) sprintf(paste0("%.", d, "f"), max(abs(apert()[[col]])))
arng <- function(col, d = 3, sep = "–") {
  x <- apert()[[col]]
  paste0(sprintf(paste0("%.", d, "f"), min(x)), sep,
         sprintf(paste0("%.", d, "f"), max(x)))
}
# the swept levels of one perturbation, as a set for the running text
alev <- function(kind, mult = 1, d = 0)
  paste(sprintf(paste0("%.", d, "f"),
                mult * sort(apert()$level[apert()$perturbation == kind])),
        collapse = ", ")
# total replicates run for one perturbation, across all of its levels
arep <- function(kind) as.character(sum(apert()$R[apert()$perturbation == kind]))
# a flip fraction as a percentage, for the interpolated crossing levels
apct <- function(x, d = 1) sprintf(paste0("%.", d, "f"), 100 * x)
# scars actually reversed at a flip fraction, and how many specimens they fall in
aflip  <- function(f) as.character(ann$scars$flips[[as.character(f)]])
atouch <- function(f) as.character(ann$scars$n_touch[[as.character(f)]])
```

# Sensitivity analysis

To evaluate the stability of the spherical harmonic framework, we
conducted a parameter sensitivity robustness analysis. The purpose of
this analysis was to determine whether the main computational results
depend on a single selected parameter setting, or whether they remain
stable across a reasonable range of values for key control parameters.
This is an important step in assessing algorithmic performance, because
the outputs of a computational method may be shaped not only by the
empirical structure of the data, but also by analyst-defined parameters
[@hambyReviewTechniquesParameter1994;
@saltelliSensitivityAnalysisImportance2002].

In computational modelling, sensitivity analysis is commonly used to
examine how changes in model inputs or assumptions affect model outputs
and the resulting inferences
[@saltelliSensitivityAnalysisImportance2002]. In our study, the tested
parameters may influence how the spherical harmonic descriptors
represent core morphology and scar patterning. Therefore, if the main
patterns of variation, group separation, effect size ranking, and
covariation results remain stable when these parameters are varied
within reasonable ranges, the method and the selected main analysis
settings can be considered robust and justified. Conversely, if the
analytical conclusions appear only under a specific parameter setting,
those conclusions would be parameter dependent, which would also limit
the broader generalizability of the analytical method.

We therefore assessed whether the main results depend on the principal
free parameters of the two descriptors. For the scar pattern descriptor
(SP-SPHARM), we varied the von Mises–Fisher kernel bandwidth ($h$) and
the scar minimum-size inclusion threshold. For the core morphology
descriptor (M-SPHARM), we varied the mesh decimation target and the
number of Laplacian smoothing iterations. We additionally varied the
truncation degree at which the power spectra enter the statistical
comparison. This last parameter differs in kind from the other four: it
does not change how either descriptor is computed, since coefficients
were calculated to the same maximum degree throughout, but only which
degrees are retained for the statistical comparison. In each parameter
sweep, all settings other than the focal parameter were held at the
values used in the main analysis. At the main analysis setting, the
recomputed descriptors reproduced the previously saved descriptor
values, confirming that each sensitivity sweep was anchored to the
results reported in the main analysis. As in the main text, we interpret
the results primarily in terms of effect magnitudes rather than
significance thresholds. The analyses below show repeatedly that
parameter perturbation affects the significance of small-to-modest
effects near the 0.05 threshold, but not the magnitude of the effects
themselves.

## Bandwidth of the von Mises–Fisher kernel

The spherical kernel density bandwidth was fixed at $h$ = 0.35
(concentration $\kappa$ = 1/$h^2$ = 8.16) in the main analysis. We
recomputed the SP-SPHARM power spectra over $h$ ∈ {0.20, 0.25, 0.30,
0.35, 0.40, 0.45, 0.50} ($\kappa$ = 4.0–25.0), holding everything else
fixed (alignment, the spherical evaluation grid and Driscoll–Healy
transform, $l_{\max}$ = 20, normalisation, and seeds). Morphology
(M-SPHARM) is independent of $h$ and was reused unchanged. At $h$ = 0.35
the recomputed spectra reproduce the cached values.

Across all $h$, SP-SPHARM power saturates early: cumulative power
through $l$ = 6 is 96.71–99.99%, and the cross specimen coefficient of
variation (CV) first exceeds 100% at degree $l$ = 7–14. The $l$ = 1–6
truncation is robust to $h$; it is most complete at moderate-to-large
$h$ (at the smallest $h$ a little reliable signal persists beyond $l$ =
6).

```{r}
#| label: fig-si-bandwidth-order
#| fig-cap: "SP-SPHARM cumulative power (top) and cross specimen CV (bottom) by spherical harmonic degree $l$, overlaid across bandwidths $h$, for the experimental (EXP) and archaeological (SDG) assemblages. Dashed lines mark the 95%/99% power and 100% CV references"
#| fig-width: 9
#| fig-height: 8
bw$fig_orderselection
```

The pairwise standardised Euclidean distance structure is highly stable:
each $h$'s distance matrix correlates with the $h$ = 0.35 matrix at
Pearson $r$ ≥ 0.961, and key separations persist at every $h$ (e.g.
discoidal vs Levallois).

```{r}
#| label: fig-si-bandwidth-im
#| fig-cap: "Idealized core standardised Euclidean distance matrices at $h$ = 0.20, 0.35 and 0.50"
#| fig-width: 11
#| fig-height: 4.5
bw$fig_im_heatmaps
```

The scar pattern PERMANOVA by experimental core type stays significant
across $h$ ($R^2$ = 0.298–0.329; $p$ \< 0.05 throughout). The pairwise
resolution profile is stable: 7 of 10 typology pairs are resolved at the
three smallest bandwidths ($h$ = 0.20–0.30) and 8 of 10 from $h$ = 0.35
to 0.50, including the chosen $h$ = 0.35), dropping to 6/10 at the two
smallest bandwidths and 7/10 at the largest. The association between
morphology and scar patterning is negligible in magnitude at every
bandwidth: Mantel $r$ ≈ 0 throughout (EXP -0.062 to -0.008; SDG 0.001 to
0.017) and RV ≈ 0.10–0.12 (EXP) and ≈ 0.09 (SDG). For SDG both measures
are non-significant across the entire range. For the experimental
assemblage the Mantel test is non-significant at every $h$, and the RV
coefficient is non-significant from $h$ = 0.20 to 0.40, becomes marginal
at $h$ = 0.45 ($p$ = 0.078), and crosses $p$ \< 0.05 only at the most
over smoothed setting ($h$ = 0.50: RV = 0.117, $p$ = 0.044), where the
association is still negligible and is not corroborated by the Mantel
test ($r$ = -0.008, $p$ = 0.54). Read on effect size, the decoupling is
robust across the full range, and the chosen $h$ = 0.35 sits well inside
the non-significant regime. The SDG scar–core type PERMANOVA remains
significant throughout ($R^2$ = 0.166–0.177; $p$ \< 0.05).

```{r}
#| label: tbl-si-bandwidth
#| tbl-cap: "Per bandwidth sensitivity metrics. The asterisk marks the main analysis setting ($h$ = 0.35). Reported metrics include cumulative power, coefficient of variation (CV), idealized-model (IM) correlation, $R^2$, and Mantel/RV estimates"
bw$metrics |>
  arrange(h) |>
  transmute(
    `h`                = ifelse(abs(h - 0.35) < 1e-9, paste0(format(h, nsmall = 2), "*"), format(h, nsmall = 2)),
    `κ`           = sprintf("%.2f", kappa),
    `EXP cum% (l=6)`   = sprintf("%.2f", exp_cumpower_l6),
    `CV>100% at l`     = exp_cv_cross_l,
    `IM corr vs .35`   = sprintf("%.3f", im_corr_vs_ref),
    `EXP R2`           = sprintf("%.3f", exp_perm_R2),
    `EXP p`            = sprintf("%.3f", exp_perm_p),
    `EXP sig pairs`    = paste0(exp_perm_nsig, "/", exp_perm_npairs),
    `SDG type R2`      = sprintf("%.3f", sdg_perm_scar_coretype_R2),
    `SDG p`            = sprintf("%.3f", sdg_perm_scar_coretype_p),
    `EXP Mantel r (p)` = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`       = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)` = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`       = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Flaking scar size threshold

To test whether the results depend on flaking scar size, and in
particular whether they are driven by small, less reliably measured
scars, we recomputed the SP-SPHARM power spectra after dropping scars at
or below increasing cutoffs $T$ ∈ {2, 5, 10} mm (retaining length \>
$T$), holding everything else fixed (alignment, $h$ = 0.35, the
evaluation and transform grids, $l_{\max}$ = 20, normalisation, seeds).
The cutoff $T$ = 2 was used as the main analysis, at which the
recomputed descriptors reproduced the saved descriptor values used in
the main analysis. The \>5 mm rule retains 95.9% (EXP) and 99.2% (SDG)
of scars and \>10 mm retains 80.7% and 88.5%. The threshold is not
applied to the idealized cores, whose scar lengths are fixed values, so
their discriminability is unchanged by construction.

```{r}
#| label: fig-si-scar-counts
#| fig-cap: "Per specimen scar retention by size threshold $T$; the dashed line is the three-scar advisory floor, below which no specimen falls"
#| fig-width: 9
#| fig-height: 4.5
scar$fig_scarcounts
```

SP-SPHARM power saturates early at every threshold: cumulative power
through $l$ = 6 is 99.91–99.92%, and the cross-specimen CV first exceeds
100% at degree $l$ = 9 for all cutoff value. The $l$ = 1–6 truncation is
robust to the inclusion rule.

```{r}
#| label: fig-si-scar-order
#| fig-cap: "SP-SPHARM power saturation and CV by degree across scar size thresholds $T$, for experimentally knapped cores and SDG cores"
#| fig-width: 9
#| fig-height: 8
scar$fig_orderselection
```

The scar pattern PERMANOVA by core type is significant at every
threshold ($R^2$ = 0.282–0.322; $p$ \< 0.05 throughout). The pairwise
resolution profile is 8/10 typology pairs at the main analysis baseline,
7/10 at \>5 mm, and 8/10 at \>10 mm. Restricting the analysis to larger,
more reliably measured scars therefore does not weaken type
discrimination; the experimental core-type signal is carried by the
larger scars and is not an artifact of small scars.

Morphology and scar patterning remain negligibly associated and
non-significant at every threshold: Mantel $r$ ≈ 0 (EXP -0.061 to 0.004;
SDG -0.047 to 0.012) and RV = 0.101–0.109 (EXP) and 0.057–0.091 (SDG),
all $p$ \> 0.05. The SDG scar–core type PERMANOVA is significant
throughout ($R^2$ = 0.168–0.171).

```{r}
#| label: tbl-si-scar
#| tbl-cap: "Per threshold sensitivity metrics. The asterisk marks the main analysis baseline, in which > 2 mm scars are retained. The EXP and SDG scar columns report the total number of retained scars"
scar$metrics |>
  arrange(threshold_mm) |>
  transmute(
    `T (mm)`           = ifelse(abs(threshold_mm) < 1e-9, "0*", as.character(as.integer(threshold_mm))),
    `EXP scars`        = scars_kept_EXP,
    `SDG scars`        = scars_kept_SDG,
    `EXP cum% (l=6)`   = sprintf("%.2f", exp_cumpower_l6),
    `CV>100% at l`     = exp_cv_cross_l,
    `EXP R2`           = sprintf("%.3f", exp_perm_R2),
    `EXP p`            = sprintf("%.3f", exp_perm_p),
    `EXP sig pairs`    = paste0(exp_perm_nsig, "/", exp_perm_npairs),
    `SDG type R2`      = sprintf("%.3f", sdg_perm_scar_coretype_R2),
    `SDG p`            = sprintf("%.3f", sdg_perm_scar_coretype_p),
    `EXP Mantel r (p)` = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`       = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)` = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`       = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Scar direction annotation error

The sensitivity analyses above examine changes in analytical parameters;
this subsection instead examines the quality of the input data. The
preceding analysis asks which scars are included, whereas this one asks
how accurately their directions are recorded. The direction vector of
each scar is defined by manually placing proximal and distal landmarks,
making this step subject to analyst error. To assess its effect, we
directly perturbed the scar unit vectors and reran the complete workflow
for each replicate, including vMF spherical kernel density estimation
($h$ = 0.35, $\kappa$ = 8.16), interpolation onto the Driscoll–Healy
grid, spherical harmonic expansion to $l_{\max}$ = 20, normalisation,
the ILR transform, and statistical comparison at $l$ = 1–6. Thus,
annotation error was simulated in the original scar-direction data
rather than by adding noise directly to the final power spectra.

We simulated two types of direction-annotation error. The first was a
polarity flip, in which a random fraction $f$ of scar directions was
reversed, exchanging the proximal and distal ends, with $f$ ∈
{`r alev("polarity", 100)`}%. The second was angular jitter, in which
each scar direction was rotated by a small random angle with standard
deviation $\sigma$ ∈ {`r alev("angle")`}°. Each level was repeated
`r aint("polarity", 0.02, "R")` times under fixed random seeds, while
the unperturbed condition was run once as the baseline. The descriptors
recomputed under this baseline condition were identical to those saved
from the main analysis, confirming that this sensitivity analysis
follows the same workflow and results as the main analysis.

We focus on polarity error because a key difference between SP-SPHARM
and fabric is whether scar-direction polarity is retained. SP-SPHARM
preserves the proximal-to-distal direction of each scar, whereas fabric
considers the orientation axis without distinguishing its two ends.
Polarity is therefore an important source of information for SP-SPHARM,
but also a potential source of sensitivity to annotation error.

To ensure a fair comparison, fabric (elongation and isotropy) and the
SPI were recomputed from the same perturbed scar directions for every
replicate. Each method retained the statistical test used in the main
analysis: SP-SPHARM and fabric were evaluated using PERMANOVA with
Holm-corrected pairwise contrasts, while the SPI was evaluated using
Kruskal–Wallis tests with Dunn post-hoc comparisons. We report two
outcome measures: the number of the 10 possible typology pairs that
remained significantly distinguishable, and the median $-\log_{10}(p)$
across all typology pairs. The first indicates how many typological
distinctions remain detectable, whereas the second captures changes in
overall separation. Using both measures avoids relying solely on whether
individual comparisons cross the significance threshold. The truncation
sensitivity analysis illustrates the same point: although the median
statistic changed little between $l$ = 1–5 and $l$ = 1–6, one typology
pair was nevertheless lost (@tbl-si-truncation).

```{r}
#| label: fig-si-perturbation
#| fig-cap: "Degradation of SP-SPHARM, fabric (elongation and isotropy) and the SPI under simulated annotation error, in the experimental assemblage. Columns are the two error types and rows the two outcome measures, the count of resolved typology pairs of 10 and the median $-\\log_{10}$ of the uncorrected pairwise $p$ values. Each method keeps the test used in the main analysis: SP-SPHARM and fabric by PERMANOVA (adonis2) with Holm-corrected pairwise contrasts, the SPI by Kruskal–Wallis with Holm-corrected Dunn post-hoc tests; nothing was harmonised to draw them together. Every level is 100 replicates, lines are medians and bands are 2.5–97.5% percentile intervals"
#| out-width: "100%"
knitr::include_graphics(ann$fig_png)
```

As the proportion of incorrectly reversed scar directions increased,
SP-SPHARM progressively lost discriminatory power. Under the baseline
condition, `r aint("baseline", 0, "n_sig_05_med")` of the 10 typology
pairs were distinguishable; this decreased to
`r aint("polarity", 0.1, "n_sig_05_med")` at a flip fraction of 10% and
to `r aint("polarity", 0.2, "n_sig_05_med")` at 20%. The continuous
measure also declined from
`r av("baseline", 0, "med_neglog10_p_med", 2)` to
`r av("polarity", 0.2, "med_neglog10_p_med", 2)`. The SPI showed a
similar decline, from `r amet("polarity", "SPI", 0, "n_sig_05_med", 0)`
typology pairs.

In contrast, fabric was completely unaffected by polarity flips. All
`r arep("polarity")` polarity replicates returned bit-identical results,
maintaining
`r amet("polarity", "Fabric (E, I)", 0.2, "n_sig_05_med", 0)` of 10
typology pairs and a median $-\log_{10}(p)$ of
`r amet("polarity", "Fabric (E, I)", 0.2, "med_neglog10_med", 2)` at
every level. This is a mathematical property of the fabric approach
rather than a computational or plotting artefact: fabric considers the
orientation axis but does not distinguish between the proximal-to-distal
and distal-to-proximal directions. Thus, $\mathbf{u}$ and $-\mathbf{u}$
are equivalent in the fabric calculation, because
$\mathbf{u}\mathbf{u}^{\top}=(-\mathbf{u})(-\mathbf{u})^{\top}$
[@markAnalysisAxialOrientation1973]. The zero-width percentile band for
fabric under polarity error in @fig-si-perturbation is a direct
consequence of this property.

The sensitivity of SP-SPHARM to polarity error also demonstrates that it
retains proximal–distal information that fabric discards. The
performance of the two methods crossed between 5% and 10% polarity
flips: the estimated crossing point was $f$ = `r apct(ann$cross_count)`%
based on the number of resolved typology pairs and
`r apct(ann$cross_med)`% based on the continuous measure. Thus, when
proximal–distal directions are reliably annotated, SP-SPHARM can provide
greater typological discrimination; when polarity errors become more
frequent, fabric is more stable. This result supports combining the
three approaches rather than treating any one of them as a complete
replacement for the others.

Angular jitter had a different effect. None of the three methods showed
substantial overall degradation at $\sigma$ ≤ 20°, and we therefore do
not rank them within this range. The two outcome measures also showed
somewhat different patterns. At $\sigma$ = 20°, the continuous measure
for SP-SPHARM remained largely unchanged
(`r av("angle", 20, "med_neglog10_p_med", 2)`, compared with
`r av("baseline", 0, "med_neglog10_p_med", 2)` at baseline), whereas the
number of resolved typology pairs decreased from
`r aint("baseline", 0, "n_sig_05_med")` to
`r aint("angle", 20, "n_sig_05_med")`. Fabric lost only one typology
pair (from `r amet("angle", "Fabric (E, I)", 0, "n_sig_05_med", 0)` to
`r amet("angle", "Fabric (E, I)", 20, "n_sig_05_med", 0)`), but showed a
larger decline in the continuous measure (from
`r amet("angle", "Fabric (E, I)", 0, "med_neglog10_med", 2)` to
`r amet("angle", "Fabric (E, I)", 20, "med_neglog10_med", 2)`).
Moreover, the percentile intervals for all methods overlapped their
respective baselines at every level, so these differences do not support
a clear ranking of the methods. Overall, small angular errors do not
substantially affect the ability of any of the three methods to
distinguish core types.

Across both polarity flips and angular jitter, the decoupling between
core morphology and scar patterning remained stable. The median Mantel
$r$ remained within \|$r$\| ≤ `r aabs("decoup_mantel_med")`, while the
median RV remained within `r arng("decoup_rv_med")`; the corresponding
baseline values were `r av("baseline", 0, "decoup_mantel_med")` and
`r av("baseline", 0, "decoup_rv_med")`. Thus, the observed decoupling
between morphology and scar patterning is robust to direction-annotation
error, consistent with the sensitivity analyses of bandwidth, scar
selection, and mesh resolution reported above (@tbl-si-bandwidth,
@tbl-si-scar, @tbl-si-mesh).

These results suggest that, when recording scar directions, correctly
identifying the proximal and distal ends is more important than
achieving very high angular precision. Keeping polarity errors below the
crossing point identified above, and particularly below 5%, allows
SP-SPHARM to retain strong typological discrimination. By contrast,
angular errors of around 10° have relatively little effect on the
overall results. Thus, the type of annotation error may matter more than
the magnitude of the error itself.

To place this tolerance in the context of the archaeological dataset,
the experimental assemblage analysed here contains `r ann$scars$n_scar`
scars from `r ann$scars$n_spec` specimens, with a median of
`r ann$scars$median` scars per specimen (range =
`r ann$scars$min`–`r ann$scars$max`). Because the flip fraction is
applied within each specimen using the rounded product $fn_i$, a flip
fraction of $f$ = 2% reverses `r aflip(0.02)` scars in total, affecting
`r atouch(0.02)` of the `r ann$scars$n_spec` specimens. These counts
provide a practical reference against which readers can assess the
expected proximal–distal annotation error rate in their own datasets.

```{r}
#| label: tbl-si-perturbation
#| tbl-cap: "Per level annotation-error sensitivity metrics for all three scar pattern methods. The asterisk marks the unperturbed anchor, which is the main analysis setting and is shared by both sweeps. Each cell gives the median over the 100 replicates of that level with the 2.5–97.5% percentile interval in brackets; pairs are the Holm-corrected typology pairs resolved of 10, and med -log10 p is the median $-\\log_{10}$ of the uncorrected pairwise $p$ values. Each method keeps its main analysis test (SP-SPHARM and fabric, PERMANOVA; SPI, Kruskal–Wallis with Dunn post-hoc)"
key <- function(pert, lvl) sprintf("%s_%.4f", pert, lvl)
dec <- ann$summary |>
  transmute(
    k        = ifelse(perturbation == "baseline", "anchor", key(perturbation, level)),
    `EXP Mantel r` = sprintf("%.3f [%.3f, %.3f]",
                             decoup_mantel_med, decoup_mantel_lo, decoup_mantel_hi),
    `EXP RV`       = sprintf("%.3f [%.3f, %.3f]",
                             decoup_rv_med, decoup_rv_lo, decoup_rv_hi))
ann$three |>
  filter(!(perturbation == "angle" & level == 0)) |>
  transmute(
    k     = ifelse(level == 0, "anchor", key(perturbation, level)),
    ord   = ifelse(level == 0, 0, ifelse(perturbation == "polarity", 1, 2)),
    level, perturbation,
    method = recode(method, "SP-SPHARM" = "SP", "Fabric (E, I)" = "fabric"),
    pairs = sprintf("%.0f/10 [%.0f, %.0f]", n_sig_05_med, n_sig_05_lo, n_sig_05_hi),
    med   = sprintf("%.2f [%.2f, %.2f]", med_neglog10_med,
                    med_neglog10_lo, med_neglog10_hi)) |>
  pivot_wider(names_from = method, values_from = c(pairs, med),
              names_glue = "{method} {.value}") |>
  left_join(dec, by = "k") |>
  arrange(ord, level) |>
  transmute(
    `sweep` = c("anchor", "polarity (f)", "angle (σ)")[ord + 1],
    `level` = ifelse(ord == 0, "0*",
              ifelse(ord == 1, sprintf("%.0f%%", 100 * level),
                                sprintf("%.0f°", level))),
    `SP pairs`, `SP med -log10 p` = `SP med`,
    `fabric pairs`, `fabric med -log10 p` = `fabric med`,
    `SPI pairs`, `SPI med -log10 p` = `SPI med`,
    `EXP Mantel r`, `EXP RV`) |>
  knitr::kable()
```

{{< pagebreak >}}

## Mesh decimation and Laplacian smoothing

Core morphology is decimated by quadric edge collapse to a fixed target
face count (main analysis = 20,000) and Laplacian-smoothed (main
analysis = 3 iterations) before the spherical harmonic expansion. We
recomputed the M-SPHARM power spectra over two one-dimensional sweeps
through the main analysis setting: decimation target ∈ {10000, 20000,
50000} faces (at production smoothing) and smoothing ∈ {0, 3, 6}
iterations (at the production target), holding everything else fixed
(the quadric-decimation algorithm, mesh normalisation, deterministic
area-weighted PCA alignment, the spherical sampling grid, $l_{\max}$ =
20, normalisation, seeds).

M-SPHARM power concentrates by $l$ = 8 (cumulative power 97.98–98.58%
EXP, 97.75–98.59% SDG) and the cross-specimen CV stays below the 90%
reliability line through $l$ = 8 (maximum CV 82.5–88.6% EXP, 54.6–69.6%
SDG). The $l$ = 1–8 truncation is robust to decimation and smoothing.

```{r}
#| label: fig-si-mesh-order
#| fig-cap: "M-SPHARM power concentration and CV by degree across mesh-preprocessing settings, for EXP and SDG; dashed line = 90% reliability"
#| fig-width: 9
#| fig-height: 8
mesh$fig_orderselection
```

The experimental morphology PERMANOVA by core type is weak but
consistently significant ($R^2$ = 0.121–0.149; 0 of 10 typology pairs
resolved). In the experimental assemblage, type information is therefore
carried far more strongly by the scar pattern descriptor (which resolves
8/10 pairs; @tbl-si-scar) than by morphology, evidence that the two
descriptors are complementary rather than redundant. Core type is the
dominant, significant morphology term across all settings ($R^2$ =
0.178–0.216; $p$ \< 0.05 throughout), while raw material is small ($R^2$
= 0.013–0.051; significant only at the heaviest smoothing) and layer is
negligible ($R^2$ = 0.007–0.030; non-significant throughout). The "core
type dominant, technological continuity across layers and raw materials"
picture holds across all settings.

With the perturbed morphology and the fixed scar descriptor, the
association between morphology and scar patterning is negligible in
magnitude at every setting: Mantel $r$ ≈ 0 (EXP -0.106 to -0.048; SDG
-0.058 to 0.102) and RV ≈ 0.10 (EXP 0.095–0.116; SDG 0.072–0.093). The
Mantel test is non-significant throughout (EXP and SDG), and the RV is
non-significant at every setting except the unsmoothed mesh (0 smoothing
iterations: EXP RV = 0.116, $p$ = 0.043) — where, as for the most over
smoothed bandwidth in @tbl-si-bandwidth, the association is still
negligible and is not corroborated by the Mantel test ($r$ = -0.069, $p$
= 0.83). Read on effect size, the decoupling holds under perturbation of
the morphology descriptor as well as the scar descriptor
(@tbl-si-bandwidth, @tbl-si-scar).

```{r}
#| label: tbl-si-mesh
#| tbl-cap: "Per setting sensitivity metrics. The asterisk marks the main analysis setting, defined as 20,000 faces and three Laplacian smoothing iterations"
mesh$metrics |>
  arrange(desc(smooth_iters == 3), face_target, smooth_iters) |>
  transmute(
    `faces`               = ifelse(is_production, paste0(face_target, "*"), as.character(face_target)),
    `smooth`              = smooth_iters,
    `EXP cum% (l=8)`         = sprintf("%.2f", exp_morph_cumpower_l8),
    `EXP maxCV(l1-8)`     = sprintf("%.1f", exp_morph_maxcv_l8),
    `SDG cum% (l=8)`         = sprintf("%.2f", sdg_morph_cumpower_l8),
    `EXP morph R2 (p)`    = pm(exp_morph_perm_R2, exp_morph_perm_p),
    `SDG type R2 (p)`     = pm(sdg_morph_coretype_R2, sdg_morph_coretype_p),
    `SDG rawmat R2 (p)`   = pm(sdg_morph_rawmat_R2, sdg_morph_rawmat_p),
    `SDG layer R2 (p)`    = pm(sdg_morph_layer_R2, sdg_morph_layer_p),
    `EXP Mantel r (p)`    = pm(exp_mantel_r, exp_mantel_p),
    `EXP RV (p)`          = pm(exp_RV, exp_RV_p),
    `SDG Mantel r (p)`    = pm(sdg_mantel_r, sdg_mantel_p),
    `SDG RV (p)`          = pm(sdg_RV, sdg_RV_p)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Truncation degree

For SP-SPHARM, the truncation range was defined a priori using two
criteria: cumulative power ≥ 99% and the onset of noise indicated by the
cross-specimen CV (approximately $l$ = 9). At the fifth degree,
cumulative power had already exceeded 99% (EXP:
`r tv("SP","EXP",5,"SP_cumpower_at_lmax",2)`%; SDG:
`r tv("SP","SDG",5,"SP_cumpower_at_lmax",2)`%), and reached
`r tv("SP","EXP",6,"SP_cumpower_at_lmax",2)`% and
`r tv("SP","SDG",6,"SP_cumpower_at_lmax",2)`% at the sixth degree. The
two criteria therefore defined an acceptable truncation range of $l$ =
5–8 rather than a single optimal value. The subsequent sensitivity sweep
was used to test whether the conclusions were stable within this
pre-defined range, not to select the truncation degree.

The sensitivity analysis showed that truncation at $l$ = 1–5 reduced
discrimination. In the experimental assemblage, the number of resolved
typology pairs decreased from `r tint("SP","EXP",6,"SP_nsig05")` of
`r tint("SP","EXP",6,"n_pairs")` to `r tint("SP","EXP",5,"SP_nsig05")`
of `r tint("SP","EXP",5,"n_pairs")`, with `r tlost("SP","EXP",5)` being
the pair lost; in SDG, the core-type PERMANOVA $p$ value increased from
`r tv("SP","SDG",6,"SP_p")` to `r tv("SP","SDG",5,"SP_p")`, crossing the
conventional significance threshold. By contrast, $l$ = 1–6 and $l$ =
1–7 produced consistent conclusions. At $l$ = 1–8, the morphology–scar
association in the experimental assemblage became significant (RV =
`r tv("SP","EXP",8,"RV")`, $p$ = `r tv("SP","EXP",8,"RV_p")`), with the
result already marginal at $l$ = 1–7 ($p$ =
`r tv("SP","EXP",7,"RV_p")`). This reversal was not observed in SDG.
Because the CV diagnostic indicates increasing noise in the
higher-degree coefficients, and the reversal occurs only in one of the
two assemblages, we interpret this isolated high-degree association as
more consistent with computational noise than with a technological
signal. This also illustrates why the upper-bound CV criterion is needed
alongside the cumulative-power criterion.

Overall, the main conclusions remain stable across $l$ =
`r tstab("SP","EXP")` in the experimental assemblage and $l$ =
`r tstab("SP","SDG")` in SDG. We therefore use $l$ = 1–`r tanchor("SP")`
for SP-SPHARM: it lies within the pre-defined stable range and, among
the settings that preserve the conclusions, uses the lowest
dimensionality. This choice follows the same principle applied to
M-SPHARM, namely to retain sufficient information for discrimination
while avoiding additional higher-degree dimensions that do not provide a
consistent improvement in the results.

```{r}
#| label: tbl-si-truncation
#| tbl-cap: "Per truncation sensitivity metrics for both sweeps. The asterisk marks the main analysis setting of each sweep ($l$ = 1–6 for SP-SPHARM, $l$ = 1–8 for M-SPHARM). Because the ILR transform maps the retained degrees onto one fewer coordinate, the number of retained coordinates (n coords) is given: models are not nested across settings and $R^2$ is therefore not directly comparable between rows (see text)"
trn$summary |>
  transmute(
    scan, assemblage, varied_lmax, is_anchor, n_coords, n_pairs,
    R2     = ifelse(scan == "SP", SP_R2, M_R2),
    p      = ifelse(scan == "SP", SP_p, M_p),
    nsig   = ifelse(scan == "SP", SP_nsig05, M_nsig05),
    med    = ifelse(scan == "SP", SP_med_neglog10_p, M_med_neglog10_p),
    mantel_r, mantel_p, RV, RV_p
  ) |>
  (\(d) left_join(filter(d, assemblage == "EXP"),
                  filter(d, assemblage == "SDG"),
                  by = c("scan", "varied_lmax"), suffix = c("_E", "_S")))() |>
  arrange(desc(scan), varied_lmax) |>
  transmute(
    `sweep`            = ifelse(scan == "SP", "SP-SPHARM", "M-SPHARM"),
    `truncation`       = ifelse(is_anchor_E, paste0("1-", varied_lmax, "*"),
                                             paste0("1-", varied_lmax)),
    `n coords`         = n_coords_E,
    `EXP R2`           = sprintf("%.3f", R2_E),
    `EXP p`            = sprintf("%.3f", p_E),
    `EXP sig pairs`    = paste0(nsig_E, "/", n_pairs_E),
    `EXP med -log10 p` = sprintf("%.2f", med_E),
    `SDG type R2`      = sprintf("%.3f", R2_S),
    `SDG p`            = sprintf("%.3f", p_S),
    `EXP Mantel r (p)` = pm(mantel_r_E, mantel_p_E),
    `EXP RV (p)`       = pm(RV_E, RV_p_E),
    `SDG Mantel r (p)` = pm(mantel_r_S, mantel_p_S),
    `SDG RV (p)`       = pm(RV_S, RV_p_S)
  ) |>
  knitr::kable()
```

{{< pagebreak >}}

## Summary

Across both descriptors, the main conclusions remain stable across the principal free parameters. These include the early truncation ranges ($l$ = 1–6 for
the scar descriptor, $l$ = 1–8 for morphology), the idealized-core discrimination, the stronger contribution of core type than raw material or layer in the archaeological morphology and scar-patterning analyses, and the decoupling between morphology and scar patterning. The latter remains unchanged when either descriptor is perturbed across the bandwidth, scar-threshold, and mesh sweeps, as well as across the stable truncation intervals of both descriptors. The truncation sweep provides the only qualification. At the extreme settings
($l$ = 1–8 for the scar descriptor, $l$ = 1–6 for morphology) the
experimental RV coefficient becomes nominally significant
(@tbl-si-truncation), although the association remains negligible in
magnitude and is not corroborated by the Mantel test in either case. The
main analysis settings sit inside the stable interval of both sweeps.

The perturbation tests also show that the results are not uniformly sensitive to input-data quality. Perturbing the scar direction
landmarks themselves leaves the decoupling result unchanged at every
error type and level, and leaves type discrimination intact under
angular jitter to $\sigma$ = 20°. Polarity annotation error is the one
factor tested anywhere in this section that has a material effect on a
reported result, taking SP-SPHARM's pairwise resolution from
`r aint("baseline", 0, "n_sig_05_med")` of 10 typology pairs to
`r aint("polarity", 0.1, "n_sig_05_med")` at a 10% flip rate
(@tbl-si-perturbation). Thus, the results are insensitive to the analytical parameter choices and to angular imprecision, but they depend on the proximal–distal judgements being largely correct.

{{< pagebreak >}}

# Statistical power of the morphology and scar patterning covariation test

The tests of covariation between core morphology and scar patterning
returned null results in both assemblages (main text; @tbl-si-bandwidth,
@tbl-si-scar, @tbl-si-mesh). Because a non-significant result does not
by itself distinguish a genuine absence of association from insufficient
power to detect one, we assessed the informativeness of these nulls
directly. Rather than computing observed power from the realized effect,
which is uninformative [@hoenigAbusePowerPervasive2001], we followed the
recommendation to characterize the range of effect sizes a fixed design
can resolve [@lakensSampleSizeJustification2022]. Three quantities were
computed for each assemblage, reusing the same ILR transformed power
spectra as the main analysis. At the main analysis setting the
recomputed coefficients reproduced the reported raw RV and Mantel
statistics, confirming that the analysis was anchored to the values in
the main text.

We evaluated the null results in three complementary ways. First,
because raw RV coefficients can be upwardly biased when the number of
variables is large relative to sample size
[@adamsEvaluatingModularityMorphometric2016;
@smildeMatrixCorrelationsHighdimensional2009], we recalculated the
association using the bias-corrected modified RV
[@smildeMatrixCorrelationsHighdimensional2009]. Second, we used a
percentile bootstrap over specimens (2,000 resamples) to estimate how
large the association might plausibly be, while keeping each specimen's
morphology and scar pattern coordinates paired. Third, we used Monte
Carlo simulation to estimate how sensitive the Mantel and RV permutation
tests were at the realized sample sizes [@rohlfUseTwoBlockPartial2000;
@manlyRandomizationBootstrapMonte2007]. In these simulations, data were
generated with a known degree of association between the two blocks,
following established practice in morphometric studies of modularity and
integration [@adamsEvaluatingModularityMorphometric2016]. Variation
within each block was drawn from the empirical mean and covariance of
that block, and a shared component was added to create a controlled
level of between-block association. The strength of this component was
calibrated to a target population RV, by which we mean the underlying
between-block association imposed in the simulated population, rather
than the RV estimated from any single finite sample. Because the
simulated association was aligned with the leading axis of each block,
the resulting minimum detectable effects should be understood as
best-case values: a less favorably structured association would require
a larger effect to be detected [@lakensSampleSizeJustification2022].
Simulation settings were kept modest, and all procedures used fixed
seeds.

These checks suggest that the observed decoupling between morphology and
scar patterning is unlikely to be explained by sample size alone. After
bias correction, the modified RV fell from `r pv("EXP","raw_RV")` to
`r pv("EXP","RV2")` in the EXP assemblage and from
`r pv("SDG","raw_RV")` to `r pv("SDG","RV2")` in the SDG assemblage.
This indicates that much of the already small raw association was
attributable to small-sample inflation rather than to a strong
underlying signal. The bootstrap results also placed only modest upper
bounds on the possible association: approximately RV =
`r pv("EXP","RV2_CI_high",2)` for EXP and `r pv("SDG","RV2_CI_high",2)`
for SDG, and Mantel $r$ = `r pv("EXP","mantel_CI_high",2)` for EXP and
`r pv("SDG","mantel_CI_high",2)` for SDG. The simulation results showed
that the RV permutation test had sufficient power to detect a moderate
association at the realized sample sizes. It reached at least 80% power
at a population RV of `r pv("EXP","MDES_RV")` in EXP and
`r pv("SDG","MDES_RV")` in SDG, and exceeded 96% power by a population
RV of about 0.13. The Mantel test was less sensitive, reaching 80% power
only at a population RV of `r pv("EXP","MDES_mantel")` in EXP and
`r pv("SDG","MDES_mantel")` in SDG, consistent with the lower power of
Mantel tests relative to covariance-based statistics. Power curves are
shown in @fig-si-power, and all values are summarized in @tbl-si-power.

Taken together, these results support the interpretation that core
morphology and scar patterning are largely decoupled at the assemblage
level. If a moderate or stronger association had been present, it would
probably have been detected by the RV test. Instead, the bias-corrected
estimates were close to zero in both assemblages. A very weak residual
association, below a population RV of roughly
`r pv("EXP","MDES_RV",2)`–`r pv("SDG","MDES_RV",2)`, cannot be excluded.
The small per-type subsamples also remain underpowered, which is why the
within-type analyses are treated as exploratory in the main text.

```{r}
#| label: tbl-si-power
#| tbl-cap: "Power and sensitivity metrics for the morphology and scar patterning covariation test, by assemblage. Raw RV and Mantel $r$ are the values reported in the main text (anchor). RV2 is the bias-corrected modified RV coefficient. The 95% confidence intervals are percentile bootstrap intervals (2,000 resamples) and are interpreted as upper bounds on the association (see text). MDES is the minimum population effect detectable at 80% power, estimated by Monte Carlo simulation, and is a lower bound. Values in parentheses are permutation p values"
pwr$metrics |>
  transmute(
    Assemblage        = assemblage,
    n                 = n,
    `Raw RV (p)`      = pm(raw_RV, raw_RV_p),
    RV2               = sprintf("%.3f", RV2),
    `RV2 95% CI`      = sprintf("[%.3f, %.3f]", RV2_CI_low, RV2_CI_high),
    `Mantel r (p)`    = pm(mantel_r, mantel_p),
    `Mantel r 95% CI` = sprintf("[%.3f, %.3f]", mantel_CI_low, mantel_CI_high),
    `MDES RV`         = sprintf("%.3f", MDES_RV),
    `MDES Mantel`     = sprintf("%.3f", MDES_mantel)
  ) |>
  knitr::kable()
```

```{r}
#| label: fig-si-power
#| fig-cap: "Monte Carlo power of the RV permutation test and the Mantel test as a function of the population RV coefficient, at the realized sample sizes of the experimental and Sandinggai assemblages. The dashed horizontal line marks 80% power and the dotted vertical line marks the observed raw RV"
#| fig-width: 9
#| fig-height: 4.5
pwr$fig_power
```

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::
